### 2019-09-04 paper/paper.tex
Initial commit

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan]{acmart}\settopmatter{}

%% Conference information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
\acmYear{2018}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%\citestyle{acmauthoryear}  %% For author/year citations
%\citestyle{acmnumeric}     %% For numeric citations
%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
                            %% sorting of references within a single citation;
                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
                            %% rendered as [14,5,2] rather than [2,5,14].
%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
                            %% compression of sequential references within a
                            %% single citation;
                            %% e.g., \cite{Baker12,Baker14,Baker16}
                            %% rendered as [2,3,4] rather than [2-4].

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from traditional SIGPLAN
%% proceedings format to PACMPL format must update the
%% '\documentclass' and topmatter commands above; see
%% 'acmart-pacmpl-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\begin{document}

%% Title information
\title[]{Quantum Hoare Types}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{Kartik Singhal}
%\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{0000-0003-1132-269X}             %% \orcid is optional
\affiliation{
%  \position{Position1}
  \department{Department of Computer Science}              %% \department is recommended
  \institution{University of Chicago}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
  \country{United States}                    %% \country is recommended
}
\email{ks@cs.uchicago.edu}          %% \email is recommended

%% Author with two affiliations and emails.
\author{John Reppy}
%\authornote{with author2 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position2a}
  \department{Department of Computer Science}             %% \department is recommended
  \institution{University of Chicago}           %% \institution is required
%  \streetaddress{Street2a Address2a}
%  \city{City2a}
%  \state{State2a}
%  \postcode{Post-Code2a}
  \country{United States}                   %% \country is recommended
}
\email{jhr@cs.uchicago.edu}         %% \email is recommended
%\affiliation{
%  \position{Position2b}
%  \department{Department2b}             %% \department is recommended
%  \institution{Institution2b}           %% \institution is required
%  \streetaddress{Street3b Address2b}
%  \city{City2b}
%  \state{State2b}
%  \postcode{Post-Code2b}
%  \country{Country2b}                   %% \country is recommended
%}
%\email{first2.last2@inst2b.org}         %% \email is recommended

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Text of abstract \ldots.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\section{Introduction}

Text of paper \ldots

\cite{DBLP:conf/oopsla/HuangM18}

\cite{Huang:2019:SAV:3307650.3322213}

\cite{nanevski_morrisett_birkedal_2008}

\cite{paykin18}

\cite{rand18}

\cite{2015arXiv151002198R}

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by
  EPiQC, an \grantsponsor{GS100000001}{NSF}{http://dx.doi.org/10.13039/100000001}
  Expedition in Computing, under Grant
  No.~\grantnum{GS100000001}{1730449}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the authors and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

%% Bibliography
\bibliography{bibfile}

%% Appendix
\appendix
\section{Appendix}

Text of appendix \ldots

\end{document}

### 2019-09-10 paper/paper.tex
some test lang
Migrate to PACMPL format

@@
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 %% For double-blind review submission, w/ CCS and ACM Reference
-%\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true}
+%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
 %% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
-%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 %% For single-blind review submission, w/ CCS and ACM Reference
-%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true}
+%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
 %% For final camera-ready submission, w/ required CCS and ACM Reference
-%\documentclass[sigplan]{acmart}\settopmatter{}
+%\documentclass[acmsmall]{acmart}\settopmatter{}
 
-%% Conference information
+%% Journal information
 %% Supplied to authors by publisher for camera-ready submission;
 %% use defaults for review submission.
-\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
+\acmJournal{PACMPL}
+\acmVolume{1}
+\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
+\acmArticle{1}
 \acmYear{2018}
-\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
+\acmMonth{1}
 \acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
@@
 %% Citation style
-%\citestyle{acmauthoryear}  %% For author/year citations
-%\citestyle{acmnumeric}     %% For numeric citations
-%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
-                            %% sorting of references within a single citation;
-                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
-                            %% rendered as [14,5,2] rather than [2,5,14].
-%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
-                            %% compression of sequential references within a
-                            %% single citation;
-                            %% e.g., \cite{Baker12,Baker14,Baker16}
-                            %% rendered as [2,3,4] rather than [2-4].
+%% Note: author/year citations are required for papers published as an
+%% issue of PACMPL.
+\citestyle{acmauthoryear}   %% For author/year citations
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Note: Authors migrating a paper from traditional SIGPLAN
-%% proceedings format to PACMPL format must update the
-%% '\documentclass' and topmatter commands above; see
-%% 'acmart-pacmpl-template.tex'.
+%% Note: Authors migrating a paper from PACMPL format to traditional
+%% SIGPLAN proceedings format must update the '\documentclass' and
+%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+From Xavier Leroy's OPLSS Slides:
+
+Arithmetic expressions:
+
+\[a ::= n \mid x \mid a1 + a2 \mid a1 - a2 \mid a1 \times a2\]
+
+Boolean expressions:
+
+\begin{align*}
+	b ::=&\ \textsf{true} \mid \textsf{false} \mid a1 = a2 \mid a1 \le a2\\
+	\mid&\ \textsf{not}\ b \mid b1\ \textsf{and}\ b2
+\end{align*}
+
+Commands (statements):
+
+\begin{align*}
+	c ::=&\ \textsf{SKIP} 														&\textsf{(do\ nothing)}\\
+	\mid&\ x ::= a 															&\textsf{(assignment)}\\
+	\mid&\ 1; ; c2 															&\textsf{(sequence)}\\
+	\mid&\ \textsf{IFB}\ b\ \textsf{THEN}\ c1\ \textsf{ELSE}\ c2\ \textsf{FI} 	&\textsf{(conditional)}\\
+	\mid&\ \textsf{WHILE}\ b\ \textsf{DO}\ c\ \textsf{END} 					&\textsf{(loop)}
+\end{align*}
+
 \cite{DBLP:conf/oopsla/HuangM18}

### 2019-09-10 paper/paper.tex
Switch to a format suitable for master's report
Add paper template from SPJ's talk

@@
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 %% For double-blind review submission, w/ CCS and ACM Reference
@@
 %% Author with two affiliations and emails.
-\author{John Reppy}
+%\author{John Reppy}
 %\authornote{with author2 note}          %% \authornote is optional;
@@
 %\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
-\affiliation{
-%  \position{Position2a}
-  \department{Department of Computer Science}             %% \department is recommended
-  \institution{University of Chicago}           %% \institution is required
-%  \streetaddress{Street2a Address2a}
-%  \city{City2a}
-%  \state{State2a}
-%  \postcode{Post-Code2a}
-  \country{United States}                   %% \country is recommended
-}
-\email{jhr@cs.uchicago.edu}         %% \email is recommended
+%\affiliation{
+%%  \position{Position2a}
+%  \department{Department of Computer Science}             %% \department is recommended
+%  \institution{University of Chicago}           %% \institution is required
+%%  \streetaddress{Street2a Address2a}
+%%  \city{City2a}
+%%  \state{State2a}
+%%  \postcode{Post-Code2a}
+%  \country{United States}                   %% \country is recommended
+%}
+%\email{jhr@cs.uchicago.edu}         %% \email is recommended
 %\affiliation{
@@
 Text of abstract \ldots.
+
+(4 sentences)
 \end{abstract}
@@
 
+\textit{- Describe the problem}
+
+\textit{- State your contributions}
+
+(1 page)
+
+\subsection{Contributions}
+In this section we present the main contributions of the paper
+
+\textit{bulleted list of contributions}
+
+\section{Problem}
+
+\textit{Use an example to introduce the problem}
+
+(1 page)
+
+\section{Idea}
+
+The main idea of this paper is \ldots
+
+(2 pages)
+
+\section{Details}
+(5 pages)
+
 From Xavier Leroy's OPLSS Slides:
@@
 
+\section{Related Work}
+(1-2 pages)
+
 \cite{DBLP:conf/oopsla/HuangM18}
@@
 \cite{2015arXiv151002198R}
+
+\section{Conclusions and Further Work}
+
+(0.5 page)
 

### 2019-09-10 paper/paper.tex
Add intro and figure from poster

@@
                         %% http://ctan.org/pkg/subcaption
+
+%% Added by kartik
+\usepackage[braket]{qcircuit}
 
@@
 \begin{abstract}
-Text of abstract \ldots.
+(4 sentences)
 
-(4 sentences)
+\textit{Text of abstract \ldots.}
+
 \end{abstract}
@@
 \section{Introduction}
-
-Text of paper \ldots
+(1 page)
 
@@
 
-(1 page)
+It is difficult to reason about the correctness of quantum programs; both while running them and while writing them.
+Sound static type systems help prevent a huge class of bugs from occurring, but since the realm of quantum programming is still new, there is not a lot of consensus on what kind of types make the most sense and whether they help programmers reason about the semantic properties associated with the quantum algorithms that they are implementing.
+
+Previous work such as Proto-Quipper\cite{2015arXiv151002198R} and QWire\cite{rand18,paykin18} utilize a linear type system and dependent types to enforce a small subset of semantic properties in the quantum setting such as the no-cloning theorem and whether a unitary operator is of the right dimension.
+These advances in quantum type systems although helpful, still fall short in encoding and enforcing even more useful properties that one would like to be able to express for the purpose of verification.
+Inspired by the use of Hoare triples in the verification of imperative programs and building on the idea of Hoare Types in classical programming languages, we propose Quantum Hoare Types aimed at enabling both sound static type checking and formal verification of quantum programs.
 
 \subsection{Contributions}
-In this section we present the main contributions of the paper
+(bulleted list of contributions)
 
-\textit{bulleted list of contributions}
+\textit{In this section we present the main contributions of the paper\ldots}
 
 \section{Problem}
+(1 page)
 
@@
 
-(1 page)
+\subsection{Motivating Example}
 
-\section{Idea}
+\begin{figure}
+	\centerline{
+		\Qcircuit @C=1em @R=.7em {
+		\lstick{\ket{q_0}} & \gate{H} & \ctrl{1} & \qw \\
+		\lstick{\ket{q_1}} & \qw         & \targ    & \qw
+	}}
+	\caption{Bell Pair Creation}
+	\label{fig:bellpair}
+\end{figure}
 
-The main idea of this paper is \ldots
+In Figure \ref{fig:bellpair}, we see a standard Bell Pair generation circuit.
+
+Assume the input is \ket{00}. A Hadamard gate is applied to qubit $q_0$ and then a CNOT gate is applied to its outcome as the control and the qubit $q_1$ as the target.
 

### 2019-09-10 paper/paper.tex
Add rest of the poster

@@
 
+The result will be a 2-qubit entangled state aka the first Bell state that will, on measurement, collapse into either \ket{00} or \ket{11} with equal probability.
+
+\subsection{Desired Properties}
+For our example, we would like to guarantee correctness of properties such as the following:
+\begin{itemize}
+	\item Before application of H gate, the qubit $q_0$ should have a deterministic integer value (on measurement)
+	\item After application of H, $q_0$ should be in 50-50 probabilistic distribution of being in the two computational states
+	\item After application of CNOT, the two qubits should be in a maximally entangled state
+	\item Given a certain input to the whole circuit, we should get the corresponding Bell state
+\end{itemize}
+
+\section{Idea}
 (2 pages)
+
+\textit{The main idea of this paper is \ldots}
+
+Recent work at Princeton as part of EPiQC\cite{DBLP:conf/oopsla/HuangM18,Huang:2019:SAV:3307650.3322213} has identified several classes of bugs in quantum programs and proposed approaches to tackle them.
+They proposed assertion checking using preconditions and post-conditions as one of the most effective techniques for most such classes of bugs.
+
+As programming languages researchers, we think it will be more useful to encode such assertions into a static type system for both formal verification and aiding the programmers in writing correct programs from the start.
 
@@
 (5 pages)
+
+\subsection{Notation and Background}
+\begin{itemize}
+	\item \textbf{Dependent Types}
+	\[ \mathrm{inner\_product}: \Pi m: \mathrm{nat}.\mathrm{vector}(m) \times \mathrm{vector}(m) \rightarrow \mathrm{nat} \]
+
+	\item \textbf{Hoare Triples} from \textbf{Floyd-Hoare Logic}
+	\[ \{P\}\ C\ \{Q\} \]
+	$P$ and $Q$ are pre and postcondition respectively on the program state, while $C$ is the command. For example:
+	\[ \{\mathrm{True}\}\ x := 3\ \{x=3\} \]
+	\item \textbf{Hoare Types} from \textbf{Hoare Type Theory} (\textbf{HTT}) that combine the strengths of dependent types and Hoare logic.
+	\[ \{P\}\ \mathrm{res}:A\ \{Q\}\]
+	for a stateful computation executed in a heap that satisfies precondition $P$ and returns a value of type $A$ in a heap that satisfies postcondition $Q$. For example:
+	\[ \mathrm{alloc} : \forall \alpha . \Pi x : \alpha. \{\mathrm{emp}\}\ y : \mathrm{nat}\ \{ y \mapsto_{\alpha} x\}\]
+\end{itemize}
+
+\subsection{Quantum Hoare Types}
+Quantum Hoare typing for the Hadamard and the CNOT operations will look something like:
+
+\abovedisplayskip=0pt
+\belowdisplayskip=12pt
+
+\[ \mathbf{H}(q_0) : \{ \mathrm{q\_state}(q_0) \in \{\ket{0}, \ket{1} \}\}\ r : \mathrm{unit}\ \{ \mathrm{q\_state}(q_0) \in \{\ket{+}, \ket{-}\} \} \]
+
+\[ \mathbf{CNOT}(q_0, q_1) : \{ \mathrm{q\_state}(q_0) \in \{ \ket{+}, \ket{-} \} \wedge \mathrm{q\_state}(q_0) \in \{\ket{0}, \ket{1} \}\ r : \mathrm{unit}\ \{ \mathrm{entangled}(q_0, q_1) \}\]
+
+Using sequential composition, type of the complete program will be:
+
+\[ \mathbf{BellPair}(q_0, q_1) : \{ \mathrm{q\_state}(q_i) \in \{\ket{0}, \ket{1} \}\}\ r : \mathrm{unit}\ \{ \mathrm{entangled}(q_0, q_1) \} \]
+
+where the predicates need to be encoded in terms of suitable dependent types. This is ongoing work.
+
+\subsubsection{Challenges}
+\begin{itemize}
+	\item The type system for Hoare Type Theory is fairly complex. We are mechanizing the type system to handle design changes easily.
+	\item We need to be able to reason about both decomposable quantum state and entangled state. This aspect may require incorporating ideas from Separation Logic.
+\end{itemize}
+
+\subsubsection{Benefits}
+\begin{itemize}
+	\item Potential to be a unified system for programming, specification and reasoning about quantum programs similar to HTT.
+	\item Allows composition of verification in the same style as Hoare logic.
+\end{itemize}
+
+\hrulefill
 

### 2019-09-12 paper/paper.tex
Use mathrm instead of textsf
Use mathpartir for multiple equations

@@
 \usepackage[braket]{qcircuit}
+\usepackage{mathpartir}
 
@@
 
-Recent work at Princeton as part of EPiQC\cite{DBLP:conf/oopsla/HuangM18,Huang:2019:SAV:3307650.3322213} has identified several classes of bugs in quantum programs and proposed approaches to tackle them.
+Recent work at Princeton as part of EPiQC\footnote{EPiQC: Enabling Practical-Scale Quantum Computation:  \url{https://epiqc.cs.uchicago.edu}}\cite{DBLP:conf/oopsla/HuangM18,Huang:2019:SAV:3307650.3322213} has identified several classes of bugs in quantum programs and proposed approaches to tackle them.
 They proposed assertion checking using preconditions and post-conditions as one of the most effective techniques for most such classes of bugs.
@@
 
-\abovedisplayskip=0pt
-\belowdisplayskip=12pt
+\begin{mathpar}
+	\mathbf{H}(q_0) : \{ \mathrm{q\_state}(q_0) \in \{\ket{0}, \ket{1} \}\}\ r : \mathrm{unit}\ \{ \mathrm{q\_state}(q_0) \in \{\ket{+}, \ket{-}\} \}
 
-\[ \mathbf{H}(q_0) : \{ \mathrm{q\_state}(q_0) \in \{\ket{0}, \ket{1} \}\}\ r : \mathrm{unit}\ \{ \mathrm{q\_state}(q_0) \in \{\ket{+}, \ket{-}\} \} \]
+	\\
 
-\[ \mathbf{CNOT}(q_0, q_1) : \{ \mathrm{q\_state}(q_0) \in \{ \ket{+}, \ket{-} \} \wedge \mathrm{q\_state}(q_0) \in \{\ket{0}, \ket{1} \}\ r : \mathrm{unit}\ \{ \mathrm{entangled}(q_0, q_1) \}\]
+	\mathbf{CNOT}(q_0, q_1) : \{ \mathrm{q\_state}(q_0) \in \{ \ket{+}, \ket{-} \} \wedge \mathrm{q\_state}(q_0) \in \{\ket{0}, \ket{1} \}\ r : \mathrm{unit}\ \{ \mathrm{entangled}(q_0, q_1) \}
+\end{mathpar}
 
@@
 \begin{align*}
-	b ::=&\ \textsf{true} \mid \textsf{false} \mid a1 = a2 \mid a1 \le a2\\
-	\mid&\ \textsf{not}\ b \mid b1\ \textsf{and}\ b2
+	b ::=&\ \mathrm{true} \mid \mathrm{false} \mid a1 = a2 \mid a1 \le a2\\
+	\mid&\ \mathrm{not}\ b \mid b1\ \mathrm{and}\ b2
 \end{align*}
@@
 \begin{align*}
-	c ::=&\ \textsf{SKIP} 														&\textsf{(do\ nothing)}\\
-	\mid&\ x ::= a 															&\textsf{(assignment)}\\
-	\mid&\ 1; ; c2 															&\textsf{(sequence)}\\
-	\mid&\ \textsf{IFB}\ b\ \textsf{THEN}\ c1\ \textsf{ELSE}\ c2\ \textsf{FI} 	&\textsf{(conditional)}\\
-	\mid&\ \textsf{WHILE}\ b\ \textsf{DO}\ c\ \textsf{END} 					&\textsf{(loop)}
+	c ::=&\ \mathrm{SKIP}	&\mathrm{(do\ nothing)}\\
+	\mid&\ x ::= a	&\mathrm{(assignment)}\\
+	\mid&\ 1; ; c2	&\mathrm{(sequence)}\\
+	\mid&\ \mathrm{IFB}\ b\ \mathrm{THEN}\ c1\ \mathrm{ELSE}\ c2\ \mathrm{FI}	&\mathrm{(conditional)}\\
+	\mid&\ \mathrm{WHILE}\ b\ \mathrm{DO}\ c\ \mathrm{END}	&\mathrm{(loop)}
 \end{align*}

### 2019-09-20 paper/paper.tex
Add exisiting bibliography along with some books
WIP
Add syntax template
Switch to quantum types and add bell pair example

@@
 
+\subsection{Syntax}
+
+See Fig. \ref{fig:syntax}. Let's assume $\alpha$ represents type of a non-specific quantum state vector (that includes entangled qubits).
+
+\begin{figure}
+\begin{tabular}{lrcl}
+	\textit{Types} & $A, B, C$ & ::= & $ \alpha \mid \mathrm{qbit} \mid A \otimes B \mid 1 \mid \Pi x{:}A.B \mid \Psi. X. \{P\} x{:}A \{Q\} \mid \forall \alpha.A$ \\
+	\textit{Monotypes} & $\tau, \sigma$ & ::= & $ \alpha \mid \mathrm{qbit} \mid \tau \otimes \sigma \mid 1 \mid \Pi x{:}\tau.\sigma \mid \Psi. X. \{P\} x{:}\tau \{Q\}$ \\
+	\textit{Assertions} & $P, Q, R$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
+		$ \mathrm{id}_A(M,N) \mid \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid P \supset Q \mid \neg P \mid \\
+		\forall x{:}A.P \mid \forall \alpha.P \mid \forall h{:}\mathrm{heap}.P \mid \mathrm{is\_entangled}\ M $
+	\end{minipage}\\
+	\textit{(Quantum) Heaps} & $H, G$ & ::= & $ h \mid \mathrm{empty} \mid \mathrm{upd}_{\tau}(H, M, N) $\\
+	\textit{Elim terms} & $K, L$ & ::= & $ x \mid K\ M \mid K\ \tau \mid M : A $\\
+	\textit{Intro terms} & $M, N, O$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
+		$ K \mid (\,) \mid \lambda x.\ M \mid \Lambda \alpha.\ M \mid \mathrm{do}\ E \mid \ket{0} \mid \ket{1} \mid (M, N) \mid M==N $
+	\end{minipage}\\
+	\textit{Commands} & $c$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
+		$ \mathrm{alloc}_\tau\ M \mid \mathrm{dealloc}\ M \mid \mathrm{if}_A\ M\ \mathrm{then}\ E_1\ \mathrm{else}\ E_2 \mid \\
+		\mathrm{fix}\ f(y{:}A){:}B = \mathrm{do}\ E\ \mathrm{in\ eval}\ f M \mid \mathrm{apply}(M, N) \mid \mathrm{capply}(M, N, O) $
+	\end{minipage}\\
+	\textit{Computations} & $E, F$ & ::= & $ \mathrm{return}\ M \mid x \gets K; E \mid x \Leftarrow c; E \mid x =_A M; E $\\
+	\textit{Variable context} & $\Delta, \Psi$ & ::= & $ \cdot \mid \Delta, x{:}A \mid \Delta, \alpha $\\
+	\textit{Heap context} & $X$ & ::= & $ \cdot \mid X, h $\\
+	\textit{Assertion context} & $\Gamma$& ::= & $\cdot \mid \Gamma, P $
+\end{tabular}
+	\caption{Syntax}
+\label{fig:syntax}
+\end{figure}
+
+\textit{Terms.} Split into introduction and elimination terms.
+Intro term for Hoare type is do $E$ that represents effectful computation and will suspend the evaluation; corresponding elimination form (which is a computation) is $x \gets K; E$.
+
+\textit{Computations.}  All effectful stuff happens here.
+Separated by semicolon.
+Should terminate with a return value.
+return $M$ and $x \gets K; E$ correspond to monadic \textit{unit} and \textit{bind}.
+
+Simple example to create a Bell Pair using QHTT commands:
+
+\begin{tabular}{ll}
+	$\mathrm{plus\_minus} =_{\mathrm{qbit} \to \mathrm{qbit}} \lambda q.\ \mathrm{do}\ ($&$a \Leftarrow \mathrm{apply}(\mathrm{Hadamard}, q);$\\
+																																	&$\mathrm{return}\ a);$\\
+	$\mathrm{share} =_{\mathrm{qbit} \to \mathrm{qbit} \otimes \mathrm{qbit}} \lambda a.\ \mathrm{do}\ ($&$q \Leftarrow \mathrm{alloc_{qbit}}\ \ket{0};$\\
+																																							   &$b \Leftarrow \mathrm{capply}(\mathrm{a, QNot}, q);$\\
+																																							   &$\mathrm{return}\ (a, b));$\\
+	$a \gets \mathrm{plus\_minus} \ket{0};$ & \\
+	$\mathrm{bell_{00}} \gets \mathrm{share}\ a;$ & \\
+	$\mathrm{return}\ \mathrm{bell_{00}}; $&
+\end{tabular}
+
+\subsubsection{Assertions}
+
+\subsubsection{Example}
+
+\subsection{Typing Rules}
+
+\subsection{Annotated Example}
+
+\subsection{Evaluation Rules}
+
 \subsection{Notation and Background}
@@
 
+\cite{PGL-045}
+
 \section{Conclusions and Further Work}

### 2019-09-27 paper/paper.tex
Add Silq (pretty close lang design to what I was thinking)
Normalize used citations

@@
 
-Previous work such as Proto-Quipper\cite{2015arXiv151002198R} and QWire\cite{rand18,paykin18} utilize a linear type system and dependent types to enforce a small subset of semantic properties in the quantum setting such as the no-cloning theorem and whether a unitary operator is of the right dimension.
+Previous work such as Proto-Quipper\cite{ross2015algebraic} and QWire\cite{paykin2018,rand2018a} utilize a linear type system and dependent types to enforce a small subset of semantic properties in the quantum setting such as the no-cloning theorem and whether a unitary operator is of the right dimension.
 These advances in quantum type systems although helpful, still fall short in encoding and enforcing even more useful properties that one would like to be able to express for the purpose of verification.
@@
 
-Recent work at Princeton as part of EPiQC\footnote{EPiQC: Enabling Practical-Scale Quantum Computation:  \url{https://epiqc.cs.uchicago.edu}}\cite{DBLP:conf/oopsla/HuangM18,Huang:2019:SAV:3307650.3322213} has identified several classes of bugs in quantum programs and proposed approaches to tackle them.
+Recent work at Princeton as part of EPiQC\footnote{EPiQC: Enabling Practical-Scale Quantum Computation:  \url{https://epiqc.cs.uchicago.edu}}\cite{huang2018,huang2019} has identified several classes of bugs in quantum programs and proposed approaches to tackle them.
 They proposed assertion checking using preconditions and post-conditions as one of the most effective techniques for most such classes of bugs.
@@
 
-\cite{DBLP:conf/oopsla/HuangM18}
+\cite{huang2018}
 
-\cite{Huang:2019:SAV:3307650.3322213}
+\cite{huang2019}
 
-\cite{nanevski_morrisett_birkedal_2008}
+\cite{nanevski2008}
 
-\cite{paykin18}
+\cite{paykin2018}
 
-\cite{rand18}
+\cite{rand2018a}
 
-\cite{2015arXiv151002198R}
+\cite{ross2015algebraic}
 
-\cite{PGL-045}
+\cite{ringer2019}
+
+\cite{silq}
 

### 2019-10-11 paper/paper.tex
Get spacing right with code example
Hacky centering of code
Remove extra semicolon, move on to types
Add one note
Remove polymorphism to reduce complexity for first attempt at this system

@@
 
-\subsection{Syntax}
+\subsection{Abstract Syntax}
 
-See Fig. \ref{fig:syntax}. Let's assume $\alpha$ represents type of a non-specific quantum state vector (that includes entangled qubits).
+See Fig. \ref{fig:syntax}.
 
@@
 \begin{tabular}{lrcl}
-	\textit{Types} & $A, B, C$ & ::= & $ \alpha \mid \mathrm{qbit} \mid A \otimes B \mid 1 \mid \Pi x{:}A.B \mid \Psi. X. \{P\} x{:}A \{Q\} \mid \forall \alpha.A$ \\
-	\textit{Monotypes} & $\tau, \sigma$ & ::= & $ \alpha \mid \mathrm{qbit} \mid \tau \otimes \sigma \mid 1 \mid \Pi x{:}\tau.\sigma \mid \Psi. X. \{P\} x{:}\tau \{Q\}$ \\
+	\textit{Types} & $\tau, \sigma$ & ::= & $ \mathrm{qbit} \mid \tau \otimes \sigma \mid 1 \mid \Pi x{:}\tau.\sigma \mid \Psi. X. \{P\} x{:}\tau \{Q\}$ \\
 	\textit{Assertions} & $P, Q, R$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
-		$ \mathrm{id}_A(M,N) \mid \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid P \supset Q \mid \neg P \mid \\
-		\forall x{:}A.P \mid \forall \alpha.P \mid \forall h{:}\mathrm{heap}.P \mid \mathrm{is\_entangled}\ M $
+		$ \mathrm{id}_\tau(M,N) \mid \mathrm{seleq}_\tau(H,M,N) \mid \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid \\
+		P \supset Q \mid \neg P \mid \forall h{:}\mathrm{heap}.P $
 	\end{minipage}\\
 	\textit{(Quantum) Heaps} & $H, G$ & ::= & $ h \mid \mathrm{empty} \mid \mathrm{upd}_{\tau}(H, M, N) $\\
-	\textit{Elim terms} & $K, L$ & ::= & $ x \mid K\ M \mid K\ \tau \mid M : A $\\
+	\textit{Elim terms} & $K, L$ & ::= & $ x \mid K\ M \mid M : \tau $\\
 	\textit{Intro terms} & $M, N, O$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
-		$ K \mid (\,) \mid \lambda x.\ M \mid \Lambda \alpha.\ M \mid \mathrm{do}\ E \mid \ket{0} \mid \ket{1} \mid (M, N) \mid M==N $
+		$ K \mid (\,) \mid \lambda x.\ M \mid \mathrm{do}\ E \mid \ket{0} \mid \ket{1} \mid (M, N) \mid M==N $
 	\end{minipage}\\
 	\textit{Commands} & $c$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
-		$ \mathrm{alloc}_\tau\ M \mid \mathrm{dealloc}\ M \mid \mathrm{if}_A\ M\ \mathrm{then}\ E_1\ \mathrm{else}\ E_2 \mid \\
-		\mathrm{fix}\ f(y{:}A){:}B = \mathrm{do}\ E\ \mathrm{in\ eval}\ f M \mid \mathrm{apply}(M, N) \mid \mathrm{capply}(M, N, O) $
+		$ \mathrm{alloc}_\tau\ M \mid \mathrm{dealloc}\ M \mid \mathrm{if}_\tau\ M\ \mathrm{then}\ E_1\ \mathrm{else}\ E_2 \mid
+		\mathrm{fix}\ f(y{:}\tau){:}\sigma = \mathrm{do}\ E\ \mathrm{in\ eval}\ f M \mid \mathrm{apply}(M, N) \mid \mathrm{capply}(M, N, O) $
 	\end{minipage}\\
-	\textit{Computations} & $E, F$ & ::= & $ \mathrm{return}\ M \mid x \gets K; E \mid x \Leftarrow c; E \mid x =_A M; E $\\
-	\textit{Variable context} & $\Delta, \Psi$ & ::= & $ \cdot \mid \Delta, x{:}A \mid \Delta, \alpha $\\
+	\textit{Computations} & $E, F$ & ::= & $ \mathrm{return}\ M \mid x \gets K; E \mid x \Leftarrow c; E \mid x =_\tau M; E $\\
+	\textit{Variable context} & $\Delta, \Psi$ & ::= & $ \cdot \mid \Delta, x{:}\tau $\\
 	\textit{Heap context} & $X$ & ::= & $ \cdot \mid X, h $\\
@@
 
+\textit{Commands.} \textit{apply} is for unitary circuit application (apply circuit M to N), \textit{capply} that with controls M, circuit N and arguments O.
+(An alternative is to formulate gate/circuit application as assignment.)
+
 \textit{Computations.} All effectful stuff happens here.
@@
 return $M$ and $x \gets K; E$ correspond to monadic \textit{unit} and \textit{bind}.
+$x =_\tau M$ is syntactic sugar for the let-binding of $M:\tau$ to $x$.
 
-Simple example to create a Bell Pair using QHTT commands:
+Simple example to create a Bell Pair using QHTT commands:\\
+
+\begin{center}
+\begingroup
+\setlength{\tabcolsep}{0pt}
+\begin{tabular}{ll}
+
+	$\mathrm{plus\_minus} =_{\mathrm{qbit} \to \mathrm{qbit}} \lambda q.\ \mathrm{do}\ ($&$a \Leftarrow \mathrm{apply}(\mathrm{Hadamard}, q);$\\
+																																	&$\mathrm{return}\ a);$\\
+\end{tabular}
 
 \begin{tabular}{ll}
-	$\mathrm{plus\_minus} =_{\mathrm{qbit} \to \mathrm{qbit}} \lambda q.\ \mathrm{do}\ ($&$a \Leftarrow \mathrm{apply}(\mathrm{Hadamard}, q);$\\
-																																	&$\mathrm{return}\ a);$\\
 	$\mathrm{share} =_{\mathrm{qbit} \to \mathrm{qbit} \otimes \mathrm{qbit}} \lambda a.\ \mathrm{do}\ ($&$q \Leftarrow \mathrm{alloc_{qbit}}\ \ket{0};$\\
@@
 	$\mathrm{bell_{00}} \gets \mathrm{share}\ a;$ & \\
-	$\mathrm{return}\ \mathrm{bell_{00}}; $&
+	$\mathrm{return}\ \mathrm{bell_{00}}$&
 \end{tabular}
+\endgroup
+\end{center}
+
+In above, we can abbreviate the last two lines to: eval(share $a$).
+
+\textit{Types.}
 

### 2019-10-29 paper/paper.tex
General cleanup

@@
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 %% For double-blind review submission, w/ CCS and ACM Reference
@@
 %% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[acmsmall,nonacm]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 %% For single-blind review submission, w/ CCS and ACM Reference
@@
 \acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
-\acmArticle{1}
+\acmArticle{}
 \acmYear{2018}
@@
 \usepackage{mathpartir}
+\usepackage{xspace}
+
+\newcommand{\sqir}{\ensuremath{s\textsc{qir}}\xspace}
+\newcommand{\qwire}{\ensuremath{\mathcal{Q}\textsc{wire}}\xspace}
+\newcommand{\denote}[1]{\llbracket #1 \rrbracket\xspace}
 
@@
 %% Title information
-\title[]{Quantum Hoare Types}         %% [Short Title] is optional;
+\title[Quantum Hoare Types]{Quantum Hoare Types}         %% [Short Title] is optional;
                                         %% when present, will be used in
@@
 %  \postcode{Post-Code1}
-  \country{United States}                    %% \country is recommended
+  \country{USA}                    %% \country is recommended
 }
@@
 \ccsdesc[500]{Software and its engineering~General programming languages}
-\ccsdesc[300]{Social and professional topics~History of programming languages}
 %% End of generated code
@@
 
-Previous work such as Proto-Quipper\cite{ross2015algebraic} and QWire\cite{paykin2018,rand2018a} utilize a linear type system and dependent types to enforce a small subset of semantic properties in the quantum setting such as the no-cloning theorem and whether a unitary operator is of the right dimension.
+Previous work such as Proto-Quipper\cite{ross2015algebraic} and \qwire\cite{paykin2018,rand2018a} utilize a linear type system and dependent types to enforce a small subset of semantic properties in the quantum setting such as the no-cloning theorem and whether a unitary operator is of the right dimension.
 These advances in quantum type systems although helpful, still fall short in encoding and enforcing even more useful properties that one would like to be able to express for the purpose of verification.
@@
 
-\begin{figure}
+\begin{figure}[hb]
 	\centerline{
@@
 
-\begin{figure}
+\begin{figure}[hb]
 \begin{tabular}{lrcl}
@@
 
-Simple example to create a Bell Pair using QHTT commands:\\
+Simple example to create a Bell Pair using QHTT commands:
 
@@
 
-\hrulefill
-
-From Xavier Leroy's OPLSS Slides:
-
-Arithmetic expressions:
-
-\[a ::= n \mid x \mid a1 + a2 \mid a1 - a2 \mid a1 \times a2\]
-
-Boolean expressions:
-
-\begin{align*}
-	b ::=&\ \mathrm{true} \mid \mathrm{false} \mid a1 = a2 \mid a1 \le a2\\
-	\mid&\ \mathrm{not}\ b \mid b1\ \mathrm{and}\ b2
-\end{align*}
-
-Commands (statements):
-
-\begin{align*}
-	c ::=&\ \mathrm{SKIP}	&\mathrm{(do\ nothing)}\\
-	\mid&\ x ::= a	&\mathrm{(assignment)}\\
-	\mid&\ 1; ; c2	&\mathrm{(sequence)}\\
-	\mid&\ \mathrm{IFB}\ b\ \mathrm{THEN}\ c1\ \mathrm{ELSE}\ c2\ \mathrm{FI}	&\mathrm{(conditional)}\\
-	\mid&\ \mathrm{WHILE}\ b\ \mathrm{DO}\ c\ \mathrm{END}	&\mathrm{(loop)}
-\end{align*}
-
 \section{Related Work}

### 2019-10-30 paper/paper.tex
More cleanup, remove template which is usually packaged

@@
-%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For double-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
-%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
 \documentclass[acmsmall,nonacm]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For single-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
-%% For final camera-ready submission, w/ required CCS and ACM Reference
-%\documentclass[acmsmall]{acmart}\settopmatter{}
 
-%% Journal information
-%% Supplied to authors by publisher for camera-ready submission;
-%% use defaults for review submission.
 \acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
+\acmVolume{}
+\acmNumber{} %
 \acmArticle{}
-\acmYear{2018}
-\acmMonth{1}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
+\acmYear{}
+\acmMonth{}
+\acmDOI{} %
 \startPage{1}
 
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission;
-%% use 'none' for review submission.
 \setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2018}           %% If different from \acmYear
 
-%% Bibliography style
 \bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
+\citestyle{acmauthoryear}   %
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Note: Authors migrating a paper from PACMPL format to traditional
-%% SIGPLAN proceedings format must update the '\documentclass' and
-%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{booktabs}   %
+\usepackage{subcaption} %
 
-%% Some recommended packages.
-\usepackage{booktabs}   %% For formal tables:
-                        %% http://ctan.org/pkg/booktabs
-\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
-                        %% http://ctan.org/pkg/subcaption
-
-%% Added by kartik
 \usepackage[braket]{qcircuit}
@@
 
-%% Title information
-\title[Quantum Hoare Types]{Quantum Hoare Types}         %% [Short Title] is optional;
-                                        %% when present, will be used in
-                                        %% header instead of Full Title.
-%\titlenote{with title note}             %% \titlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
-%\subtitle{Subtitle}                     %% \subtitle is optional
-%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
+\title[Quantum Hoare Types]{Quantum Hoare Types}         %
 
-%% Author information
-%% Contents and number of authors suppressed with 'anonymous'.
-%% Each author should be introduced by \author, followed by
-%% \authornote (optional), \orcid (optional), \affiliation, and
-%% \email.
-%% An author may have multiple affiliations and/or emails; repeat the
-%% appropriate command.
-%% Many elements are not rendered, but should be provided for metadata
-%% extraction tools.
-
-%% Author with single affiliation.
 \author{Kartik Singhal}
-%\authornote{with author1 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-\orcid{0000-0003-1132-269X}             %% \orcid is optional
+\orcid{0000-0003-1132-269X}             %
 \affiliation{
-%  \position{Position1}
-  \department{Department of Computer Science}              %% \department is recommended
-  \institution{University of Chicago}            %% \institution is required
-%  \streetaddress{Street1 Address1}
-%  \city{City1}
-%  \state{State1}
-%  \postcode{Post-Code1}
-  \country{USA}                    %% \country is recommended
+  \department{Department of Computer Science}              %
+  \institution{University of Chicago}            %
+  \country{USA}                    %
 }
-\email{ks@cs.uchicago.edu}          %% \email is recommended
-
-%% Author with two affiliations and emails.
-%\author{John Reppy}
-%\authornote{with author2 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
-%\affiliation{
-%%  \position{Position2a}
-%  \department{Department of Computer Science}             %% \department is recommended
-%  \institution{University of Chicago}           %% \institution is required
-%%  \streetaddress{Street2a Address2a}
-%%  \city{City2a}
-%%  \state{State2a}
-%%  \postcode{Post-Code2a}
-%  \country{United States}                   %% \country is recommended
-%}
-%\email{jhr@cs.uchicago.edu}         %% \email is recommended
-%\affiliation{
-%  \position{Position2b}
-%  \department{Department2b}             %% \department is recommended
-%  \institution{Institution2b}           %% \institution is required
-%  \streetaddress{Street3b Address2b}
-%  \city{City2b}
-%  \state{State2b}
-%  \postcode{Post-Code2b}
-%  \country{Country2b}                   %% \country is recommended
-%}
-%\email{first2.last2@inst2b.org}         %% \email is recommended
-
-%% Abstract
-%% Note: \begin{abstract}...\end{abstract} environment must come
-%% before \maketitle command
-\begin{abstract}
-(4 sentences)
+\email{ks@cs.uchicago.edu}          %
 
@@
 
-%% 2012 ACM Computing Classification System (CSS) concepts
-%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
 \begin{CCSXML}
@@
 \ccsdesc[500]{Software and its engineering~General programming languages}
-%% End of generated code
 
-%% Keywords
-%% comma separated list
-\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
+\keywords{Quantum Computing, Programming Languages, Type Systems, Dependent Types}  %
 
-%% \maketitle
-%% Note: \maketitle command must come after title commands, author
-%% commands, abstract environment, Computing Classification System
-%% environment and commands, and keywords command.
 \maketitle
@@
 \section{Introduction}
-(1 page)
-
-\textit{- Describe the problem}
-
-\textit{- State your contributions}
+%(1 page)
+%
+%\textit{- Describe the problem}
+%
+%\textit{- State your contributions}
 
@@
 
-\subsection{Contributions}
-(bulleted list of contributions)
-
-\textit{In this section we present the main contributions of the paper\ldots}
+%\subsection{Contributions}
+%(bulleted list of contributions)
+%
+%\textit{In this section we present the main contributions of the paper\ldots}
 
 \section{Problem}
-(1 page)
-
-\textit{Use an example to introduce the problem}
+%(1 page)
+%
+%\textit{Use an example to introduce the problem}
 
@@
 \section{Idea}
-(2 pages)
-
-\textit{The main idea of this paper is \ldots}
+%(2 pages)
+%
+%\textit{The main idea of this paper is \ldots}
 
@@
 \section{Details}
-(5 pages)
+%(5 pages)
 
@@
 \section{Related Work}
-(1-2 pages)
-
-\cite{huang2018}
-
-\cite{huang2019}
-
-\cite{nanevski2008}
-
-\cite{paykin2018}
-
-\cite{rand2018a}
-
-\cite{ross2015algebraic}
-
-\cite{ringer2019}
-
-\cite{silq}
+%(1-2 pages)
 
@@
 
-(0.5 page)
+%(0.5 page)
 
-%% Acknowledgments
-\begin{acks}                            %% acks environment is optional
-                                        %% contents suppressed with 'anonymous'
-  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
-  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
-  %% acknowledge financial support and will be used by metadata
-  %% extraction tools.
+\begin{acks}                            %
   This material is based upon work supported by
@@
 
-%% Bibliography
 \bibliography{bibfile}
 
-%% Appendix
-\appendix
-\section{Appendix}
+%\appendix
+%\section{Appendix}
 
-Text of appendix \ldots
+%Text of appendix \ldots
 

### 2019-10-30 paper/paper.tex
Add abstract

@@
 
-\textit{Text of abstract \ldots.}
+\begin{abstract}
+	Quantum Computing is a new and upcoming way to utilize Quantum Mechanics to do computation.
+It is easy to introduce bugs while programming quantum computers similar to classical computing.
+Strong static type systems have prevented a lot of bugs from being introduced in the classical paradigm.
+There is a similar need for well-designed types in the domain of quantum computing.
+We introduce Quantum Hoare Types that let programmers reason statically about certain semantic properties of their quantum programs and further allow limited form of formal verification.
 

### 2019-10-30 paper/paper.tex
Even more cleanup

@@
-\documentclass[acmsmall,nonacm]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[acmsmall,nonacm,review,timestamp]{acmart}
 
 \acmJournal{PACMPL}
-\acmVolume{}
-\acmNumber{} %
-\acmArticle{}
-\acmYear{}
-\acmMonth{}
-\acmDOI{} %
-\startPage{1}
 
-\setcopyright{none}
+\authorsaddresses{}
 
@@
 
-\title[Quantum Hoare Types]{Quantum Hoare Types}         %
+\title{Quantum Hoare Types}         %
 
@@
 
-\begin{CCSXML}
-<ccs2012>
-<concept>
-<concept_id>10011007.10011006.10011008</concept_id>
-<concept_desc>Software and its engineering~General programming languages</concept_desc>
-<concept_significance>500</concept_significance>
-</concept>
-</ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~General programming languages}
-
 \keywords{Quantum Computing, Programming Languages, Type Systems, Dependent Types}  %
@@
 These advances in quantum type systems although helpful, still fall short in encoding and enforcing even more useful properties that one would like to be able to express for the purpose of verification.
-Inspired by the use of Hoare triples in the verification of imperative programs and building on the idea of Hoare Types in classical programming languages, we propose Quantum Hoare Types aimed at enabling both sound static type checking and formal verification of quantum programs.
+Inspired by the use of Hoare triples in the verification of imperative programs and building on the idea of Hoare Types in classical programming languages\cite{nanevski2008}, we propose Quantum Hoare Types aimed at enabling both sound static type checking and formal verification of quantum programs.
 

### 2019-10-31 paper/paper.tex
Add a todo package, start using stabilizer types
Add some partial examples

@@
 \usepackage{xspace}
+\usepackage[nowatermark]{fixmetodonotes}
 
@@
 \newcommand{\denote}[1]{\llbracket #1 \rrbracket\xspace}
+
+\newcommand{\Z}{\ensuremath{\mathbf{Z}}\xspace}
+\newcommand{\X}{\ensuremath{\mathbf{X}}\xspace}
+\newcommand{\E}{\ensuremath{\mathbf{E}}\xspace}
+\renewcommand{\G}{\ensuremath{\mathbf{G}}\xspace}
+
+\newcommand\hoare[3]{\{\,#1\,\}&#2&\{\,#3\,\}}
 
@@
 
+\TODO{Figure out if stabilizer types fit well as assertions (need to frame them as containment relation, at the least)}
+
 See Fig. \ref{fig:syntax}.
@@
 		$ \mathrm{id}_\tau(M,N) \mid \mathrm{seleq}_\tau(H,M,N) \mid \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid \\
-		P \supset Q \mid \neg P \mid \forall h{:}\mathrm{heap}.P $
+		P \supset Q \mid \neg P \mid \forall h{:}\mathrm{heap}.P \mid \Z \mid \X \mid \E \mid \G $
 	\end{minipage}\\
@@
 
-\textit{Types.}
+\textit{Types.} Primitive types such as qbit, pair, unit, dependent functions, and the Hoare type.
 
-\subsubsection{Assertions}
+\textit{Assertions.} We introduce the notion of stabilizer types \Z, \X, \E, \G as type assertions. \TODO{shouldn't these be really treated as type refinements?}
 
 \subsubsection{Example}
+\begin{enumerate}
+	\item Bell Pair:
+\[ \begin{array}{rcl}
+\hoare{a, b : \Z}{H(a);}{a : \X; b: \Z} \\
+\hoare{a : \X; b: \Z}{CX(a, b);}{(a, b): \E} \\
+\end{array} \]
+
+	\item Superdense coding:
+\[ \begin{array}{rcl}
+\hoare{x, y : \Z; (a, b): \E}{CZ(x,a); \; CX(y, a); \; transmit(A, B);}{x, y : \Z; (a, b): \E} \\
+\hoare{x, y : \Z; (a, b): \E}{CX(a, b);}{x, y : \Z; (a, b): \X \ast \Z} \\
+\hoare{x, y, b : \Z; a: \X}{H(a);}{x, y, a, b : \Z} \\
+\end{array} \]
+
+or for the complete program:
+
+\[ \begin{array}{rcl}
+\hoare{x, y : \Z; (a, b): \E}{ superdense(x, y, a, b);}{x, y, a, b : \Z} \\
+\end{array} \]
+
+\TODO{figure out ownership types and pure vs mixed state types}
+
+	\item Entanglement Swapping (variant of teleportation):\\
+	Initially Alice ($a_1$) and Carol ($c$) share an EPR pair and so do Alice ($a_2$) and Bob ($b$)
+\[ \begin{array}{rcl}
+\hoare{(c, a_1), (a_2, b): \E}{CX(a_1, a_2);}{??} \\
+\hoare{??}{H(a_1)}{??}\\
+\hoare{??}{meas(a_1); \; meas(a_2);}{(c,b): \E; a_1, a_2 : qbit}\\
+\end{array} \]
+
+\item BB84 Protocol for key distribution:
+\end{enumerate}
 

### 2019-11-07 paper/paper.tex
Add partial teleportation example
Add explanation of states for teleport example

@@
 
+\newpage
+
+	\item Quantum Teleportation
+
+	We have three qubits.
+Qubit 0 represents the message in state $\ket{\psi} = \alpha \ket{0} + \beta \ket{1}$.
+Qubit 1 corresponds to Alice and qubit 2 corresponds to Bob and they both start in the state $\ket{0}$.
+To send the message, first step in this protocol is for qubits 1 and 2 to be entangled.
+\[ \begin{array}{rcl}
+\hoare{q_0: qbit = \ket{\psi}; q_1, q_2 : \Z}{H(q_1)}{q_0: qbit; q_2 : \Z; q_1 : \X} \\
+\hoare{q_0: qbit = \ket{\psi}; q_2 : \Z; q_1 : \X}{CX(q_1, q_2)}{q_0: qbit; (q_1,q_2) : \E} \\
+\end{array} \]
+
+Once entangled they are in the state $\ket{\phi^+} = \frac{\ket{0} + \ket{1}}{2}$. The complete quantum state looks like:
+
+\[  \ket{\psi}\ket{\phi^+} = (\alpha\ket{0} + \beta\ket{1})(\frac{1}{\sqrt{2}}(\ket{00} + \ket{11})) \]
+
+or, normalized:
+
+\[  \ket{\psi}\ket{\phi^+} = \frac{\alpha}{\sqrt{2}}\ket{000} + \frac{\alpha}{\sqrt{2}}\ket{011} + \frac{\beta}{\sqrt{2}}\ket{100} + \frac{\beta}{\sqrt{2}}\ket{111} \]
+
+In the second step we apply CNOT to qubits 0 and 1 and then a Hadamard to qubit 0.
+
+\[ \begin{array}{rcl}
+\hoare{q_0: qbit = \ket{\psi}; (q_1,q_2) : \E}{CX(q_0, q_1)}{??} \\
+\hoare{??}{H(q_0)}{??} \\
+\end{array} \]
+
+After the CNOT application, the resulting state is:
+
+\[ \frac{\alpha}{\sqrt{2}}\ket{000} + \frac{\alpha}{\sqrt{2}}\ket{011} + \frac{\beta}{\sqrt{2}}\ket{110} + \frac{\beta}{\sqrt{2}}\ket{101} \]
+
+and after Hadamard application to qubit 0, it becomes:
+
+\[ \frac{\alpha}{\sqrt{2}}(\frac{1}{\sqrt{2}}(\ket{0} + \ket{1}))\ket{00} + \frac{\alpha}{\sqrt{2}}(\frac{1}{\sqrt{2}}(\ket{0} + \ket{1}))\ket{11} + \frac{\beta}{\sqrt{2}}(\frac{1}{\sqrt{2}}(\ket{0} - \ket{1}))\ket{10} + \frac{\beta}{\sqrt{2}}(\frac{1}{\sqrt{2}}(\ket{0} - \ket{1}))\ket{01} \]
+
+This can be simplified to:
+
+\[ \frac{1}{2}\big[\alpha\ket{00\;0} + \beta\ket{00\;1} + \alpha\ket{01\;1} + \beta\ket{01\;0} + \alpha\ket{10\;0} - \beta\ket{10\;1} + \alpha\ket{11\;1} - \beta\ket{11\;0}\big] \]
+
+The next step is the measurement of the first two qubits:
+
+\[ \begin{array}{rcl}
+\hoare{??}{meas(q_0); \; meas(q_1);}{q_0, q_1 : bit, q_2: qbit}\\
+\end{array} \]
+
+This leads to 4 possibilities of getting $(0, 0), (0, 1), (1, 0, (1, 1)$.
+Depending on what we got, we need to make a correction on the resulting state of qubit 2, so we get the original message $\ket{\psi}$.
+This is done by applying X and Z gates conditioned on the two measured bits.
+
+\[ \begin{array}{rcl}
+\hoare{q_0, q_1 : bit, q_2: qbit}{CX(q_1, q_2); \; CZ(q_0, q_2);}{q_0, q_1 : bit, q_2: qbit = \ket{\psi}}\\
+\end{array} \]
+
+\TODO{Need a way to assert that $q_2$ has the same quantum state at the end of the program as $q_0$ did at the beginning}
+
 	\item Entanglement Swapping (variant of teleportation):\\
@@
 \hoare{??}{H(a_1)}{??}\\
-\hoare{??}{meas(a_1); \; meas(a_2);}{(c,b): \E; a_1, a_2 : qbit}\\
+\hoare{??}{meas(a_1); \; meas(a_2);}{(c,b): \E; a_1, a_2 : bit}\\
 \end{array} \]
@@
 \end{enumerate}
+
+\TODO{Can we assign such type refinements to all/most quantum state patterns that have recognizable structure for human programmers?}
 
@@
 
+\listofnotes
+
 %(0.5 page)

### 2019-12-10 ms-paper/ms-paper.tex
Add initial draft of master's thesis with outline

\documentclass[acmsmall,nonacm,timestamp,review]{acmart}

\acmJournal{PACMPL}

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}

\usepackage{booktabs}
\usepackage{subcaption}

\usepackage[braket]{qcircuit}
\usepackage{mathpartir}
\usepackage{mathtools}
\usepackage{xspace}

\usepackage{listings}
\usepackage{lstcoq}

\definecolor{ltblue}{rgb}{0,0.4,0.4}
\definecolor{dkblue}{rgb}{0,0.1,0.6}
\definecolor{dkgreen}{rgb}{0,0.35,0}
\definecolor{dkviolet}{rgb}{0.3,0,0.5}
\definecolor{dkred}{rgb}{0.5,0,0}

\lstset{language=coq, basicstyle=\ttfamily\scriptsize, columns=flexible,
	keepspaces=true}

\newcommand{\sqir}{\ensuremath{s\textsc{qir}}\xspace}
\newcommand{\qwire}{\ensuremath{\mathcal{Q}\textsc{wire}}\xspace}
\newcommand{\denote}[1]{\llbracket #1 \rrbracket\xspace}
\newcommand{\qdenote}[1]{(\!| #1 |\!)\xspace}

\title{Quantum Hoare Types}
\subtitle{A Master's Thesis}

\author{Kartik Singhal}
\orcid{0000-0003-1132-269X}
\affiliation{
	\position{PhD Student}
	\department{Department of Computer Science}
	\institution{University of Chicago}
	\state{IL}
%	\postcode{60637}
	\country{USA}}
\email{ks@cs.uchicago.edu}

\authornote{Homepage: \url{http://ks.cs.uchicago.edu}}

%\authorsaddresses{}

\begin{document}

\begin{abstract}
Quantum Computing is a new and upcoming way to utilize quantum mechanics for computation. It is easy to introduce bugs while programming quantum computers similar to classical computing. Strong static type systems have prevented a lot of bugs from being introduced in the classical paradigm. There is a similar need for well-designed types in the domain of quantum computing. We propose Quantum Hoare Types that let programmers reason statically about certain semantic properties of their quantum programs and further allow a limited form of formal verification.
\end{abstract}

\maketitle

\renewcommand*\contentsname{Table of Contents}
\tableofcontents

\listoffigures

\newpage

\section{Introduction / Background}
It is difficult to reason about the correctness of quantum programs. Sound static type systems help prevent a huge class of bugs from occurring but since the realm of quantum programming is still new there is not a lot of consensus on what kind of types make the most sense. Further, it is unclear how much they help programmers reason about the semantic properties associated with the quantum algorithms that they are implementing.

Recent work~\cite{huang_qdb:_2018,huang_statistical_2019} as part of EPiQC\footnote{EPiQC: Enabling Practical-Scale Quantum Computation:  \url{https://epiqc.cs.uchicago.edu}} has identified several classes of bugs in quantum programs and proposed approaches to tackle them. The technique that holds the most promise is assertion checking using preconditions and postconditions. But assertions are usually checked dynamically during runtime, which can be wasteful of precious quantum computing or simulation resources.

As programming languages researchers, we think it will be more useful to encode such assertions into a static type system both for formal verification and to aid the programmers in writing correct programs from the start. Inspired by the use of Hoare triples in the verification of imperative programs and building on the idea of Hoare Types in classical programming languages~\cite{nanevski_hoare_2008}, we propose Quantum Hoare Types aimed at enabling both sound static type checking and formal verification of quantum programs.

\subsection{Quantum Computing}
The basic unit of information in quantum computing is a qubit, that can be represented using a unit vector in a 2-dimensional complex vector space. Using Dirac's bra-ket notation, we can write an arbitrary qubit, $\ket{\psi}$, as a linear combination of the computational basis vectors: $\ket{\psi} = \alpha\ket{0} + \beta\ket{1}$, where $\ket{0}=\begin{bsmallmatrix}
1 \\
0
\end{bsmallmatrix}$ and $\ket{1}=\begin{bsmallmatrix}
0 \\
1
\end{bsmallmatrix}$, and the two constants are complex amplitudes that are normalized using the constraint $|\alpha|^2 +|\beta|^2 = 1$. This means that when measured the probability of the qubit being in the state 0 is $|\alpha|^2$ and that for the state 1 is $|\beta|^2$. To work with multiple qubits, we need to take their tensor product, denoted by $\otimes$. Computation is performed by the application of unitary quantum gates, such as Hadamard (H) or controlled NOT (CX), that can be represented as square matrices. Mathematically, a gate application is equivalent to multiplying the matrix representing the gate with the state vector of the qubit(s).

This basic unit of information is not the only reason why quantum computing is fundamentally different from classical computing---certain quantum mechanical phenomenon are highly counterintuitive! One such phenomenon is \textit{quantum entanglement} which, at its most basic level, involves a correlation between two qubits such that when one is measured the outcome obtained necessarily influences the measurement outcome of the other qubit (even if the two qubits are far apart). We will see how this can be utilized for computation in the quantum teleportation algorithm that we describe and program in \S\ref{telep}. We refer the reader to the standard textbook~\cite{nielsen_quantum_2010} or Nielsen's excellent series of essays~\cite{matuschak2019} for more background.

\subsection{Type Theory}
\textit{Basics of type theory for non-specialists\ldots}

\subsection{Program Proof}
\textit{Basics of program proof and techniques for non-specialists\ldots}

\section{Hoare Types for Quantum Programming Languages}
In this section we describe Quantum Hoare Types using an example of programming quantum teleportation protocol. We start with the background necessary to understand teleportation and then present how Quantum Hoare Types help us write and verify the  correctness of the protocol.

\subsection{Example: Quantum Teleportation}\label{telep}
Teleportation is a simple quantum algorithm that is used to communicate a quantum state using two classical bits and a shared entangled pair of qubits. Alice is trying to send a message encoded in the qubit to Bob. There are three broad steps in the protocol: 1) preparation, ie, entanglement creation between two qubits to be shared between Alice and Bob in advance; 2) sending the message by Alice; and, 3) reconstruction of the message by Bob. Program and circuit follow:
\begin{figure}[h]
	\begin{subfigure}{3in}
		% 	a       ← init0 ();
		%     b       ← init0 ();
		\begin{lstlisting}
		teleport : Π\kp:qbit.{q↦qbit\kp /\ a↦qbit|0⟩ /\ b↦qbit|0⟩} c:Circuit(qbit⊗qbit⊗qbit, qbit) {b↦qbit\kp} =
		box (q, a, b) ⇒
		(* 1. Bell state preparation for Alice and Bob *)
		a       ← H a;
		(a, b)  ← CX (a, b);
		(* 2. Bell measurement by Alice *)
		(q, a)  ← CX (q, a);
		q       ← H q;
		x       ← meas q;
		y       ← meas a;
		(* 3. Correction by Bob *)
		(y, b)  ← CX (y, b);
		(x, b)  ← CZ (x, b);
		output b
		\end{lstlisting}
		% ()      ← discard y;
		%     ()      ← discard x;
	\end{subfigure}
	\begin{subfigure}{1in}
		% 	\centerline{
		\Qcircuit @C=.4em @R=.4em @! {
			\lstick{q=\ket{\psi}} & \qw & \qw & \ctrl{1} & \gate{H} & \meter & \cw & \control \cw\\
			\lstick{a=\ket{0}} & \gate{H} & \ctrl{1} & \gate{X} & \qw & \meter & \control \cw & \\
			\lstick{b=\ket{0}} & \qw & \gate{X} & \qw & \qw & \qw & \gate{X} \cwx & \gate{Z} \cwx[-2] & \rstick{\ket{\psi}} \qw
			\gategroup{2}{2}{3}{3}{.7em}{.}
			\gategroup{1}{4}{2}{6}{.7em}{.}
		}
		% }
	\end{subfigure}
	\Description{On the left, \qwire code to implement Quantum Teleportation algorithm annotated by its Hoare Type that provides specification for the algorithm. On the right, the corresponding quantum circuit.}
	\caption{Quantum Teleportation (Code and Circuit)}
	\label{fig:teleport}
\end{figure}

In Fig. \ref{fig:teleport} we show a function \lstinline{teleport} that implements the quantum teleportation circuit. It returns one qubit and takes three qubits as arguments: one qubit \lstinline{q} that could be in any unknown state $\ket{\psi}$ which Alice is trying to send to Bob; and two qubits, \lstinline{a} and \lstinline{b}, initially in the state $\ket{0}$. The function first entangles the latter two qubits into a Bell state. Then, Alice performs a Bell measurement of her two qubits \lstinline{q} and \lstinline{a} and sends the resulting classical bits (\lstinline{x} and \lstinline{y}) to Bob. Finally, Bob applies corrective measurement using the quantum gates X and Z controlled by the classical bits that he received from Alice. The result is that Bob's qubit is now in the same state as the initial state of the message qubit \lstinline{q}. The specification that the function must meet is written using a Hoare type, which we describe next.

\subsection{Hoare Types}
A Hoare type encodes preconditions and postconditions in the same spirit as Hoare triples. The type shown on the left below can be read as `for a stateful computation executed in a heap that satisfies precondition $P$, return a value of type $A$ in a heap that satisfies postcondition $Q$.' We show an example, the \lstinline{alloc} primitive from HTT~\cite{nanevski_hoare_2008}, to demonstrate the expressiveness of a Hoare type:
\begin{mathpar}

	\{P\}\; \texttt{x}:A\; \{Q\}

	\and

	\texttt{alloc} : \forall \alpha. \Pi x:\alpha. \{\texttt{emp}\}\; y:\texttt{nat}\; \{y \mapsto_{\alpha} x\}

\end{mathpar}

\begin{figure}[h]
	\begin{tabular}{lrcl}
		\textit{Types} & $\tau, \sigma$ & ::= & $ \mathrm{qbit} \mid \tau \otimes \sigma \mid 1 \mid \Pi x{:}\tau.\sigma \mid \Psi. X. \{P\} x{:}\tau \{Q\}$ \\
		\textit{Assertions} & $P, Q, R$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
			$ \mathrm{id}_\tau(M,N) \mid \mathrm{seleq}_\tau(H,M,N) \mid \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid \\
			P \supset Q \mid \neg P \mid \forall h{:}\mathrm{heap}.P$
		\end{minipage}\\
		\textit{(Quantum) Heaps} & $H, G$ & ::= & $ h \mid \mathrm{empty} \mid \mathrm{upd}_{\tau}(H, M, N) $\\
		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K\ M \mid M : \tau $\\
		\textit{Intro terms} & $M, N, O$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
			$ K \mid (\,) \mid \lambda x.\ M \mid \mathrm{do}\ E \mid \ket{0} \mid \ket{1} \mid (M, N) \mid M==N $
		\end{minipage}\\
		\textit{Commands} & $c$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
			$ \mathrm{alloc}_\tau\ M \mid \mathrm{dealloc}\ M \mid \mathrm{if}_\tau\ M\ \mathrm{then}\ E_1\ \mathrm{else}\ E_2 \mid
			\mathrm{fix}\ f(y{:}\tau){:}\sigma = \mathrm{do}\ E\ \mathrm{in\ eval}\ f M \mid \mathrm{apply}(M, N) \mid \mathrm{capply}(M, N, O) $
		\end{minipage}\\
		\textit{Computations} & $E, F$ & ::= & $ \mathrm{return}\ M \mid x \gets K; E \mid x \Leftarrow c; E \mid x =_\tau M; E $\\
		\textit{Variable context} & $\Delta, \Psi$ & ::= & $ \cdot \mid \Delta, x{:}\tau $\\
		\textit{Heap context} & $X$ & ::= & $ \cdot \mid X, h $\\
		\textit{Assertion context} & $\Gamma$& ::= & $\cdot \mid \Gamma, P $
	\end{tabular}
	\Description{Syntax of Hoare Type Theory from \cite{nanevski_hoare_2008}}
	\caption{Syntax of Hoare Type Theory}
	\label{fig:syntax}
\end{figure}

In order to extend HTT to support quantum circuits, we augment its syntax with that of \qwire\cite{paykin_qwire:_2017,rand_formally_2018}. \qwire is a core quantum circuit language that can cooperate with an arbitrary classical host language. This requires minimal support from the host language to interface with the circuit language. In our case, HTT acts as the classical host language for \qwire.

\qwire's circuit language has a wire type W that can be a unit, a bit, a qubit or a tuple of wires. The host language (HTT) is augmented with additional types to support interaction between the two languages and to treat circuits as data:
\begin{mathpar}

	W ::= 1 \mid \texttt{bit} \mid \texttt{qbit} \mid W1 \otimes W2

	\and

	A ::= \ldots \mid \texttt{Unit} \mid \texttt{Bool} \mid A \times A \mid \texttt{Circuit}(W1, W2)

\end{mathpar}

\subsection{Verification}
Here we describe how HTT lets us verify the correctness of teleportation based on the specification provided as the type of the function. Type checking in HTT involves generation of strongest postconditions at each step of the program with the initial precondition as input. Verification is a two-step process: the first phase does basic type checking and verification condition generation which is decidable, the second phase needs to show the validity of the generated verification conditions, which is undecidable. This second phase can be deferred to an automated theorem prover. For illustration, we show the assertions as we step through step 1 of teleportation. The final state here is that qubits \lstinline{a} and \lstinline{b} are entangled into the first Bell state, $\ket{\phi^+} = \frac{1}{\sqrt{2}}(\ket{00} + \ket{11})$.

% \lstset{aboveskip=1pt,belowskip=1pt}
\begin{lstlisting}
(* P0 (given): {q↦qbit\kp /\ a↦qbit|0⟩ /\ b↦qbit|0⟩} *)
a      ← H a;
(* P1: {q↦qbit\kp /\ a↦qbit1/sqrt2(|0⟩+|1⟩) /\ b↦qbit|0⟩} *)
(a, b) ← CX (a, b);
(* P2: {q↦qbit\kp /\ (a,b)↦qbit|φ+⟩ *)
\end{lstlisting}

\subsection{Soundness / Type Checking Algorithm}

\section{Related Work}
\textit{Objective of this section\ldots}

\subsection{Programming Languages for Quantum Computing}
\textit{A chronology of langugage developments\ldots}

\subsection{Verification of Quantum Programs}
Previous work such as Proto-Quipper~\cite{ross_algebraic_2015,mahmoud_formalization_2019,rios_categorical_2017} and \qwire~\cite{paykin_qwire:_2017,rand_qwire_2017,rand_formally_2018} utilize a linear type system and dependent types to enforce a small subset of semantic properties, such as the no-cloning theorem and whether a unitary gate is of the right dimension. These advances in quantum type systems, although helpful, still fall short in encoding and enforcing even more useful properties that one would like to be able to express for the purpose of verification.

Our approach builds upon previous work in reasoning about quantum programs such as Quantum Weakest Preconditions~\cite{dhondt_quantum_2006} and Quantum Hoare Logic~\cite{ying_floydhoare_2012} in the spirit of Hoare~\cite{hoare_axiomatic_1969} and Dijkstra~\cite{dijkstra_discipline_1976} but attempts to bring those reasoning techniques into the type system. The hope is that programmers will be able to encode some of the semantic properties that they expect of their programs as specifications in their code and type checking will ensure correctness of some of those properties. In the classical setting, Hoare Type Theory (HTT)~\cite{nanevski_hoare_2008} accomplishes exactly this goal. Here, we attempt to merge these ideas for the quantum realm.

\section{Conclusion and Perspectives}
This approach has the potential to be a unified system for programming, specification and reasoning about quantum programs. We need to overcome challenges such as compactly specifying properties of quantum state when dealing with more than a few qubits. We also need to maintain the syntax directed nature of HTT's bidirectional typechecking while incorporating quantum wire types.

We are working on programming other interesting quantum algorithms that can serve as a litmus test for this technique, specifically, a cryptographic protocol called Quantum Key Distribution and the well-known Grover's search algorithm. These examples involve some form of classical control or iteration that makes them non-trivial for our technique.

We also look to explore how our types will evolve when we start incorporating higher-order constructs in our quantum language such as those demonstrated in languages like Quantum $\lambda$-calculus~\cite{selinger_lambda_2006} and Quipper~\cite{green_quipper:_2013}.

Another venue for exploration is to incorporate more precise types that can distinguish between qubits in pure classical state vs. those in superposition vs. those in entanglement inspired by the various quantum resource theories (see \cite{rand_type_2019} and references therein).

\begin{acks}
	I would like to thank my advisor John Reppy for suggesting me to look at Hoare Type Theory that led to this master's thesis and for his constant support and encouragement to see it through. I also thank my committee members Fred Chong and Ravi Chugh for reading drafts of this document and providing useful feedback.

	This material is based upon work supported by
	EPiQC, an \grantsponsor{GS100000001}{NSF}{http://dx.doi.org/10.13039/100000001}
	Expedition in Computing, under Grant
	No.~\grantnum{GS100000001}{1730449}.  Any opinions, findings, and
	conclusions or recommendations expressed in this material are those
	of the authors and do not necessarily reflect the views of the
	National Science Foundation.
\end{acks}

\newpage

\bibliography{references}

\newpage

\appendix
\section{Sample Proof}
\textit{Coq code / Annotated proof objects from implementation\ldots}

\end{document}

### 2019-12-11 ms-paper/ms-paper.tex
Fix typo, remove review option
Minor text refinements

@@
-\documentclass[acmsmall,nonacm,timestamp,review]{acmart}
+\documentclass[acmsmall,nonacm,timestamp]{acmart}
 
@@
 Strong static type systems have prevented a lot of bugs from being introduced in the classical paradigm.
-There is a similar need for well-designed types in the domain of quantum computing.
+There is a similar need for well-designed static types in the domain of quantum computing.
 We propose Quantum Hoare Types that let programmers reason statically about certain semantic properties of their quantum programs and further allow a limited form of formal verification.
@@
 
-\section{Introduction / Background}
+\section{Introduction}
 It is difficult to reason about the correctness of quantum programs.
@@
 
+\section{Background}
+
 \subsection{Quantum Computing}
-The basic unit of information in quantum computing is a qubit, that can be represented using a unit vector in a 2-dimensional complex vector space.
+The basic unit of information in quantum computing is a qubit, that can be represented using a unit vector in a two-dimensional complex vector space.
 Using Dirac's bra-ket notation, we can write an arbitrary qubit, $\ket{\psi}$, as a linear combination of the computational basis vectors: $\ket{\psi} = \alpha\ket{0} + \beta\ket{1}$, where $\ket{0}=\begin{bsmallmatrix}
@@
 \end{bsmallmatrix}$, and the two constants are complex amplitudes that are normalized using the constraint $|\alpha|^2 +|\beta|^2 = 1$.
-This means that when measured the probability of the qubit being in the state 0 is $|\alpha|^2$ and that for the state 1 is $|\beta|^2$.
+This means that when measured, the probability of the qubit being in the state 0 is $|\alpha|^2$ and that for the state 1 is $|\beta|^2$.
 To work with multiple qubits, we need to take their tensor product, denoted by $\otimes$.
-Computation is performed by the application of unitary quantum gates, such as Hadamard (H) or controlled NOT (CX), that can be represented as square matrices.
+Computation is performed by the application of unitary quantum gates, such as the Hadamard (H) gate or the controlled NOT (CX) gate, that can be represented as square matrices.
 Mathematically, a gate application is equivalent to multiplying the matrix representing the gate with the state vector of the qubit(s).
@@
 This basic unit of information is not the only reason why quantum computing is fundamentally different from classical computing---certain quantum mechanical phenomenon are highly counterintuitive!
-One such phenomenon is \textit{quantum entanglement} which, at its most basic level, involves a correlation between two qubits such that when one is measured the outcome obtained necessarily influences the measurement outcome of the other qubit (even if the two qubits are far apart).
+One such phenomenon is \textit{quantum entanglement} which, at its most basic level, involves a correlation between two qubits such that when one is measured, the outcome obtained necessarily influences the measurement outcome of the other qubit (even if the two qubits are far apart).
 We will see how this can be utilized for computation in the quantum teleportation algorithm that we describe and program in \S\ref{telep}.
@@
 There are three broad steps in the protocol: 1) preparation, ie, entanglement creation between two qubits to be shared between Alice and Bob in advance; 2) sending the message by Alice; and, 3) reconstruction of the message by Bob.
-Program and circuit follow:
 \begin{figure}[h]
@@
 In Fig.
-\ref{fig:teleport} we show a function \lstinline{teleport} that implements the quantum teleportation circuit.
+\ref{fig:teleport}, on the left, we show a function \lstinline{teleport} written in \qwire-like syntax (TODO: add ref) that implements the quantum teleportation circuit shown on the right.
 It returns one qubit and takes three qubits as arguments: one qubit \lstinline{q} that could be in any unknown state $\ket{\psi}$ which Alice is trying to send to Bob; and two qubits, \lstinline{a} and \lstinline{b}, initially in the state $\ket{0}$.
@@
 This second phase can be deferred to an automated theorem prover.
-For illustration, we show the assertions as we step through step 1 of teleportation.
+For illustration, we show the assertions as we step through the first step of teleportation.
 The final state here is that qubits \lstinline{a} and \lstinline{b} are entangled into the first Bell state, $\ket{\phi^+} = \frac{1}{\sqrt{2}}(\ket{00} + \ket{11})$.
@@
 \subsection{Programming Languages for Quantum Computing}
-\textit{A chronology of langugage developments\ldots}
+\textit{A chronology of language developments\ldots}
 
@@
 
-Our approach builds upon previous work in reasoning about quantum programs such as Quantum Weakest Preconditions~\cite{dhondt_quantum_2006} and Quantum Hoare Logic~\cite{ying_floydhoare_2012} in the spirit of Hoare~\cite{hoare_axiomatic_1969} and Dijkstra~\cite{dijkstra_discipline_1976} but attempts to bring those reasoning techniques into the type system.
+Our approach builds upon previous work in reasoning about quantum programs such as Quantum Weakest Preconditions~\cite{dhondt_quantum_2006} and Quantum Hoare Logic~\cite{ying_floydhoare_2012} in the spirit of \citet{hoare_axiomatic_1969} and \citet{dijkstra_discipline_1976} but attempts to bring those reasoning techniques into the type system.
 The hope is that programmers will be able to encode some of the semantic properties that they expect of their programs as specifications in their code and type checking will ensure correctness of some of those properties.
@@
 
-We also look to explore how our types will evolve when we start incorporating higher-order constructs in our quantum language such as those demonstrated in languages like Quantum $\lambda$-calculus~\cite{selinger_lambda_2006} and Quipper~\cite{green_quipper:_2013}.
+We also look to explore how our types will evolve when we start incorporating higher-order constructs in our quantum language such as those demonstrated in languages like quantum lambda calculus~\cite{selinger_lambda_2006} and Quipper~\cite{green_quipper:_2013}.
 

### 2019-12-15 ms-paper/ms-paper.tex
Add John's suggestions, some minor bib updates

@@
 
-\subsection{Type Theory}
-\textit{Basics of type theory for non-specialists\ldots}
+\subsection{Hoare Style Reasoning}
+\textit{Basics of Hoare-triples for non-specialists\ldots}
 
-\subsection{Program Proof}
+\subsection{Program Proof and HTT}
 \textit{Basics of program proof and techniques for non-specialists\ldots}
+
+\section{Running Example}
+\textit{Move teleportation example here}
 
@@
 
+\section{Evaluation}
+
+\subsection{More Examples}
+\textit{add at least one example with scaling knob}
+
+\subsection{Circuit-complexity Considerations}
+
 \section{Related Work}
@@
 
-\subsection{Programming Languages for Quantum Computing}
-\textit{A chronology of language developments\ldots}
+\subsection{Quantum Languages and Types}
+\textit{A chronology of language developments wrt to usefulness of types\ldots}
 
@@
 Here, we attempt to merge these ideas for the quantum realm.
+
+\textit{Add from Mingsheng Ying's book, papers, etc.}
 
@@
 
+\textit{Future work about using Proto-Quipper/\qwire as languages, then extraction into QASM and checking generated verification conditions.}
+
 \begin{acks}
-	I would like to thank my advisor John Reppy for suggesting me to look at Hoare Type Theory that led to this master's thesis and for his constant support and encouragement to see it through.
+%	I would like to thank my advisor John Reppy for suggesting me to look at Hoare Type Theory that led to this master's thesis and for his constant support and encouragement to see it through.
 I also thank my committee members Fred Chong and Ravi Chugh for reading drafts of this document and providing useful feedback.

### 2020-02-09 ms-paper/ms-paper.tex
Improve syntax

@@
 \newcommand{\qdenote}[1]{(\!| #1 |\!)\xspace}
+
+\newcommand{\type}[1]{\textrm{\textbf{#1}}}
 
@@
 	\begin{tabular}{lrcl}
-		\textit{Types} & $\tau, \sigma$ & ::= & $ \mathrm{qbit} \mid \tau \otimes \sigma \mid 1 \mid \Pi x{:}\tau.\sigma \mid \Psi. X. \{P\} x{:}\tau \{Q\}$ \\
-		\textit{Assertions} & $P, Q, R$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
-			$ \mathrm{id}_\tau(M,N) \mid \mathrm{seleq}_\tau(H,M,N) \mid \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid \\
-			P \supset Q \mid \neg P \mid \forall h{:}\mathrm{heap}.P$
+		\textit{Types} & $A, B, C$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
+			$ \type{Unit} \mid \type{Bit} \mid \type{Qubit} \mid A \otimes B \mid \type{Nat} \mid \Pi x{ : }A.B \mid ~!A  \mid \type{Circ}(S, U) \mid \{P\} x{ : }A \{Q\}$
 					\end{minipage}\\
-		\textit{(Quantum) Heaps} & $H, G$ & ::= & $ h \mid \mathrm{empty} \mid \mathrm{upd}_{\tau}(H, M, N) $\\
-		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K\ M \mid M : \tau $\\
-		\textit{Intro terms} & $M, N, O$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
-			$ K \mid (\,) \mid \lambda x.\ M \mid \mathrm{do}\ E \mid \ket{0} \mid \ket{1} \mid (M, N) \mid M==N $
+		\textit{Simple Types} & $S, U$ & ::= & $ \type{Unit} \mid \type{Bit} \mid \type{Qubit} \mid S \otimes U$ \\
+		\textit{Propositions} & $P, Q, R$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
+			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid P \supset Q \mid \neg P \mid \exists x{:}A.P \mid \forall x.P$
 		\end{minipage}\\
-		\textit{Commands} & $c$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
-			$ \mathrm{alloc}_\tau\ M \mid \mathrm{dealloc}\ M \mid \mathrm{if}_\tau\ M\ \mathrm{then}\ E_1\ \mathrm{else}\ E_2 \mid
-			\mathrm{fix}\ f(y{:}\tau){:}\sigma = \mathrm{do}\ E\ \mathrm{in\ eval}\ f M \mid \mathrm{apply}(M, N) \mid \mathrm{capply}(M, N, O) $
+%		\textit{(Quantum) Heaps} & $H, G$ & ::= & $ h \mid \mathrm{empty} \mid \mathrm{upd}_{\tau}(H, M, N) $\\
+%		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K\ M \mid M : \tau $\\
+		\textit{Terms} & $M, N$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
+			$ c \mid l \mid x \mid \lambda x.M \mid MN \mid \mathrm{unit} \mid (a, \mathcal{C}, b) \mid \mathrm{unbox}~M \mid \mathrm{force}~M \mid \mathrm{lift}~M \mid \mathrm{box_U}~M \mid (M, N) \mid \mathrm{let}~ (x, y) = N ~\mathrm{in}~ M $
 		\end{minipage}\\
-		\textit{Computations} & $E, F$ & ::= & $ \mathrm{return}\ M \mid x \gets K; E \mid x \Leftarrow c; E \mid x =_\tau M; E $\\
-		\textit{Variable context} & $\Delta, \Psi$ & ::= & $ \cdot \mid \Delta, x{:}\tau $\\
-		\textit{Heap context} & $X$ & ::= & $ \cdot \mid X, h $\\
-		\textit{Assertion context} & $\Gamma$& ::= & $\cdot \mid \Gamma, P $
+		\textit{Simple Terms} & $a, b, c$ & ::= & $ l \mid \mathrm{unit} \mid (a, b)$ \\
+		\textit{Circuits} & $\mathcal{C}, \mathcal{D}$ & : & $ \Sigma \to \Sigma' $ \\
+		\textit{Label Contexts} & $\Sigma$ & ::= & $ \cdot \mid l : \type{Qubit}, \Sigma$ \\
+%		\textit{Commands} & $c$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
+%			$ \mathrm{alloc}_\tau\ M \mid \mathrm{dealloc}\ M \mid \mathrm{if}_\tau\ M\ \mathrm{then}\ E_1\ \mathrm{else}\ E_2 \mid
+%			\mathrm{fix}\ f(y{:}\tau){:}\sigma = \mathrm{do}\ E\ \mathrm{in\ eval}\ f M \mid \mathrm{apply}(M, N) \mid \mathrm{capply}(M, N, O) $
+%		\end{minipage}\\
+%		\textit{Computations} & $E, F$ & ::= & $ \mathrm{return}\ M \mid x \gets K; E \mid x \Leftarrow c; E \mid x =_\tau M; E $\\
+%		\textit{Variable context} & $\Delta, \Psi$ & ::= & $ \cdot \mid \Delta, x{:}\tau $\\
+%		\textit{Heap context} & $X$ & ::= & $ \cdot \mid X, h $\\
+		\textit{Contexts} & $\Gamma$ & ::= & $ \cdot \mid x : A, \Gamma \mid l : \type{Qubit}, \Sigma$ \\
+		\textit{Values} & $V$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
+	$ x \mid l \mid \lambda x.M \mid \mathrm{lift}~M \mid (a, \mathcal{C}, b) \mid \mathrm{unbox}~V \mid \mathrm{force}(\mathrm{unbox}~V) \mid $
+\end{minipage}\\
 	\end{tabular}

### 2020-02-11 ms-paper/ms-paper.tex
Minor clarifications
Almost finalize syntax, add types for simple gates

@@
 
-\begin{figure}[h]
+\begin{figure}[t]
 	\begin{tabular}{lrcl}
 		\textit{Types} & $A, B, C$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
-			$ \type{Unit} \mid \type{Bit} \mid \type{Qubit} \mid A \otimes B \mid \type{Nat} \mid \Pi x{ : }A.B \mid ~!A  \mid \type{Circ}(S, U) \mid \{P\} x{ : }A \{Q\}$
+			$ \type{Unit} \mid \type{Bit} \mid \type{Qubit} \mid A \otimes B \mid \Pi x{ : }A.B \mid ~!A  \mid \type{Circ}(S, U) \mid \{P\} x{ : }A \{Q\}$
 					\end{minipage}\\
-		\textit{Simple Types} & $S, U$ & ::= & $ \type{Unit} \mid \type{Bit} \mid \type{Qubit} \mid S \otimes U$ \\
+		\textit{Wire Types} & $S, U$ & ::= & $ \type{Unit} \mid \type{Bit} \mid \type{Qubit} \mid S \otimes U$ \\
 		\textit{Propositions} & $P, Q, R$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
-			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid P \supset Q \mid \neg P \mid \exists x{:}A.P \mid \forall x.P$
+			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid P \supset Q \mid \neg P \mid \exists x{:}A.P \mid \forall x{:}A.P \mid \mathrm{Id_A}(M, N)$
 		\end{minipage}\\
@@
 		\textit{Terms} & $M, N$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
-			$ c \mid l \mid x \mid \lambda x.M \mid MN \mid \mathrm{unit} \mid (a, \mathcal{C}, b) \mid \mathrm{unbox}~M \mid \mathrm{force}~M \mid \mathrm{lift}~M \mid \mathrm{box_U}~M \mid (M, N) \mid \mathrm{let}~ (x, y) = N ~\mathrm{in}~ M $
+			$ c \mid l \mid x \mid \lambda x.M \mid MN \mid \mathrm{unit} \mid (a, \mathcal{C}, b) \mid \mathrm{box_S}~M \mid \mathrm{unbox}~M \mid \mathrm{force}~M \mid \mathrm{lift}~M \mid (M, N) \mid \mathrm{let}~ (x, y) = N ~\mathrm{in}~ M $
 		\end{minipage}\\
-		\textit{Simple Terms} & $a, b, c$ & ::= & $ l \mid \mathrm{unit} \mid (a, b)$ \\
+		\textit{Simple Terms} & $a, b$ & ::= & $ l \mid \mathrm{unit} \mid (a, b)$ \\
 		\textit{Circuits} & $\mathcal{C}, \mathcal{D}$ & : & $ \Sigma \to \Sigma' $ \\
+		\textit{Proposition Contexts} & $\Delta$ & ::= & $ \cdot \mid P, \Delta$ \\
 		\textit{Label Contexts} & $\Sigma$ & ::= & $ \cdot \mid l : \type{Qubit}, \Sigma$ \\
@@
 %		\textit{Heap context} & $X$ & ::= & $ \cdot \mid X, h $\\
-		\textit{Contexts} & $\Gamma$ & ::= & $ \cdot \mid x : A, \Gamma \mid l : \type{Qubit}, \Sigma$ \\
+		\textit{Contexts} & $\Gamma$ & ::= & $ \cdot \mid x : A, \Gamma \mid l : \type{Qubit}, \Gamma$ \\
 		\textit{Values} & $V$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
@@
 	\Description{Syntax of Hoare Type Theory from \cite{nanevski_hoare_2008}}
-	\caption{Syntax of Hoare Type Theory}
+	\caption{Syntax of Quantum (Linear) Hoare Type Theory}
 	\label{fig:syntax}
@@
 
-In order to extend HTT to support quantum circuits, we augment its syntax with that of \qwire\cite{paykin_qwire:_2017,rand_formally_2018}.
+Figure \ref{fig:syntax} shows syntax of Quantum HTT, inspired by the syntax of Proto-Quipper for the quantum parts.
+Symbol $c$ ranges over constants of the language such as primitive quantum gates (Hadamard, X, Y, Z, CNOT, etc), while $l$ identifies wires (bits or qubits).
+$!A$ represents duplicable/reusable data.
+The Pi type can be simplified to $A \multimap B$ if $B$ does not depend on $x$.
+$\mathrm{Id_A}(M, N)$ is propositional equality.
+$\mathrm{box_S}$ converts a circuit-producing function of type $S \multimap U$ into a circuit treated as data of type $\type{Circ}(S, U)$; $\mathrm{unbox}$ does the reverse.
+
+%In order to extend HTT to support quantum circuits, we augment its syntax with that of \qwire\cite{paykin_qwire:_2017,rand_formally_2018}.
 \qwire is a core quantum circuit language that can cooperate with an arbitrary classical host language.
@@
 In our case, HTT acts as the classical host language for \qwire.
+%
+%\qwire's circuit language has a wire type W that can be a unit, a bit, a qubit or a tuple of wires.
+The host language (HTT) is augmented with additional types to support interaction between the two languages and to treat circuits as data:
+%\begin{mathpar}
+%
+%	W ::= 1 \mid \texttt{bit} \mid \texttt{qbit} \mid W1 \otimes W2
+%
+%	\and
+%
+%	A ::= \ldots \mid \texttt{Unit} \mid \texttt{Bool} \mid A \times A \mid \texttt{Circuit}(W1, W2)
+%
+%\end{mathpar}
 
-\qwire's circuit language has a wire type W that can be a unit, a bit, a qubit or a tuple of wires.
-The host language (HTT) is augmented with additional types to support interaction between the two languages and to treat circuits as data:
+The primitive quantum gates that are treated as constants in the language have types such as follows:
 \begin{mathpar}
-
-	W ::= 1 \mid \texttt{bit} \mid \texttt{qbit} \mid W1 \otimes W2
-
+	H : \type{Qubit} \multimap \type{Qubit}
 	\and
-
-	A ::= \ldots \mid \texttt{Unit} \mid \texttt{Bool} \mid A \times A \mid \texttt{Circuit}(W1, W2)
-
+	CNOT : \type{Qubit} \otimes \type{Qubit} \multimap \type{Qubit}  \otimes \type{Qubit}
 \end{mathpar}

### 2020-02-12 ms-paper/ms-paper.tex
Add cointoss example

@@
 		\textit{Terms} & $M, N$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
-			$ c \mid l \mid x \mid \lambda x.M \mid MN \mid \mathrm{unit} \mid (a, \mathcal{C}, b) \mid \mathrm{box_S}~M \mid \mathrm{unbox}~M \mid \mathrm{force}~M \mid \mathrm{lift}~M \mid (M, N) \mid \mathrm{let}~ (x, y) = N ~\mathrm{in}~ M $
+			$ c \mid l \mid x \mid \lambda x.M \mid MN \mid \mathrm{unit} \mid (a, \mathcal{C}, b) \mid \mathrm{box_S}~M \mid \mathrm{unbox}~M \mid \mathrm{force}~M \mid \mathrm{lift}~M \mid (M, N) \mid \mathrm{let}~ x = N ~\mathrm{in}~ M $
 		\end{minipage}\\
@@
 \begin{mathpar}
-	H : \type{Qubit} \multimap \type{Qubit}
+	\mathrm{H} : \type{Qubit} \multimap \type{Qubit}
 	\and
-	CNOT : \type{Qubit} \otimes \type{Qubit} \multimap \type{Qubit}  \otimes \type{Qubit}
+	\mathrm{CNOT} : \type{Qubit} \otimes \type{Qubit} \multimap \type{Qubit}  \otimes \type{Qubit}
+	\and
+	\mathrm{Init0} : \type{Unit} \multimap \type{Qubit}
+	\and
+	\mathrm{meas} : \type{Qubit} \multimap \type{Bit}
 \end{mathpar}
+
+Simple quantum programs can be written using the above syntax, for example, cointoss:
+
+\begin{lstlisting}
+cointoss : Unit ⊸ Bit
+cointoss = let
+					q = Init0 unit
+					q = H q
+					b = meas q
+				in b
+\end{lstlisting}
+
+To write a Hoare type, we need to box the circuit as follows:
+
+\begin{lstlisting}
+boxed_cointoss : {⊤} c : Circ(Unit, Bit) {⊤}
+boxed_cointoss = box_unit cointoss
+\end{lstlisting}
+
+The propositions shown above are trivial. More interesting examples will follow.
 

### 2020-02-14 ms-paper/ms-paper.tex
Add a line for lift and force
Minor suggested changes incorporated

@@
 
-As programming languages researchers, we think it will be more useful to encode such assertions into a static type system both for formal verification and to aid the programmers in writing correct programs from the start.
+As programming language researchers, we think it will be more useful to encode such assertions into a static type system both for formal verification and to aid the programmers in writing correct programs from the start.
 Inspired by the use of Hoare triples in the verification of imperative programs and building on the idea of Hoare Types in classical programming languages~\cite{nanevski_hoare_2008}, we propose Quantum Hoare Types aimed at enabling both sound static type checking and formal verification of quantum programs.
@@
 One such phenomenon is \textit{quantum entanglement} which, at its most basic level, involves a correlation between two qubits such that when one is measured, the outcome obtained necessarily influences the measurement outcome of the other qubit (even if the two qubits are far apart).
-We will see how this can be utilized for computation in the quantum teleportation algorithm that we describe and program in \S\ref{telep}.
+We will see how this property can be utilized for computation in the quantum teleportation algorithm that we describe and program in \S\ref{telep}.
 We refer the reader to the standard textbook~\cite{nielsen_quantum_2010} or Nielsen's excellent series of essays~\cite{matuschak2019} for more background.
@@
 \section{Running Example}
-\textit{Move teleportation example here}
-
-\section{Hoare Types for Quantum Programming Languages}
-In this section we describe Quantum Hoare Types using an example of programming quantum teleportation protocol.
-We start with the background necessary to understand teleportation and then present how Quantum Hoare Types help us write and verify the  correctness of the protocol.
 
@@
 The specification that the function must meet is written using a Hoare type, which we describe next.
+\section{Hoare Types for Quantum Programming Languages}
+In this section we describe Quantum Hoare Types using an example of programming quantum teleportation protocol.
+We start with the background necessary to understand teleportation and then present how Quantum Hoare Types help us write and verify the  correctness of the protocol.
 
@@
 $\mathrm{box_S}$ converts a circuit-producing function of type $S \multimap U$ into a circuit treated as data of type $\type{Circ}(S, U)$; $\mathrm{unbox}$ does the reverse.
+lift and force convert between linear and reusable types.
 
@@
 	\mathrm{CNOT} : \type{Qubit} \otimes \type{Qubit} \multimap \type{Qubit}  \otimes \type{Qubit}
-	\and
+	\\
 	\mathrm{Init0} : \type{Unit} \multimap \type{Qubit}

### 2020-02-20 ms-paper/ms-paper.tex
More rearrangement

@@
 
-\section{Running Example}
+\section{Hoare Types for Quantum Programming Languages}
+% In this section we describe Quantum Hoare Types using an example of programming quantum teleportation protocol.
+We start with the background necessary to understand teleportation and then present how Quantum Hoare Types help us write and verify the  correctness of the protocol.
+
+\subsection{Hoare Types}
+A Hoare type encodes preconditions and postconditions in the same spirit as Hoare triples.
+The type shown on the left below can be read as `for a stateful computation executed in a heap that satisfies precondition $P$, return a value of type $A$ in a heap that satisfies postcondition $Q$.' We show an example, the \lstinline{alloc} primitive from HTT~\cite{nanevski_hoare_2008}, to demonstrate the expressiveness of a Hoare type:
+\begin{mathpar}
+
+	\{P\}\; \texttt{x}:A\; \{Q\}
+
+	\and
+
+	\texttt{alloc} : \forall \alpha. \Pi x:\alpha. \{\texttt{emp}\}\; y:\texttt{nat}\; \{y \mapsto_{\alpha} x\}
+
+\end{mathpar}
+
+Figure \ref{fig:syntax} shows syntax of Quantum HTT, inspired by the syntax of Proto-Quipper for the quantum parts.
+Symbol $c$ ranges over constants of the language such as primitive quantum gates (Hadamard, X, Y, Z, CNOT, etc), while $l$ identifies wires (bits or qubits).
+$!A$ represents duplicable/reusable data.
+The Pi type can be simplified to $A \multimap B$ if $B$ does not depend on $x$.
+$\mathrm{Id_A}(M, N)$ is propositional equality.
+$\mathrm{box_S}$ converts a circuit-producing function of type $S \multimap U$ into a circuit treated as data of type $\type{Circ}(S, U)$; $\mathrm{unbox}$ does the reverse.
+lift and force convert between linear and reusable types.
+
+\begin{figure}[h]
+	\begin{tabular}{lrcl}
+		\textit{Types} & $A, B, C$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
+			$ \type{Unit} \mid \type{Bit} \mid \type{Qubit} \mid A \otimes B \mid \Pi x{ : }A.B \mid ~!A  \mid \type{Circ}(S, U) \mid \{P\} x{ : }A \{Q\}$
+					\end{minipage}\\
+		\textit{Wire Types} & $S, U$ & ::= & $ \type{Unit} \mid \type{Bit} \mid \type{Qubit} \mid S \otimes U$ \\
+		\textit{Propositions} & $P, Q, R$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
+			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid P \supset Q \mid \neg P \mid \exists x{:}A.P \mid \forall x{:}A.P \mid \mathrm{Id_A}(M, N)$
+		\end{minipage}\\
+%		\textit{(Quantum) Heaps} & $H, G$ & ::= & $ h \mid \mathrm{empty} \mid \mathrm{upd}_{\tau}(H, M, N) $\\
+%		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K\ M \mid M : \tau $\\
+		\textit{Terms} & $M, N$ & ::= & \begin{minipage}[t]{0.55\columnwidth}%
+			$ c \mid l \mid x \mid \lambda x.M \mid MN \mid \mathrm{unit} \mid (a, \mathcal{C}, b) \mid \mathrm{box_S}~M \mid \mathrm{unbox}~M \mid \mathrm{force}~M \mid \mathrm{lift}~M \mid (M, N) \mid \mathrm{let}~ x = N ~\mathrm{in}~ M $
+		\end{minipage}\\
+		\textit{Simple Terms} & $a, b$ & ::= & $ l \mid \mathrm{unit} \mid (a, b)$ \\
+		\textit{Circuits} & $\mathcal{C}, \mathcal{D}$ & : & $ \Sigma \to \Sigma' $ \\
+		\textit{Proposition Contexts} & $\Delta$ & ::= & $ \cdot \mid P, \Delta$ \\
+		\textit{Label Contexts} & $\Sigma$ & ::= & $ \cdot \mid l : \type{Qubit}, \Sigma$ \\
+%		\textit{Commands} & $c$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
+%			$ \mathrm{alloc}_\tau\ M \mid \mathrm{dealloc}\ M \mid \mathrm{if}_\tau\ M\ \mathrm{then}\ E_1\ \mathrm{else}\ E_2 \mid
+%			\mathrm{fix}\ f(y{:}\tau){:}\sigma = \mathrm{do}\ E\ \mathrm{in\ eval}\ f M \mid \mathrm{apply}(M, N) \mid \mathrm{capply}(M, N, O) $
+%		\end{minipage}\\
+%		\textit{Computations} & $E, F$ & ::= & $ \mathrm{return}\ M \mid x \gets K; E \mid x \Leftarrow c; E \mid x =_\tau M; E $\\
+%		\textit{Variable context} & $\Delta, \Psi$ & ::= & $ \cdot \mid \Delta, x{:}\tau $\\
+%		\textit{Heap context} & $X$ & ::= & $ \cdot \mid X, h $\\
+		\textit{Contexts} & $\Gamma$ & ::= & $ \cdot \mid x : A, \Gamma \mid l : \type{Qubit}, \Gamma$ \\
+		\textit{Values} & $V$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
+	$ x \mid l \mid \lambda x.M \mid \mathrm{lift}~M \mid (a, \mathcal{C}, b) \mid \mathrm{unbox}~V \mid \mathrm{force}(\mathrm{unbox}~V) \mid $
+\end{minipage}\\
+	\end{tabular}
+	\Description{Syntax of Hoare Type Theory from \cite{nanevski_hoare_2008}}
+	\caption{Syntax of Quantum (Linear) Hoare Type Theory}
+	\label{fig:syntax}
+\end{figure}
+
+%In order to extend HTT to support quantum circuits, we augment its syntax with that of \qwire\cite{paykin_qwire:_2017,rand_formally_2018}.
+\qwire is a core quantum circuit language that can cooperate with an arbitrary classical host language.
+This requires minimal support from the host language to interface with the circuit language.
+In our case, HTT acts as the classical host language for \qwire.
+%
+%\qwire's circuit language has a wire type W that can be a unit, a bit, a qubit or a tuple of wires.
+The host language (HTT) is augmented with additional types to support interaction between the two languages and to treat circuits as data:
+%\begin{mathpar}
+%
+%	W ::= 1 \mid \texttt{bit} \mid \texttt{qbit} \mid W1 \otimes W2
+%
+%	\and
+%
+%	A ::= \ldots \mid \texttt{Unit} \mid \texttt{Bool} \mid A \times A \mid \texttt{Circuit}(W1, W2)
+%
+%\end{mathpar}
+
+The primitive quantum gates that are treated as constants in the language have types such as follows:
+\begin{mathpar}
+	\mathrm{H} : \type{Qubit} \multimap \type{Qubit}
+	\and
+	\mathrm{CNOT} : \type{Qubit} \otimes \type{Qubit} \multimap \type{Qubit}  \otimes \type{Qubit}
+	\\
+	\mathrm{Init0} : \type{Unit} \multimap \type{Qubit}
+	\and
+	\mathrm{meas} : \type{Qubit} \multimap \type{Bit}
+\end{mathpar}
+
+Simple quantum programs can be written using the above syntax, for example, cointoss:
+
+\begin{lstlisting}
+cointoss : Unit ⊸ Bit
+cointoss = let
+					q = Init0 unit
+					q = H q
+					b = meas q
+				in b
+\end{lstlisting}
+
+To write a Hoare type, we need to box the circuit as follows:
+
+\begin{lstlisting}
+boxed_cointoss : {⊤} c : Circ(Unit, Bit) {⊤}
+boxed_cointoss = box_unit cointoss
+\end{lstlisting}
+
+The propositions shown above are trivial. More interesting examples will follow.
 
@@
 The result is that Bob's qubit is now in the same state as the initial state of the message qubit \lstinline{q}.
-The specification that the function must meet is written using a Hoare type, which we describe next.
-\section{Hoare Types for Quantum Programming Languages}
-In this section we describe Quantum Hoare Types using an example of programming quantum teleportation protocol.
-We start with the background necessary to understand teleportation and then present how Quantum Hoare Types help us write and verify the  correctness of the protocol.
-
-\subsection{Hoare Types}
-A Hoare type encodes preconditions and postconditions in the same spirit as Hoare triples.
-The type shown on the left below can be read as `for a stateful computation executed in a heap that satisfies precondition $P$, return a value of type $A$ in a heap that satisfies postcondition $Q$.' We show an example, the \lstinline{alloc} primitive from HTT~\cite{nanevski_hoare_2008}, to demonstrate the expressiveness of a Hoare type:
-\begin{mathpar}
-
-	\{P\}\; \texttt{x}:A\; \{Q\}
-
-	\and
-
-	\texttt{alloc} : \forall \alpha. \Pi x:\alpha. \{\texttt{emp}\}\; y:\texttt{nat}\; \{y \mapsto_{\alpha} x\}
-
-\end{mathpar}
-
-\begin{figure}[t]
-	\begin{tabular}{lrcl}
-		\textit{Types} & $A, B, C$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
-			$ \type{Unit} \mid \type{Bit} \mid \type{Qubit} \mid A \otimes B \mid \Pi x{ : }A.B \mid ~!A  \mid \type{Circ}(S, U) \mid \{P\} x{ : }A \{Q\}$
-					\end{minipage}\\
-		\textit{Wire Types} & $S, U$ & ::= & $ \type{Unit} \mid \type{Bit} \mid \type{Qubit} \mid S \otimes U$ \\
-		\textit{Propositions} & $P, Q, R$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
-			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid P \supset Q \mid \neg P \mid \exists x{:}A.P \mid \forall x{:}A.P \mid \mathrm{Id_A}(M, N)$
-		\end{minipage}\\
-%		\textit{(Quantum) Heaps} & $H, G$ & ::= & $ h \mid \mathrm{empty} \mid \mathrm{upd}_{\tau}(H, M, N) $\\
-%		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K\ M \mid M : \tau $\\
-		\textit{Terms} & $M, N$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
-			$ c \mid l \mid x \mid \lambda x.M \mid MN \mid \mathrm{unit} \mid (a, \mathcal{C}, b) \mid \mathrm{box_S}~M \mid \mathrm{unbox}~M \mid \mathrm{force}~M \mid \mathrm{lift}~M \mid (M, N) \mid \mathrm{let}~ x = N ~\mathrm{in}~ M $
-		\end{minipage}\\
-		\textit{Simple Terms} & $a, b$ & ::= & $ l \mid \mathrm{unit} \mid (a, b)$ \\
-		\textit{Circuits} & $\mathcal{C}, \mathcal{D}$ & : & $ \Sigma \to \Sigma' $ \\
-		\textit{Proposition Contexts} & $\Delta$ & ::= & $ \cdot \mid P, \Delta$ \\
-		\textit{Label Contexts} & $\Sigma$ & ::= & $ \cdot \mid l : \type{Qubit}, \Sigma$ \\
-%		\textit{Commands} & $c$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
-%			$ \mathrm{alloc}_\tau\ M \mid \mathrm{dealloc}\ M \mid \mathrm{if}_\tau\ M\ \mathrm{then}\ E_1\ \mathrm{else}\ E_2 \mid
-%			\mathrm{fix}\ f(y{:}\tau){:}\sigma = \mathrm{do}\ E\ \mathrm{in\ eval}\ f M \mid \mathrm{apply}(M, N) \mid \mathrm{capply}(M, N, O) $
-%		\end{minipage}\\
-%		\textit{Computations} & $E, F$ & ::= & $ \mathrm{return}\ M \mid x \gets K; E \mid x \Leftarrow c; E \mid x =_\tau M; E $\\
-%		\textit{Variable context} & $\Delta, \Psi$ & ::= & $ \cdot \mid \Delta, x{:}\tau $\\
-%		\textit{Heap context} & $X$ & ::= & $ \cdot \mid X, h $\\
-		\textit{Contexts} & $\Gamma$ & ::= & $ \cdot \mid x : A, \Gamma \mid l : \type{Qubit}, \Gamma$ \\
-		\textit{Values} & $V$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
-	$ x \mid l \mid \lambda x.M \mid \mathrm{lift}~M \mid (a, \mathcal{C}, b) \mid \mathrm{unbox}~V \mid \mathrm{force}(\mathrm{unbox}~V) \mid $
-\end{minipage}\\
-	\end{tabular}
-	\Description{Syntax of Hoare Type Theory from \cite{nanevski_hoare_2008}}
-	\caption{Syntax of Quantum (Linear) Hoare Type Theory}
-	\label{fig:syntax}
-\end{figure}
-
-Figure \ref{fig:syntax} shows syntax of Quantum HTT, inspired by the syntax of Proto-Quipper for the quantum parts.
-Symbol $c$ ranges over constants of the language such as primitive quantum gates (Hadamard, X, Y, Z, CNOT, etc), while $l$ identifies wires (bits or qubits).
-$!A$ represents duplicable/reusable data.
-The Pi type can be simplified to $A \multimap B$ if $B$ does not depend on $x$.
-$\mathrm{Id_A}(M, N)$ is propositional equality.
-$\mathrm{box_S}$ converts a circuit-producing function of type $S \multimap U$ into a circuit treated as data of type $\type{Circ}(S, U)$; $\mathrm{unbox}$ does the reverse.
-lift and force convert between linear and reusable types.
-
-%In order to extend HTT to support quantum circuits, we augment its syntax with that of \qwire\cite{paykin_qwire:_2017,rand_formally_2018}.
-\qwire is a core quantum circuit language that can cooperate with an arbitrary classical host language.
-This requires minimal support from the host language to interface with the circuit language.
-In our case, HTT acts as the classical host language for \qwire.
-%
-%\qwire's circuit language has a wire type W that can be a unit, a bit, a qubit or a tuple of wires.
-The host language (HTT) is augmented with additional types to support interaction between the two languages and to treat circuits as data:
-%\begin{mathpar}
-%
-%	W ::= 1 \mid \texttt{bit} \mid \texttt{qbit} \mid W1 \otimes W2
-%
-%	\and
-%
-%	A ::= \ldots \mid \texttt{Unit} \mid \texttt{Bool} \mid A \times A \mid \texttt{Circuit}(W1, W2)
-%
-%\end{mathpar}
-
-The primitive quantum gates that are treated as constants in the language have types such as follows:
-\begin{mathpar}
-	\mathrm{H} : \type{Qubit} \multimap \type{Qubit}
-	\and
-	\mathrm{CNOT} : \type{Qubit} \otimes \type{Qubit} \multimap \type{Qubit}  \otimes \type{Qubit}
-	\\
-	\mathrm{Init0} : \type{Unit} \multimap \type{Qubit}
-	\and
-	\mathrm{meas} : \type{Qubit} \multimap \type{Bit}
-\end{mathpar}
-
-Simple quantum programs can be written using the above syntax, for example, cointoss:
-
-\begin{lstlisting}
-cointoss : Unit ⊸ Bit
-cointoss = let
-					q = Init0 unit
-					q = H q
-					b = meas q
-				in b
-\end{lstlisting}
-
-To write a Hoare type, we need to box the circuit as follows:
-
-\begin{lstlisting}
-boxed_cointoss : {⊤} c : Circ(Unit, Bit) {⊤}
-boxed_cointoss = box_unit cointoss
-\end{lstlisting}
-
-The propositions shown above are trivial. More interesting examples will follow.
+The specification that the function must meet is written using a Hoare type, which we described in the previous section.
 

### 2020-04-22 ms-paper/ms-paper.tex
Transfer from QPL draft; minor changes

--- 2020-02-12 ms-paper/ms-paper.tex
@@
 \usepackage{xspace}
+\usepackage{fancyvrb}
+\usepackage{cleveref}
+\crefformat{section}{\S#2#1#3}
 
@@
 \newcommand{\type}[1]{\textrm{\textbf{#1}}}
+\newcommand{\kw}[1]{\,\mathrm{\texttt{#1}}\,}
+\newcommand{\todo}[1]{\textcolor{red}{#1}}
 
-\title{Quantum Hoare Types}
-\subtitle{A Master's Thesis}
+\newcommand{\cket}[1]{{$\ket{#1}$}}
+\newcommand\cin{{$\in$}}
+\newcommand\cmt{{$\mapsto$}}
+\newcommand\cmm{{$\multimap$}}
+\newcommand\ccc{{$\circ$}}
+\newcommand{\cpi}{$\Pi$}
+
+\title{Quantum Hoare Type Theory}
+\subtitle{Master's Paper}
 
@@
 \begin{abstract}
-Quantum Computing is a new and upcoming way to utilize quantum mechanics for computation.
+	% Foreward
+	%Context and Need
+	As quantum computers become real, it is high time we come up with effective techniques that help programmers write correct quantum programs.
+	%Need
+	In classical computing, formal verification and sound static type systems prevent several classes of bugs from being introduced. There is a need for similar techniques in the quantum regime.
+	%Task
+	Inspired by Hoare Type Theory in the classical paradigm, we propose Quantum Hoare Types by extending the Quantum IO Monad by indexing it with pre- and post-conditions that serve as program specifications.
+	%Object
+	In this paper, we introduce Quantum Hoare Type Theory (QHTT), present its syntax and typing rules, and demonstrate its effectiveness with the help of examples.
+	% Summary
+	%Findings
+
+	%Conclusion
+	QHTT has the potential to be a unified system for programming, specifying, and reasoning about quantum programs.
+	%Perspectives
+%	This is a work in progress.
+%Quantum Computing is a new and upcoming way to utilize quantum mechanics for computation.
 It is easy to introduce bugs while programming quantum computers similar to classical computing.
@@
 
-\listoffigures
+%\listoffigures
 
-\newpage
+%\newpage
 
@@
 
-As programming languages researchers, we think it will be more useful to encode such assertions into a static type system both for formal verification and to aid the programmers in writing correct programs from the start.
-Inspired by the use of Hoare triples in the verification of imperative programs and building on the idea of Hoare Types in classical programming languages~\cite{nanevski_hoare_2008}, we propose Quantum Hoare Types aimed at enabling both sound static type checking and formal verification of quantum programs.
+As programming languages researchers, we think it will be better to encode such assertions into a static type system both for formal verification and to aid the programmers in writing correct programs from the start.
+Inspired by the use of Hoare triples in the verification of imperative programs and building on the idea of Hoare Type Theory (HTT) for classical programming languages~\cite{nanevski_hoare_2008}, we propose Quantum Hoare Type Theory (by extending the Quantum IO Monad interface\cite{qio}) aimed at enabling both sound static type checking and formal verification of quantum programs.
+
+The main idea is that instead of just indexing the QIO monad type with the type of computation result, we can also index it with pre-conditions and post-conditions so as to integrate Hoare-style reasoning into the type system itself.
+The resulting Quantum Hoare Type, $\{P\} x{ : }A \{Q\}$, specifies the preconditions, $P$, that hold on the quantum state before execution; the result, $x$, and its type, $A$, on successful execution; and, the postconditions, $Q$, that are true for the quantum state after successful execution.
+In this way the effectful quantum fragment of the program is effectively encapsulated inside the Quantum Hoare monad.
+Our theory, like HTT, allows the usual equational reasoning for the pure classical fragments of the program and uses syntax-directed type checking for generating strongest postconditions for the quantum fragment.
+
+This paper is organized as follows.
+We discuss background in the next section (\cref{sec:bg}); specifically, quantum computing (\cref{sec:qc}),  Hoare Type Theory in the classical setting (\cref{sec:htt}) and then the Quantum IO Monad interface (\cref{sec:qio}).
+Then we introduce our contributions in merging these ideas together for specification and enforcement of useful semantic properties in section \cref{sec:qhtt} along with its syntax (\cref{sec:syntax}), typing (\cref{sec:typing}) and, some examples (\cref{sec:examples}).
+Some related work follows in \cref{related}.
+Finally, we conclude and share ideas for future work in the final section (\cref{sec:conclusion}).
 
 \section{Background}
+\label{sec:bg}
+In this section, we provide background on the core ideas from existing literature that form the foundation for our work -- Hoare Type Theory (HTT) and the Quantum IO Monad interface.
+We start with some basics of quantum computing.
+% and refer the reader to Nielsen's excellent series of essays~\cite{matuschak2019} for a first introduction and to the standard textbook~\cite{nielsen2010} for advanced material.
 
 \subsection{Quantum Computing}
+\label{sec:qc}
 The basic unit of information in quantum computing is a qubit, that can be represented using a unit vector in a two-dimensional complex vector space.
@@
 One such phenomenon is \textit{quantum entanglement} which, at its most basic level, involves a correlation between two qubits such that when one is measured, the outcome obtained necessarily influences the measurement outcome of the other qubit (even if the two qubits are far apart).
-We will see how this can be utilized for computation in the quantum teleportation algorithm that we describe and program in \S\ref{telep}.
+%We will see how this property can be utilized for computation in the quantum teleportation algorithm that we describe and program in \S\ref{telep}.
 We refer the reader to the standard textbook~\cite{nielsen_quantum_2010} or Nielsen's excellent series of essays~\cite{matuschak2019} for more background.
 
-\subsection{Hoare Style Reasoning}
-\textit{Basics of Hoare-triples for non-specialists\ldots}
+\subsection{Hoare Type Theory}
+\label{sec:htt}
+A Hoare type encodes preconditions and postconditions in the same spirit as Hoare triples to allow both specification and verification of effectful classical programs.
+The type shown on the left below can be read as `for a stateful computation executed in a heap that satisfies precondition $P$, return a value of type $A$ in a heap that satisfies postcondition $Q$.' We show an example, the \texttt{alloc} primitive from HTT~\cite{nanevski_hoare_2008}, to demonstrate the expressiveness of a Hoare type:
+\begin{mathpar}
+	\{P\}\; \texttt{x}:A\; \{Q\}
 
-\subsection{Program Proof and HTT}
-\textit{Basics of program proof and techniques for non-specialists\ldots}
+	\and
 
-\section{Running Example}
-\textit{Move teleportation example here}
+	\texttt{alloc} : \forall \alpha. \Pi x:\alpha. \{\texttt{emp}\}\; y:\texttt{nat}\; \{y \mapsto_{\alpha} x\}
+\end{mathpar}
 
-\section{Hoare Types for Quantum Programming Languages}
-In this section we describe Quantum Hoare Types using an example of programming quantum teleportation protocol.
-We start with the background necessary to understand teleportation and then present how Quantum Hoare Types help us write and verify the  correctness of the protocol.
+Type checking in HTT involves generation of strongest postconditions at each step of the program.
+Verification is a two-step process: the first phase does basic type checking and verification condition generation which is decidable, the second phase needs to show the validity of the generated verification conditions, which is undecidable.
+This second phase can be deferred to an automated theorem prover.
 
-\subsection{Example: Quantum Teleportation}\label{telep}
-Teleportation is a simple quantum algorithm that is used to communicate a quantum state using two classical bits and a shared entangled pair of qubits.
+We provide more details when we introduce QHTT in section \cref{sec:qhtt}.
+
+\subsection{Quantum IO Monad}
+\label{sec:qio}
+
+The Quantum IO (QIO) monad~\cite{qio} is a purely functional interface for quantum programming that provides a separation between the unitary (reversible) and non-unitary (irreversible) fragments of quantum computation.
+It provides isolation of quantum effects inside a monad similar to what we need to provide in our Quantum Hoare monad.
+
+Here we discuss the relevant bits from the QIO interface.
+QIO provides a type for referring to qubits, \texttt{Qbit}; a type for unitary operations, \texttt{U}; and, the \texttt{QIO} type operator which is a monad indexed by the result of the quantum computation.
+The primitive quantum operations defined using the QIO monad below (Haskell syntax) are:
+\begin{verbatim}
+mkQbit : Bool -> QIO Qbit       -- initialization
+applyU : U -> QIO ()            -- apply a unitary to quantum state
+measQbit : Qbit -> QIO Bool     -- measurement
+\end{verbatim}
+
+Further, \texttt{U} is defined as a monoid with neutral element, \texttt{mempty}, and operation, \texttt{mappend}, that encodes reversible operations on quantum state.
+Here are the core unitary primitives that we will need:
+
+\begin{verbatim}
+rot :: Qbit -> ((Bool, Bool) -> C) -> U
+cond :: Qbit -> (Bool -> U) -> U
+\end{verbatim}
+
+Here, \texttt{rot} takes a qubit and a two-by-two complex-valued unitary matrix (represented as a function from the matrix indices to $\mathbb{C}$) and lets one define arbitrary rotation on a single qubit.
+\texttt{rot} can then be used to define standard gates such as the hadamard, H, and the Pauli gates, X, Y and Z.
+\texttt{cond} is the unitary conditional that is used to perform branching unitary operations.
+For the present paper, we use a simpler conditional, \texttt{ifQ}, that is defined as follows:
+
+\begin{verbatim}
+ifQ :: Qbit -> U -> U
+ifQ q u = cond q (\x -> if x then u else mempty)
+\end{verbatim}
+
+That is, \texttt{ifQ} acts as the standard control operator in the quantum circuit model that runs its second argument based on the truth value of its first argument.
+Then, it is easy to define controlled gates such as CNOT with the expression $\kw{ifQ q1 (X q2)}$.
+
+Quantum state in the QIO interface is modeled as a normalized vector that intuitively stores pairs representing complex amplitudes associated with each basis state (represented as a map from Qbit to Bool types).
+The operations defined over the vector class ensure that the quantum state is kept normalized throughout.
+
+QIO further ensures that the monoidal structure of unitary operations lets one run a unitary operation over the complete quantum state.
+We elide details here and refer the reader to \cite{qio} for more.
+
+%\subsection{Hoare Style Reasoning}
+%\textit{Basics of Hoare-triples for non-specialists\ldots}
+%
+%\subsection{Program Proof and HTT}
+%\textit{Basics of program proof and techniques for non-specialists\ldots}
+
+\section{Quantum Hoare Type Theory}
+\label{sec:qhtt}
+In this section, we introduce the Quantum Hoare Type inspired by the QIO monad and its type theory (QHTT) by replacing the classical effectful portion of Hoare Type Theory with quantum effects.
+The core idea is to encapsulate any quantum effect inside a monadic Quantum Hoare type, $\Delta.
+\Psi .\{P\} x{ : }A \{Q\}$, and formalize reasoning of quantum effects using strongest postconditions in a similar fashion as HTT in the classical setting.
+
+The next two sections discuss the syntax and typing of QHTT. The last section show examples written in QHTT.
+
+\subsection{Syntax}
+\label{sec:syntax}
+
+Figure \ref{fig:syntax} shows the syntax of Quantum Hoare Type Theory. Our presentation closely follows that of HTT~\cite{nanevski_hoare_2008}.
+
+Complete grammar written using Ott tool at \url{https://github.com/k4rtik/QHT/blob/master/ott/qhtt.ott}.
+
+\begin{figure}[ht]
+	\begin{tabular}{lrcl}
+		\textit{Types} & $A, B, C$ & ::= &
+		\begin{minipage}[t]{0.45\columnwidth}%
+			$ \type{1} \mid \type{Bool} \mid \type{Qbit} \mid \type{U} \mid \type{Pure} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
+		\end{minipage}\\ \\
+		\textit{Assertions} & $P, Q, R$ & ::= &
+		\begin{minipage}[t]{0.45\columnwidth}%
+			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid P \supset Q \mid \neg P \mid \exists x{:}A.P \mid \forall x{:}A.P \mid \exists h{:}\kw{heap}.P \mid \forall h{:}\kw{heap}.P \mid \kw{Id}_A(M, N) \mid \kw{HId}(H, G) \mid \kw{indom}(H,M) $
+		\end{minipage}\\ \\
+		\textit{Q. Heaps} & $H, G$ & ::= & $ h \mid \kw{empty} \mid \kw{upd}(H, M, N) $\\ \\
+		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K M \mid M : A $\\
+		\textit{Intro terms} & $M, N, O$ & ::= & $ K \mid () \mid \lambda x.M \mid \kw{do} E \mid \kw{true} \mid \kw{false} $\\ \\
+		\textit{Q. Commands} & $c$ & ::= &
+		\begin{minipage}[t]{0.45\columnwidth}%
+			$ \kw{mkQbit} M \mid \kw{measQbit} M \mid \kw{applyU} M \mid \kw{if} M \kw{then} N \kw{else} O$
+		\end{minipage}\\
+		\textit{Computations} & $E, F$ & ::= & $ \kw{return} M \mid x \leftarrow K; E \mid x \Leftarrow c; E \mid x =_A M; E $\\ \\
+		\textit{Variable context} & $\Delta$ & ::= & $ \cdot \mid x : A, \Delta $ \\
+		\textit{Q. Heap context} & $\Psi$ & ::= & $ \cdot \mid h, \Psi$ \\
+		\textit{Assertion context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
+	\end{tabular}
+	\Description{Syntax of Quantum Hoare Type Theory}
+	\caption{Syntax of Quantum Hoare Type Theory}
+	\label{fig:syntax}
+\end{figure}
+
+\paragraph{Types} We include primitive types for unit, booleans, qubits and unitary operations, and a type \type{Pure} for representing pure quantum state vectors (complex valued vectors in Hilbert space); these correspond to similar types in the QIO work.
+There are also type constructors for pairs, dependent functions and the Hoare type from HTT.
+
+\paragraph{Assertions}
+Apart from the usual first order logic, we have assertions for reasoning about propositional equality of terms, $\kw{Id}_A(M, N)$, and heaps, $\kw{HId}(H, G)$.
+
+As done in HTT, some common convenience assertions can be defined using the base primitives.
+Particularly, $\kw{emp}$ denotes that the heap is empty.
+M $\mapsto$ N represents that this is the only mapping in the heap.
+M $\hookrightarrow$ N represents that looking up M in the heap returns N.
+$\kw{seleq}$(H, M, N) asserts that in heap H, location M points to N.
+
+\paragraph{Quantum State}
+
+We use the terms quantum state and quantum heaps interchangeably throughout.
+A quantum heap is a partial function from qubits (that can in turn be thought of as locations, represented using natural numbers) to at most one quantum state vector.
+
+For the purpose of the type theory, heaps are functional, so that, for example, $\kw{upd}(H, M, N)$ returns a new heap after updating heap H at location M with N.
+
+\paragraph{Terms}
+The terms are divided into introduction and elimination sorts for the purpose of bidirectional typechecking like in HTT.
+
+The pure fragment allows higher-order functions while the impure (effectful) fragment is encapsulated within the monadic \texttt{do} $E$ construct and supports a simple imperative quantum language inspired by QIO constructs.
+It is worth noting that \texttt{do} $E$ represents suspended computation and is considered pure.
+
+\paragraph{Quantum Commands}
+These are the quantum-specific effectful commands from QIO for initialization, unitary application and measurement of qubits.
+We also support classical control using the \texttt{if} $M$ \texttt{then} $N$ \texttt{else} $O$ construct similar to HTT.
+
+\paragraph{Computations} The computations are the usual monadic return and (three) bind operations as in HTT.
+
+\paragraph{Contexts} Contexts are modelled exactly the same way as in HTT. The only difference is that instead of classical heaps, we have quantum heaps.
+
+\subsection{Typing Rules}
+\label{sec:typing}
+
+In this section, we provide some details about type checking in HTT and how it is different in the quantum case.
+
+\subsubsection{Judgments}
+The type system is designed to be bidirectional and syntax directed and hences include separate judgements for intro and elim terms. We are able to use them unchanged for QHTT.
+
+The type checking process also involves generating canonical forms for each term.
+A canonical form in HTT means a beta-normal (containing no beta-redexes) and eta-long (all intro terms are eta-expanded) form.
+We elide much detail here.
+
+As usual elim terms synthesize types ($\Rightarrow$) and intro terms are checked against the given type ($\Leftarrow$). Canonical forms are always synthesized. For the pure fragment:
+
+\begin{center}
+	\begin{tabular}{ll}
+		$\Delta \vdash K \Rightarrow A [N']$ & Elim term K has type A and canonical form $N'$ \\
+		$\Delta \vdash M \Leftarrow A [M']$ & Intro term M has type A and canonical form $M'$
+	\end{tabular}
+\end{center}
+
+The judgments for computations involve synthesizing the strongest postconditions and checking whether a given postcondition applies:
+
+\begin{center}
+	\begin{tabular}{ll}
+		$\Delta; P \vdash E \Rightarrow x:A.Q[E']$ & Computation E with precondition P has \textit{strongest} postcondition Q\\ & and returns value x of type A. Its canonical form is $E'$. \\
+		$\Delta; P \vdash E \Leftarrow x:A.Q[E']$ & Computation E with precondition P has postcondition Q\\ & and returns value x of type A. Its canonical form is $E'$.
+	\end{tabular}
+\end{center}
+
+The type system includes rules for primitive effectful commands and those for structuring composition such as monadic unit and bind.
+We reuse most of the HTT rules except those for primitives for quantum effects that need to specify the strongest postconditions for each primitive, we discuss that next.
+
+\subsubsection{Strongest Postconditions}
+Here we show the strongest postconditions for the primitive quantum commands of QHTT encoded in their typing rules.
+Given a quantum command for initialization, unitary application or measurement, its strongest postcondition is an assertion that most precisely captures the relationship between the initial state and the modified state after the execution.
+These rules along with the \texttt{consq} (consequent) rule of HTT (that ensures weakening of the given strongest postcondition to an arbitrary postcondition) work together to ensure composition of verification.
+
+We need to use the relational composition connective from HTT ($P \circ Q$) that captures how heap evolves with computation.
+It basically reads: $Q$ holds of the current heap which is obtained after modification of a prior heap for which P holds.
+
+We also use HTT's difference operator ($\multimap$) below that captures changes to only the interesting fragment of the heap without modifying the rest of the heap.
+
+We can now show some typing rules:
+
+\paragraph{Initialization}
+$x \Leftarrow \kw{mkQbit} M; E$
+
+\begin{mathpar}
+	\inferrule
+	{\Delta \vdash M : \type{Bool}
+		\\ \Delta, x : \type{Qbit}; P \circ (x \mapsto \kw{state}(M)) \vdash E \Rightarrow y:B . Q}
+	{\Delta; P \vdash x \Leftarrow \kw{mkQbit} M; E \Rightarrow y:B. (\exists x:\type{Qbit}.Q)}
+\end{mathpar}
+where x is fresh and \texttt{state} is a function that translates a classical representation of quantum state (such as boolean here) to the equivalent quantum state vector.
+
+\paragraph{Measurement}
+$x \Leftarrow \kw{measQbit} M; E$
+
+\begin{mathpar}
+	\inferrule
+	{\Delta \vdash M : \type{Qbit}
+		\\ \Delta \vdash P \supset (M \hookrightarrow -)
+		\\ \Delta; P \circ ((M \mapsto -) \multimap \kw{emp}) \vdash E \Rightarrow y:B . Q}
+	{\Delta; P \vdash x \Leftarrow \kw{measQbit} M; E \Rightarrow y:B. Q}
+\end{mathpar}
+
+In measurement, we first need to ensure that the precondition P implies that M is allocated in the heap and that the fragment of heap that only refers to location M becomes empty.
+
+As may be apparent, initialization and measurement are analogous to HTT's \texttt{alloc} and \texttt{dealloc} primitives.
+We are working out precise details for unitary application, but at a high-level, they involve ensuring that the given unitary term actually represents a unitary matrix and updating the quantum state with the result of unitary application.
+This involves incorporating the monoidal structure of unitaries into our theory (as is done in QIO).
+This step also involves generating verification conditions that cannot be checked during this first phase of typechecking.
+
+\subsection{Sample programs}
+\label{sec:examples}
+
+In this section, we show some simple examples from the QIO work~\cite{qio} translated in our language and annotated with their program specifications using their Quantum Hoare types.
+An observation is that translating the programs was a very simple process except for coming up with the right specifications for the programs.
+We take some liberty in writing our typeset syntax as plain text for convenience and provide some commentary on assertions specified for these programs.
+
+At the end of this section, we show a sample verification using QHTT for Bell pair generation.
+
+\subsubsection*{Hello Quantum World}
+\begin{verbatim}
+hqw : {emp} r : Bool {emp /\ Id(r, false)}
+    = do q <= mkQbit false;
+         measQbit q
+\end{verbatim}
+
+We start out in an empty state and assert that the result is equal to \texttt{false}.
+
+\subsubsection*{Coin Toss}
+\begin{verbatim}
+rnd : {emp} r : Bool {emp}
+    = do q <= mkQbit false;
+         applyU (H q);
+         measQbit q
+\end{verbatim}
+
+As the result can be in either of its two values \texttt{false} or \texttt{true}, we need not specify any special postcondition as type checking the result will be sufficient for correctness.
+
+\subsubsection*{Testing Bell Pair}
+\begin{Verbatim}[commandchars=\$\[\]]
+testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}  -- a,b $cin {true, false}
+         = do qa <= mkQbit false;
+              applyU (H qa);
+              qb <= mkQbit false;
+              applyU (ifQ qa (X qb));
+              (measQbit qa, measQbit qb)
+\end{Verbatim}
+
+Here, we are asserting that the returned booleans $a$ and $b$ hold the same value, which is what we expect from a Bell pair.
+
+\paragraph{Verification}
+We would like to describe how QHTT lets us verify the correctness of the above program based on the specification provided as the type of the function.
+For illustration, we show the assertions annotated as comments as we step through the \texttt{testBell} program below.
+We have elided specific details about unitary application and definition of the previously mentioned \texttt{state} function as they are still being worked out.
+
+\begin{Verbatim}[commandchars=\$\[\]]
+testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}
+              -- P0: emp
+         = do qa <= mkQbit false;
+              -- P1: P0 $ccc (qa $cmt state(false))
+              applyU (H qa);
+              -- P2: P1 $ccc (qa $cmt state({(0.5, false), (0.5, true)}))
+              qb <= mkQbit false;
+              -- P3: P2 $ccc (qa $cmt state(false))
+              applyU (ifQ qa (X qb));
+              -- P4: P3 $ccc ((qa, qb) $cmt state({(0.5, (false, false),
+              --                              (0.5, (true, true))}))
+                (measQbit qa, measQbit qb)
+              -- P5: P4 $ccc ((qa $cmt -) $cmm emp) $ccc ((qb $cmt -) $cmm emp)
+\end{Verbatim}
+Finally, the \texttt{consq} rule should ensure that the assertion P5 can lead to the specified postcondition in the type of the program.
+
+We show some two more examples in the appendix for a modular version of Bell pair generation (\cref{app:mbp}) that shows use of higher-order features of QHTT and another for teleportation protocol (\cref{app:teleport}).
+Both of these involve specifying assertions over quantum state whose theory we still need to work out.
+
+%Figure \ref{fig:syntax} shows syntax of Quantum HTT, inspired by the syntax of Proto-Quipper for the quantum parts.
+Symbol $c$ ranges over constants of the language such as primitive quantum gates (Hadamard, X, Y, Z, CNOT, etc), while $l$ identifies wires (bits or qubits).
+$!A$ represents duplicable/reusable data.
+The Pi type can be simplified to $A \multimap B$ if $B$ does not depend on $x$.
+$\mathrm{Id_A}(M, N)$ is propositional equality.
+$\mathrm{box_S}$ converts a circuit-producing function of type $S \multimap U$ into a circuit treated as data of type $\type{Circ}(S, U)$; $\mathrm{unbox}$ does the reverse.
+lift and force convert between linear and reusable types.
+
+%The primitive quantum gates that are treated as constants in the language have types such as follows:
+%\begin{mathpar}
+%	\mathrm{H} : \type{Qubit} \multimap \type{Qubit}
+%	\and
+%	\mathrm{CNOT} : \type{Qubit} \otimes \type{Qubit} \multimap \type{Qubit}  \otimes \type{Qubit}
+%	\\
+%	\mathrm{Init0} : \type{Unit} \multimap \type{Qubit}
+%	\and
+%	\mathrm{meas} : \type{Qubit} \multimap \type{Bit}
+%\end{mathpar}
+%
+%Simple quantum programs can be written using the above syntax, for example, cointoss:
+%
+%\begin{lstlisting}
+%cointoss : Unit ⊸ Bit
+%cointoss = let
+%					q = Init0 unit
+%					q = H q
+%					b = meas q
+%				in b
+%\end{lstlisting}
+%
+%To write a Hoare type, we need to box the circuit as follows:
+%
+%\begin{lstlisting}
+%boxed_cointoss : {⊤} c : Circ(Unit, Bit) {⊤}
+%boxed_cointoss = box_unit cointoss
+%\end{lstlisting}
+%
+%The propositions shown above are trivial. More interesting examples will follow.
+%
+%\subsection{Example: Quantum Teleportation}\label{telep}
+%Teleportation is a simple quantum algorithm that is used to communicate a quantum state using two classical bits and a shared entangled pair of qubits.
 Alice is trying to send a message encoded in the qubit to Bob.
 There are three broad steps in the protocol: 1) preparation, ie, entanglement creation between two qubits to be shared between Alice and Bob in advance; 2) sending the message by Alice; and, 3) reconstruction of the message by Bob.
-\begin{figure}[h]
-	\begin{subfigure}{3in}
-		% 	a       ← init0 ();
-		%     b       ← init0 ();
-		\begin{lstlisting}
-		teleport : Π\kp:qbit.{q↦qbit\kp /\ a↦qbit|0⟩ /\ b↦qbit|0⟩} c:Circuit(qbit⊗qbit⊗qbit, qbit) {b↦qbit\kp} =
-		box (q, a, b) ⇒
-		(* 1. Bell state preparation for Alice and Bob *)
-		a       ← H a;
-		(a, b)  ← CX (a, b);
-		(* 2. Bell measurement by Alice *)
-		(q, a)  ← CX (q, a);
-		q       ← H q;
-		x       ← meas q;
-		y       ← meas a;
-		(* 3. Correction by Bob *)
-		(y, b)  ← CX (y, b);
-		(x, b)  ← CZ (x, b);
-		output b
-		\end{lstlisting}
-		% ()      ← discard y;
-		%     ()      ← discard x;
-	\end{subfigure}
-	\begin{subfigure}{1in}
-		% 	\centerline{
-		\Qcircuit @C=.4em @R=.4em @! {
-			\lstick{q=\ket{\psi}} & \qw & \qw & \ctrl{1} & \gate{H} & \meter & \cw & \control \cw\\
-			\lstick{a=\ket{0}} & \gate{H} & \ctrl{1} & \gate{X} & \qw & \meter & \control \cw & \\
-			\lstick{b=\ket{0}} & \qw & \gate{X} & \qw & \qw & \qw & \gate{X} \cwx & \gate{Z} \cwx[-2] & \rstick{\ket{\psi}} \qw
-			\gategroup{2}{2}{3}{3}{.7em}{.}
-			\gategroup{1}{4}{2}{6}{.7em}{.}
-		}
+%\begin{figure}[h]
+%	\begin{subfigure}{3in}
+%		% 	a       ← init0 ();
+%		%     b       ← init0 ();
+%		\begin{lstlisting}
+%		teleport : Π\kp:qbit.{q↦qbit\kp /\ a↦qbit|0⟩ /\ b↦qbit|0⟩} c:Circuit(qbit⊗qbit⊗qbit, qbit) {b↦qbit\kp} =
+%		box (q, a, b) ⇒
+%		(* 1. Bell state preparation for Alice and Bob *)
+%		a       ← H a;
+%		(a, b)  ← CX (a, b);
+%		(* 2. Bell measurement by Alice *)
+%		(q, a)  ← CX (q, a);
+%		q       ← H q;
+%		x       ← meas q;
+%		y       ← meas a;
+%		(* 3. Correction by Bob *)
+%		(y, b)  ← CX (y, b);
+%		(x, b)  ← CZ (x, b);
+%		output b
+%		\end{lstlisting}
+%		% ()      ← discard y;
+%		%     ()      ← discard x;
+%	\end{subfigure}
+%	\begin{subfigure}{1in}
+%		% 	\centerline{
+%		\Qcircuit @C=.4em @R=.4em @! {
+%			\lstick{q=\ket{\psi}} & \qw & \qw & \ctrl{1} & \gate{H} & \meter & \cw & \control \cw\\
+%			\lstick{a=\ket{0}} & \gate{H} & \ctrl{1} & \gate{X} & \qw & \meter & \control \cw & \\
+%			\lstick{b=\ket{0}} & \qw & \gate{X} & \qw & \qw & \qw & \gate{X} \cwx & \gate{Z} \cwx[-2] & \rstick{\ket{\psi}} \qw
+%			\gategroup{2}{2}{3}{3}{.7em}{.}
+%			\gategroup{1}{4}{2}{6}{.7em}{.}
 %		}
-	\end{subfigure}
-	\Description{On the left, \qwire code to implement Quantum Teleportation algorithm annotated by its Hoare Type that provides specification for the algorithm.
+%		% }
+%	\end{subfigure}
+%	\Description{On the left, \qwire code to implement Quantum Teleportation algorithm annotated by its Hoare Type that provides specification for the algorithm.
 On the right, the corresponding quantum circuit.}
-	\caption{Quantum Teleportation (Code and Circuit)}
-	\label{fig:teleport}
-\end{figure}
-
-In Fig.
+%	\caption{Quantum Teleportation (Code and Circuit)}
+%	\label{fig:teleport}
+%\end{figure}
+%
+%In Fig.
 \ref{fig:teleport}, on the left, we show a function \lstinline{teleport} written in \qwire-like syntax (TODO: add ref) that implements the quantum teleportation circuit shown on the right.
@@
 The result is that Bob's qubit is now in the same state as the initial state of the message qubit \lstinline{q}.
-The specification that the function must meet is written using a Hoare type, which we describe next.
-
-\subsection{Hoare Types}
-A Hoare type encodes preconditions and postconditions in the same spirit as Hoare triples.
-The type shown on the left below can be read as `for a stateful computation executed in a heap that satisfies precondition $P$, return a value of type $A$ in a heap that satisfies postcondition $Q$.' We show an example, the \lstinline{alloc} primitive from HTT~\cite{nanevski_hoare_2008}, to demonstrate the expressiveness of a Hoare type:
-\begin{mathpar}
-
-	\{P\}\; \texttt{x}:A\; \{Q\}
-
-	\and
-
-	\texttt{alloc} : \forall \alpha. \Pi x:\alpha. \{\texttt{emp}\}\; y:\texttt{nat}\; \{y \mapsto_{\alpha} x\}
-
-\end{mathpar}
-
-\begin{figure}[t]
-	\begin{tabular}{lrcl}
-		\textit{Types} & $A, B, C$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
-			$ \type{Unit} \mid \type{Bit} \mid \type{Qubit} \mid A \otimes B \mid \Pi x{ : }A.B \mid ~!A  \mid \type{Circ}(S, U) \mid \{P\} x{ : }A \{Q\}$
-					\end{minipage}\\
-		\textit{Wire Types} & $S, U$ & ::= & $ \type{Unit} \mid \type{Bit} \mid \type{Qubit} \mid S \otimes U$ \\
-		\textit{Propositions} & $P, Q, R$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
-			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid P \supset Q \mid \neg P \mid \exists x{:}A.P \mid \forall x{:}A.P \mid \mathrm{Id_A}(M, N)$
-		\end{minipage}\\
-%		\textit{(Quantum) Heaps} & $H, G$ & ::= & $ h \mid \mathrm{empty} \mid \mathrm{upd}_{\tau}(H, M, N) $\\
-%		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K\ M \mid M : \tau $\\
-		\textit{Terms} & $M, N$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
-			$ c \mid l \mid x \mid \lambda x.M \mid MN \mid \mathrm{unit} \mid (a, \mathcal{C}, b) \mid \mathrm{box_S}~M \mid \mathrm{unbox}~M \mid \mathrm{force}~M \mid \mathrm{lift}~M \mid (M, N) \mid \mathrm{let}~ x = N ~\mathrm{in}~ M $
-		\end{minipage}\\
-		\textit{Simple Terms} & $a, b$ & ::= & $ l \mid \mathrm{unit} \mid (a, b)$ \\
-		\textit{Circuits} & $\mathcal{C}, \mathcal{D}$ & : & $ \Sigma \to \Sigma' $ \\
-		\textit{Proposition Contexts} & $\Delta$ & ::= & $ \cdot \mid P, \Delta$ \\
-		\textit{Label Contexts} & $\Sigma$ & ::= & $ \cdot \mid l : \type{Qubit}, \Sigma$ \\
-%		\textit{Commands} & $c$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
-%			$ \mathrm{alloc}_\tau\ M \mid \mathrm{dealloc}\ M \mid \mathrm{if}_\tau\ M\ \mathrm{then}\ E_1\ \mathrm{else}\ E_2 \mid
-%			\mathrm{fix}\ f(y{:}\tau){:}\sigma = \mathrm{do}\ E\ \mathrm{in\ eval}\ f M \mid \mathrm{apply}(M, N) \mid \mathrm{capply}(M, N, O) $
-%		\end{minipage}\\
-%		\textit{Computations} & $E, F$ & ::= & $ \mathrm{return}\ M \mid x \gets K; E \mid x \Leftarrow c; E \mid x =_\tau M; E $\\
-%		\textit{Variable context} & $\Delta, \Psi$ & ::= & $ \cdot \mid \Delta, x{:}\tau $\\
-%		\textit{Heap context} & $X$ & ::= & $ \cdot \mid X, h $\\
-		\textit{Contexts} & $\Gamma$ & ::= & $ \cdot \mid x : A, \Gamma \mid l : \type{Qubit}, \Gamma$ \\
-		\textit{Values} & $V$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
-	$ x \mid l \mid \lambda x.M \mid \mathrm{lift}~M \mid (a, \mathcal{C}, b) \mid \mathrm{unbox}~V \mid \mathrm{force}(\mathrm{unbox}~V) \mid $
-\end{minipage}\\
-	\end{tabular}
-	\Description{Syntax of Hoare Type Theory from \cite{nanevski_hoare_2008}}
-	\caption{Syntax of Quantum (Linear) Hoare Type Theory}
-	\label{fig:syntax}
-\end{figure}
-
-Figure \ref{fig:syntax} shows syntax of Quantum HTT, inspired by the syntax of Proto-Quipper for the quantum parts.
-Symbol $c$ ranges over constants of the language such as primitive quantum gates (Hadamard, X, Y, Z, CNOT, etc), while $l$ identifies wires (bits or qubits).
-$!A$ represents duplicable/reusable data.
-The Pi type can be simplified to $A \multimap B$ if $B$ does not depend on $x$.
-$\mathrm{Id_A}(M, N)$ is propositional equality.
-$\mathrm{box_S}$ converts a circuit-producing function of type $S \multimap U$ into a circuit treated as data of type $\type{Circ}(S, U)$; $\mathrm{unbox}$ does the reverse.
-
-%In order to extend HTT to support quantum circuits, we augment its syntax with that of \qwire\cite{paykin_qwire:_2017,rand_formally_2018}.
-\qwire is a core quantum circuit language that can cooperate with an arbitrary classical host language.
-This requires minimal support from the host language to interface with the circuit language.
-In our case, HTT acts as the classical host language for \qwire.
+The specification that the function must meet is written using a Hoare type, which we described in the previous section.
 %
-%\qwire's circuit language has a wire type W that can be a unit, a bit, a qubit or a tuple of wires.
-The host language (HTT) is augmented with additional types to support interaction between the two languages and to treat circuits as data:
-%\begin{mathpar}
-%
-%	W ::= 1 \mid \texttt{bit} \mid \texttt{qbit} \mid W1 \otimes W2
-%
-%	\and
-%
-%	A ::= \ldots \mid \texttt{Unit} \mid \texttt{Bool} \mid A \times A \mid \texttt{Circuit}(W1, W2)
-%
-%\end{mathpar}
-
-The primitive quantum gates that are treated as constants in the language have types such as follows:
-\begin{mathpar}
-	\mathrm{H} : \type{Qubit} \multimap \type{Qubit}
-	\and
-	\mathrm{CNOT} : \type{Qubit} \otimes \type{Qubit} \multimap \type{Qubit}  \otimes \type{Qubit}
-	\and
-	\mathrm{Init0} : \type{Unit} \multimap \type{Qubit}
-	\and
-	\mathrm{meas} : \type{Qubit} \multimap \type{Bit}
-\end{mathpar}
-
-Simple quantum programs can be written using the above syntax, for example, cointoss:
-
-\begin{lstlisting}
-cointoss : Unit ⊸ Bit
-cointoss = let
-					q = Init0 unit
-					q = H q
-					b = meas q
-				in b
-\end{lstlisting}
-
-To write a Hoare type, we need to box the circuit as follows:
-
-\begin{lstlisting}
-boxed_cointoss : {⊤} c : Circ(Unit, Bit) {⊤}
-boxed_cointoss = box_unit cointoss
-\end{lstlisting}
-
-The propositions shown above are trivial. More interesting examples will follow.
-
-\subsection{Verification}
-Here we describe how HTT lets us verify the correctness of teleportation based on the specification provided as the type of the function.
+%\subsection{Verification}
+%Here we describe how HTT lets us verify the correctness of teleportation based on the specification provided as the type of the function.
 Type checking in HTT involves generation of strongest postconditions at each step of the program with the initial precondition as input.
@@
 The final state here is that qubits \lstinline{a} and \lstinline{b} are entangled into the first Bell state, $\ket{\phi^+} = \frac{1}{\sqrt{2}}(\ket{00} + \ket{11})$.
+%
+%% \lstset{aboveskip=1pt,belowskip=1pt}
+%\begin{lstlisting}
+%(* P0 (given): {q↦qbit\kp /\ a↦qbit|0⟩ /\ b↦qbit|0⟩} *)
+%a      ← H a;
+%(* P1: {q↦qbit\kp /\ a↦qbit1/sqrt2(|0⟩+|1⟩) /\ b↦qbit|0⟩} *)
+%(a, b) ← CX (a, b);
+%(* P2: {q↦qbit\kp /\ (a,b)↦qbit|φ+⟩ *)
+%\end{lstlisting}
 
-% \lstset{aboveskip=1pt,belowskip=1pt}
-\begin{lstlisting}
-(* P0 (given): {q↦qbit\kp /\ a↦qbit|0⟩ /\ b↦qbit|0⟩} *)
-a      ← H a;
-(* P1: {q↦qbit\kp /\ a↦qbit1/sqrt2(|0⟩+|1⟩) /\ b↦qbit|0⟩} *)
-(a, b) ← CX (a, b);
-(* P2: {q↦qbit\kp /\ (a,b)↦qbit|φ+⟩ *)
-\end{lstlisting}
+%\subsection{Soundness / Type Checking Algorithm}
 
-\subsection{Soundness / Type Checking Algorithm}
+%\section{Evaluation}
 
-\section{Evaluation}
+%\subsection{More Examples}
+%\textit{add at least one example with scaling knob}
 
-\subsection{More Examples}
-\textit{add at least one example with scaling knob}
-
-\subsection{Circuit-complexity Considerations}
+%\subsection{Circuit-complexity Considerations}
 
 \section{Related Work}
-\textit{Objective of this section\ldots}
+\label{related}
+%\textit{Objective of this section\ldots}
 
-\subsection{Quantum Languages and Types}
-\textit{A chronology of language developments wrt to usefulness of types\ldots}
+%\subsection{Quantum Languages and Types}
+%\textit{A chronology of language developments wrt to usefulness of types\ldots}
 
 \subsection{Verification of Quantum Programs}
-Previous work such as Proto-Quipper~\cite{ross_algebraic_2015,mahmoud_formalization_2019,rios_categorical_2017} and \qwire~\cite{paykin_qwire:_2017,rand_qwire_2017,rand_formally_2018} utilize a linear type system and dependent types to enforce a small subset of semantic properties, such as the no-cloning theorem and whether a unitary gate is of the right dimension.
+Previous work, such as Proto-Quipper~\cite{ross_algebraic_2015,rios_categorical_2017} and \qwire~\cite{paykin_qwire:_2017,rand_qwire_2017,rand_formally_2018} utilize a linear type system and dependent types to enforce a small subset of semantic properties, such as the no-cloning restriction and whether a unitary gate is of the right dimension.
 These advances in quantum type systems, although helpful, still fall short in encoding and enforcing even more useful properties that one would like to be able to express for the purpose of verification.
@@
 In the classical setting, Hoare Type Theory (HTT)~\cite{nanevski_hoare_2008} accomplishes exactly this goal.
-Here, we attempt to merge these ideas for the quantum realm.
+Our attempt is to merge these ideas for the quantum realm.
 
-\textit{Add from Mingsheng Ying's book, papers, etc.}
+%\textit{Add from Mingsheng Ying's book, papers, etc.}
 
 \section{Conclusion and Perspectives}
-This approach has the potential to be a unified system for programming, specification and reasoning about quantum programs.
+\label{sec:conclusion}
+In this paper, we have described our ongoing work on Quantum Hoare Type Theory, which modifies HTT for quantum computing using ideas from the Quantum IO monad work.
+This approach has the potential to be a unified system for programming, specification, and reasoning about quantum programs.
 We need to overcome challenges such as compactly specifying properties of quantum state when dealing with more than a few qubits.
@@
 
-We are working on programming other interesting quantum algorithms that can serve as a litmus test for this technique, specifically, a cryptographic protocol called Quantum Key Distribution and the well-known Grover's search algorithm.
-These examples involve some form of classical control or iteration that makes them non-trivial for our technique.
+There are several avenues of future work to explore:
 
-We also look to explore how our types will evolve when we start incorporating higher-order constructs in our quantum language such as those demonstrated in languages like quantum lambda calculus~\cite{selinger_lambda_2006} and Quipper~\cite{green_quipper:_2013}.
+\paragraph{Mechanization} There are multiple implementations of HTT in Coq such as Ynot~\cite{ynot2008}.
+We are similarly working on mechanizing QHTT in Coq for higher assurance of the usefulness and soundness of our theory.
+This will also enable us to extract verified circuits in a lower level quantum language such as OpenQASM for execution on real machines.
 
-Another venue for exploration is to incorporate more precise types that can distinguish between qubits in pure classical state vs.
+\paragraph{Quantum Assertion Logic} An obvious need is to come up with an assertion logic similar to Separation Logic~\cite{reynolds2002} for quantum computing so as to be able to reason about only the interesting portions of the quantum state while still ensuring correctness of non-local effects such as entanglement.
+Various Quantum Hoare Logics that exist~\cite{ying_floydhoare_2012} currently do not support frame rules that provide Separation Logic its power.
+
+\paragraph{Linearity} Peter Selinger and collaborators have recently proposed a linear dependent typed version of Proto-Quipper (dubbed Proto-Quipper-D)~\cite{selinger2020}.
+It is an interesting challenge to reconcile linearity in our theory based on their proposal.
+
+\paragraph{Circuits as Arrows} Further, Proto-Quipper treats quantum circuits as first class citizens of the language.
+We would like to explore modifying our theory to treat Quantum Hoare types as arrows instead of as monads as was suggested by Vizzotto et al~\cite{so-arrows}.
+It makes sense from the perspective of sequential composition as arrows can have an arbitrary number of input/outputs as opposed to monads.
+
+\paragraph{Behavioural Types} Another venue for exploration is to incorporate more precise types that can distinguish between qubits in pure classical state vs.
 those in superposition vs.
-those in entanglement inspired by the various quantum resource theories (see \cite{rand_type_2019} and references therein).
+those in entanglement~\cite{JorrandPerdrix2009} such as those inspired by the various quantum resource theories (see \cite{rand_type_2019} and references therein).
+This may help us provide more specific postconditions that quantum programmers expect to hold true of their programs.
 
-\textit{Future work about using Proto-Quipper/\qwire as languages, then extraction into QASM and checking generated verification conditions.}
+\paragraph{Classical Effects} Finally, it will be an interesting challenge to reconcile both classical and quantum effects together in a single grand unified theory for effects.
+
+%\textit{Future work about using Proto-Quipper/\qwire as languages, then extraction into QASM and checking generated verification conditions.}
 
@@
 
+\newpage
+
 \appendix
-\section{Sample Proof}
-\textit{Coq code / Annotated proof objects from implementation\ldots}
+
+\section{Modular Bell Pair}
+\label{app:mbp}
+Writing a modular version of \texttt{testBell}.
+
+\subsection{Hadamard basis states}
+\begin{Verbatim}[commandchars=\$\[\]]
+qplus : {emp} r : Qbit {Id(r, $cket[+])}
+      = do q <= mkQbit false;
+           applyU (H q);
+           return q
+
+qminus : {emp} r : Qbit {Id(r, $cket[-])}
+       = do q <= mkQbit true;
+            applyU (H q);
+            return q
+\end{Verbatim}
+
+\subsection{Creating entanglement}
+\begin{Verbatim}[commandchars=\$\[\]]
+share : $cpi a : Qbit.
+        {a $cin {$cket[+], $cket[-]}}   -- a $cin {b, c} is short for Id(a,b) \/ Id(a,c)
+           b : Qbit
+        {Id(a, b) /\ a $cin {$cket[0], $cket[1]}}
+      = \a.do b <= mkQbit false;
+              applyU (ifQ a (X b));
+              return b
+\end{Verbatim}
+
+\subsection{Bell pair}
+\begin{Verbatim}[commandchars=\$\[\]]
+bell : {emp} (a, b) : (Qbit, Qbit) {Id(a, b) /\ a $cin {$cket[0], $cket[1]}}
+     = do qa <- qplus;
+          qb <- share qa;
+          return (qa, qb)
+\end{Verbatim}
+
+\subsection{Testing modular Bell pair}
+\begin{Verbatim}[commandchars=\$\[\]]
+testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}
+         = do (qa, qb) <- bell;
+              (measQbit qa, measQbit qb)
+\end{Verbatim}
+
+\section{Teleportation}
+\label{app:teleport}
+% \todo{TODO: Refine assertions}
+\subsection{Alice's circuit}
+\begin{Verbatim}[commandchars=\$\[\]]
+alice : $cpi a : Qbit. $cpi e : Qbit
+        {(Id(a, -) /\ entangled(e)}
+           r : (Bool, Bool)
+        {emp}
+      = \a.\e.do applyU (ifQ a (X e));
+                 applyU (H a);
+                 (measQbit a, measQbit e)
+\end{Verbatim}
+
+\subsection{Bob's circuit}
+\begin{Verbatim}[commandchars=\$\[\]]
+bob : $cpi m1 m2 : Bool. $cpi e : Qbit
+      {entangled(e)}
+         r : Qbit
+      {(Id(r, -)}
+      = \m1.\m2.\e.do if m1 then applyU (Z e) else ();
+                      if m2 then applyU (X e) else ();
+                      return e
+\end{Verbatim}
+
+\subsection{Teleport}
+\begin{Verbatim}[commandchars=\$\[\]]
+teleport : $cpi q : Qbit. x : Pure.
+      {Id(q, x)}
+         r : Qbit
+      {Id(r, x)}
+      = \q.do (a, b) <- bell;
+              (m1, m2) <- alice q a;
+              tq <- bob m1 m2 b;
+              return tq
+\end{Verbatim}
+
+%\section{Sample Proof}
+%\textit{Coq code / Annotated proof objects from implementation\ldots}
 

### 2020-04-29 ms-paper/ms-paper.tex
Switch to listings package with Haskell language

@@
 \usepackage{xspace}
-\usepackage{fancyvrb}
 \usepackage{cleveref}
@@
 
-\usepackage{listings}
-\usepackage{lstcoq}
-
+\usepackage{xcolor}
 \definecolor{ltblue}{rgb}{0,0.4,0.4}
-\definecolor{dkblue}{rgb}{0,0.1,0.6}
+\definecolor{dkblue}{rgb}{0,0,1}
 \definecolor{dkgreen}{rgb}{0,0.35,0}
@@
 
-\lstset{language=coq, basicstyle=\ttfamily\scriptsize, columns=flexible,
-	keepspaces=true}
+\usepackage{listings}
+\lstloadlanguages{Haskell}
+\lstdefinelanguage{QHaskell}{%
+	language     = Haskell,
+	morekeywords = {Qbit, U, C},
+}
+\lstnewenvironment{QCode}
+{\lstset{language=QHaskell}%
+	\csname lst@SetFirstLabel\endcsname}
+{\csname lst@SaveFirstLabel\endcsname}
+\lstset{
+	basicstyle=\small\ttfamily,
+	flexiblecolumns=false,
+	basewidth={0.5em,0.45em},
+	commentstyle=\color{dkgreen},
+	keywordstyle=\bfseries\color{dkblue},
+	stringstyle=\color{dkviolet},
+	literate={+}{{$+$}}1 {/}{{$/$}}1 {*}{{$*$}}1 {=}{{$=$}}1
+	{>}{{$>$}}1 {<}{{$<$}}1 {\\}{{$\lambda$}}1
+	{\\\\}{{\char`\\\char`\\}}1
+	{->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
+	{<=}{{$\Leftarrow$}}2 {=>}{{$\Rightarrow$}}2
+	{\ .}{{$\circ$}}2 {\ .\ }{{$\circ$}}2
+	{>>}{{>>}}2 {>>=}{{>>=}}2
+	{|}{{$\mid$}}1
+	{/\\}{{$\wedge$}}1 {\\in}{{$\in$}}1
+}
 
@@
 The primitive quantum operations defined using the QIO monad below (Haskell syntax) are:
-\begin{verbatim}
+\begin{QCode}
 mkQbit : Bool -> QIO Qbit       -- initialization
@@
 measQbit : Qbit -> QIO Bool     -- measurement
-\end{verbatim}
+\end{QCode}
 
@@
 
-\begin{verbatim}
+\begin{QCode}
 rot :: Qbit -> ((Bool, Bool) -> C) -> U
 cond :: Qbit -> (Bool -> U) -> U
-\end{verbatim}
+\end{QCode}
 
@@
 
-\begin{verbatim}
+\begin{QCode}
 ifQ :: Qbit -> U -> U
 ifQ q u = cond q (\x -> if x then u else mempty)
-\end{verbatim}
+\end{QCode}
 
@@
 \subsubsection*{Hello Quantum World}
-\begin{verbatim}
+\begin{QCode}
 hqw : {emp} r : Bool {emp /\ Id(r, false)}
@@
          measQbit q
-\end{verbatim}
+\end{QCode}
 
@@
 \subsubsection*{Coin Toss}
-\begin{verbatim}
+\begin{QCode}
 rnd : {emp} r : Bool {emp}
@@
 
-\begin{Verbatim}[commandchars=\$\[\]]
+\begin{QCode}
 testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}
@@
               -- P5: P4 $ccc ((qa $cmt -) $cmm emp) $ccc ((qb $cmt -) $cmm emp)
-\end{Verbatim}
+\end{QCode}
 Finally, the \texttt{consq} rule should ensure that the assertion P5 can lead to the specified postcondition in the type of the program.
@@
 %
-%\begin{lstlisting}
+%\begin{QCode}
 %cointoss : Unit ⊸ Bit
@@
 %				in b
-%\end{lstlisting}
+%\end{QCode}
 %
@@
 %
-%\begin{lstlisting}
+%\begin{QCode}
 %boxed_cointoss : {⊤} c : Circ(Unit, Bit) {⊤}
 %boxed_cointoss = box_unit cointoss
-%\end{lstlisting}
+%\end{QCode}
 %
@@
 %		%     b       ← init0 ();
-%		\begin{lstlisting}
+%		\begin{QCode}
 %		teleport : Π\kp:qbit.{q↦qbit\kp /\ a↦qbit|0⟩ /\ b↦qbit|0⟩} c:Circuit(qbit⊗qbit⊗qbit, qbit) {b↦qbit\kp} =
@@
 %		output b
-%		\end{lstlisting}
+%		\end{QCode}
 %		% ()      ← discard y;
@@
 %% \lstset{aboveskip=1pt,belowskip=1pt}
-%\begin{lstlisting}
+%\begin{QCode}
 %(* P0 (given): {q↦qbit\kp /\ a↦qbit|0⟩ /\ b↦qbit|0⟩} *)
@@
 %(* P2: {q↦qbit\kp /\ (a,b)↦qbit|φ+⟩ *)
-%\end{lstlisting}
+%\end{QCode}
 
@@
 \subsection{Hadamard basis states}
-\begin{Verbatim}[commandchars=\$\[\]]
+\begin{QCode}
 qplus : {emp} r : Qbit {Id(r, $cket[+])}
@@
             return q
-\end{Verbatim}
+\end{QCode}
 
 \subsection{Creating entanglement}
-\begin{Verbatim}[commandchars=\$\[\]]
+\begin{QCode}
 share : $cpi a : Qbit.
@@
               return b
-\end{Verbatim}
+\end{QCode}
 
 \subsection{Bell pair}
-\begin{Verbatim}[commandchars=\$\[\]]
+\begin{QCode}
 bell : {emp} (a, b) : (Qbit, Qbit) {Id(a, b) /\ a $cin {$cket[0], $cket[1]}}
@@
           return (qa, qb)
-\end{Verbatim}
+\end{QCode}
 
 \subsection{Testing modular Bell pair}
-\begin{Verbatim}[commandchars=\$\[\]]
+\begin{QCode}
 testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}
@@
               (measQbit qa, measQbit qb)
-\end{Verbatim}
+\end{QCode}
 
@@
 \subsection{Alice's circuit}
-\begin{Verbatim}[commandchars=\$\[\]]
+\begin{QCode}
 alice : $cpi a : Qbit. $cpi e : Qbit
@@
                  (measQbit a, measQbit e)
-\end{Verbatim}
+\end{QCode}
 
 \subsection{Bob's circuit}
-\begin{Verbatim}[commandchars=\$\[\]]
+\begin{QCode}
 bob : $cpi m1 m2 : Bool. $cpi e : Qbit
@@
                       return e
-\end{Verbatim}
+\end{QCode}
 
 \subsection{Teleport}
-\begin{Verbatim}[commandchars=\$\[\]]
+\begin{QCode}
 teleport : $cpi q : Qbit. x : Pure.
@@
               return tq
-\end{Verbatim}
+\end{QCode}
 

### 2020-04-29 ms-paper/ms-paper.tex
Convert remaining symbols to listings literates

@@
 	language     = Haskell,
-	morekeywords = {Qbit, U, C},
+	morekeywords = {Qbit, U, C, Pure},
 }
@@
 	{>>}{{>>}}2 {>>=}{{>>=}}2
-	{|}{{$\mid$}}1
-	{/\\}{{$\wedge$}}1 {\\in}{{$\in$}}1
+	{|}{{$\vert$}}1 {\\>}{{$\rangle$}}1
+	{/\\}{{$\wedge$}}1 {\\/}{{$\vee$}}1
+	{\\in}{{$\in$}}1 {\\o}{{$\circ$}}1 {\\Pi}{{$\Pi$}}1
+	{|->}{{$\mapsto$}}2 {-o}{{$\multimap$}}2
 }
@@
 \newcommand{\todo}[1]{\textcolor{red}{#1}}
-
-\newcommand{\cket}[1]{{$\ket{#1}$}}
-\newcommand\cin{{$\in$}}
-\newcommand\cmt{{$\mapsto$}}
-\newcommand\cmm{{$\multimap$}}
-\newcommand\ccc{{$\circ$}}
-\newcommand{\cpi}{$\Pi$}
 
@@
 \subsubsection*{Testing Bell Pair}
-\begin{Verbatim}[commandchars=\$\[\]]
-testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}  -- a,b $cin {true, false}
+\begin{QCode}
+testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}  -- a,b \in {true, false}
          = do qa <= mkQbit false;
@@
          = do qa <= mkQbit false;
-              -- P1: P0 $ccc (qa $cmt state(false))
+              -- P1: P0 \o (qa |-> state(false))
               applyU (H qa);
-              -- P2: P1 $ccc (qa $cmt state({(0.5, false), (0.5, true)}))
+              -- P2: P1 \o (qa |-> state({(0.5, false), (0.5, true)}))
               qb <= mkQbit false;
-              -- P3: P2 $ccc (qa $cmt state(false))
+              -- P3: P2 \o (qa |-> state(false))
               applyU (ifQ qa (X qb));
-              -- P4: P3 $ccc ((qa, qb) $cmt state({(0.5, (false, false),
+              -- P4: P3 \o ((qa, qb) |-> state({(0.5, (false, false),
               --                              (0.5, (true, true))}))
                 (measQbit qa, measQbit qb)
-              -- P5: P4 $ccc ((qa $cmt -) $cmm emp) $ccc ((qb $cmt -) $cmm emp)
+              -- P5: P4 \o ((qa |-> -) -o emp) \o ((qb |-> -) -o emp)
 \end{QCode}
@@
 \begin{QCode}
-qplus : {emp} r : Qbit {Id(r, $cket[+])}
+qplus : {emp} r : Qbit {Id(r, |+\>)}
       = do q <= mkQbit false;
@@
 
-qminus : {emp} r : Qbit {Id(r, $cket[-])}
+qminus : {emp} r : Qbit {Id(r, |-\>)}
        = do q <= mkQbit true;
@@
 \begin{QCode}
-share : $cpi a : Qbit.
-        {a $cin {$cket[+], $cket[-]}}   -- a $cin {b, c} is short for Id(a,b) \/ Id(a,c)
+share : \Pi a : Qbit.
+        {a \in {|+\>, |-\>}}   -- a \in {b, c} is short for Id(a,b) \/ Id(a,c)
            b : Qbit
-        {Id(a, b) /\ a $cin {$cket[0], $cket[1]}}
+        {Id(a, b) /\ a \in {|0\>, |1\>}}
       = \a.do b <= mkQbit false;
@@
 \begin{QCode}
-bell : {emp} (a, b) : (Qbit, Qbit) {Id(a, b) /\ a $cin {$cket[0], $cket[1]}}
+bell : {emp} (a, b) : (Qbit, Qbit) {Id(a, b) /\ a \in {|0\>, |1\>}}
      = do qa <- qplus;
@@
 \begin{QCode}
-alice : $cpi a : Qbit. $cpi e : Qbit
+alice : \Pi a : Qbit. \Pi e : Qbit
         {(Id(a, -) /\ entangled(e)}
@@
 \begin{QCode}
-bob : $cpi m1 m2 : Bool. $cpi e : Qbit
+bob : \Pi m1 m2 : Bool. \Pi e : Qbit
       {entangled(e)}
@@
 \begin{QCode}
-teleport : $cpi q : Qbit. x : Pure.
+teleport : \Pi q : Qbit. x : Pure.
       {Id(q, x)}

### 2020-04-29 ms-paper/ms-paper.tex
Switch to ACM recommended color from xcolor
Get rid of warnings and commented stuff

@@
 
-\usepackage{xcolor}
+\usepackage{color}
 \definecolor{ltblue}{rgb}{0,0.4,0.4}
@@
 \lstloadlanguages{Haskell}
-\lstdefinelanguage{QHaskell}{%
+\lstdefinelanguage{QHaskell}{
 	language     = Haskell,
@@
 }
-\lstnewenvironment{QCode}
-{\lstset{language=QHaskell}%
-	\csname lst@SetFirstLabel\endcsname}
-{\csname lst@SaveFirstLabel\endcsname}
+
 \lstset{
@@
 The primitive quantum operations defined using the QIO monad below (Haskell syntax) are:
-\begin{QCode}
+\begin{lstlisting}[language=QHaskell]
 mkQbit : Bool -> QIO Qbit       -- initialization
@@
 measQbit : Qbit -> QIO Bool     -- measurement
-\end{QCode}
+\end{lstlisting}
 
@@
 
-\begin{QCode}
+\begin{lstlisting}[language=QHaskell]
 rot :: Qbit -> ((Bool, Bool) -> C) -> U
 cond :: Qbit -> (Bool -> U) -> U
-\end{QCode}
+\end{lstlisting}
 
@@
 
-\begin{QCode}
+\begin{lstlisting}[language=QHaskell]
 ifQ :: Qbit -> U -> U
 ifQ q u = cond q (\x -> if x then u else mempty)
-\end{QCode}
+\end{lstlisting}
 
@@
 \subsubsection*{Hello Quantum World}
-\begin{QCode}
+\begin{lstlisting}[language=QHaskell]
 hqw : {emp} r : Bool {emp /\ Id(r, false)}
@@
          measQbit q
-\end{QCode}
+\end{lstlisting}
 
@@
 \subsubsection*{Coin Toss}
-\begin{QCode}
+\begin{lstlisting}[language=QHaskell]
 rnd : {emp} r : Bool {emp}
@@
 \subsubsection*{Testing Bell Pair}
-\begin{QCode}
+\begin{lstlisting}[language=QHaskell]
 testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}  -- a,b \in {true, false}
@@
               (measQbit qa, measQbit qb)
-\end{QCode}
+\end{lstlisting}
 
@@
 
-\begin{QCode}
+\begin{lstlisting}[language=QHaskell]
 testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}
@@
               -- P5: P4 \o ((qa |-> -) -o emp) \o ((qb |-> -) -o emp)
-\end{QCode}
+\end{lstlisting}
 Finally, the \texttt{consq} rule should ensure that the assertion P5 can lead to the specified postcondition in the type of the program.
@@
 Both of these involve specifying assertions over quantum state whose theory we still need to work out.
-
-%Figure \ref{fig:syntax} shows syntax of Quantum HTT, inspired by the syntax of Proto-Quipper for the quantum parts.
-Symbol $c$ ranges over constants of the language such as primitive quantum gates (Hadamard, X, Y, Z, CNOT, etc), while $l$ identifies wires (bits or qubits).
-$!A$ represents duplicable/reusable data.
-The Pi type can be simplified to $A \multimap B$ if $B$ does not depend on $x$.
-$\mathrm{Id_A}(M, N)$ is propositional equality.
-$\mathrm{box_S}$ converts a circuit-producing function of type $S \multimap U$ into a circuit treated as data of type $\type{Circ}(S, U)$; $\mathrm{unbox}$ does the reverse.
-lift and force convert between linear and reusable types.
-
-%The primitive quantum gates that are treated as constants in the language have types such as follows:
-%\begin{mathpar}
-%	\mathrm{H} : \type{Qubit} \multimap \type{Qubit}
-%	\and
-%	\mathrm{CNOT} : \type{Qubit} \otimes \type{Qubit} \multimap \type{Qubit}  \otimes \type{Qubit}
-%	\\
-%	\mathrm{Init0} : \type{Unit} \multimap \type{Qubit}
-%	\and
-%	\mathrm{meas} : \type{Qubit} \multimap \type{Bit}
-%\end{mathpar}
-%
-%Simple quantum programs can be written using the above syntax, for example, cointoss:
-%
-%\begin{QCode}
-%cointoss : Unit ⊸ Bit
-%cointoss = let
-%					q = Init0 unit
-%					q = H q
-%					b = meas q
-%				in b
-%\end{QCode}
-%
-%To write a Hoare type, we need to box the circuit as follows:
-%
-%\begin{QCode}
-%boxed_cointoss : {⊤} c : Circ(Unit, Bit) {⊤}
-%boxed_cointoss = box_unit cointoss
-%\end{QCode}
-%
-%The propositions shown above are trivial. More interesting examples will follow.
-%
-%\subsection{Example: Quantum Teleportation}\label{telep}
-%Teleportation is a simple quantum algorithm that is used to communicate a quantum state using two classical bits and a shared entangled pair of qubits.
-Alice is trying to send a message encoded in the qubit to Bob.
-There are three broad steps in the protocol: 1) preparation, ie, entanglement creation between two qubits to be shared between Alice and Bob in advance; 2) sending the message by Alice; and, 3) reconstruction of the message by Bob.
-%\begin{figure}[h]
-%	\begin{subfigure}{3in}
-%		% 	a       ← init0 ();
-%		%     b       ← init0 ();
-%		\begin{QCode}
-%		teleport : Π\kp:qbit.{q↦qbit\kp /\ a↦qbit|0⟩ /\ b↦qbit|0⟩} c:Circuit(qbit⊗qbit⊗qbit, qbit) {b↦qbit\kp} =
-%		box (q, a, b) ⇒
-%		(* 1. Bell state preparation for Alice and Bob *)
-%		a       ← H a;
-%		(a, b)  ← CX (a, b);
-%		(* 2. Bell measurement by Alice *)
-%		(q, a)  ← CX (q, a);
-%		q       ← H q;
-%		x       ← meas q;
-%		y       ← meas a;
-%		(* 3. Correction by Bob *)
-%		(y, b)  ← CX (y, b);
-%		(x, b)  ← CZ (x, b);
-%		output b
-%		\end{QCode}
-%		% ()      ← discard y;
-%		%     ()      ← discard x;
-%	\end{subfigure}
-%	\begin{subfigure}{1in}
-%		% 	\centerline{
-%		\Qcircuit @C=.4em @R=.4em @! {
-%			\lstick{q=\ket{\psi}} & \qw & \qw & \ctrl{1} & \gate{H} & \meter & \cw & \control \cw\\
-%			\lstick{a=\ket{0}} & \gate{H} & \ctrl{1} & \gate{X} & \qw & \meter & \control \cw & \\
-%			\lstick{b=\ket{0}} & \qw & \gate{X} & \qw & \qw & \qw & \gate{X} \cwx & \gate{Z} \cwx[-2] & \rstick{\ket{\psi}} \qw
-%			\gategroup{2}{2}{3}{3}{.7em}{.}
-%			\gategroup{1}{4}{2}{6}{.7em}{.}
-%		}
-%		% }
-%	\end{subfigure}
-%	\Description{On the left, \qwire code to implement Quantum Teleportation algorithm annotated by its Hoare Type that provides specification for the algorithm.
-On the right, the corresponding quantum circuit.}
-%	\caption{Quantum Teleportation (Code and Circuit)}
-%	\label{fig:teleport}
-%\end{figure}
-%
-%In Fig.
-\ref{fig:teleport}, on the left, we show a function \lstinline{teleport} written in \qwire-like syntax (TODO: add ref) that implements the quantum teleportation circuit shown on the right.
-It returns one qubit and takes three qubits as arguments: one qubit \lstinline{q} that could be in any unknown state $\ket{\psi}$ which Alice is trying to send to Bob; and two qubits, \lstinline{a} and \lstinline{b}, initially in the state $\ket{0}$.
-The function first entangles the latter two qubits into a Bell state.
-Then, Alice performs a Bell measurement of her two qubits \lstinline{q} and \lstinline{a} and sends the resulting classical bits (\lstinline{x} and \lstinline{y}) to Bob.
-Finally, Bob applies corrective measurement using the quantum gates X and Z controlled by the classical bits that he received from Alice.
-The result is that Bob's qubit is now in the same state as the initial state of the message qubit \lstinline{q}.
-The specification that the function must meet is written using a Hoare type, which we described in the previous section.
-%
-%\subsection{Verification}
-%Here we describe how HTT lets us verify the correctness of teleportation based on the specification provided as the type of the function.
-Type checking in HTT involves generation of strongest postconditions at each step of the program with the initial precondition as input.
-Verification is a two-step process: the first phase does basic type checking and verification condition generation which is decidable, the second phase needs to show the validity of the generated verification conditions, which is undecidable.
-This second phase can be deferred to an automated theorem prover.
-For illustration, we show the assertions as we step through the first step of teleportation.
-The final state here is that qubits \lstinline{a} and \lstinline{b} are entangled into the first Bell state, $\ket{\phi^+} = \frac{1}{\sqrt{2}}(\ket{00} + \ket{11})$.
-%
-%% \lstset{aboveskip=1pt,belowskip=1pt}
-%\begin{QCode}
-%(* P0 (given): {q↦qbit\kp /\ a↦qbit|0⟩ /\ b↦qbit|0⟩} *)
-%a      ← H a;
-%(* P1: {q↦qbit\kp /\ a↦qbit1/sqrt2(|0⟩+|1⟩) /\ b↦qbit|0⟩} *)
-%(a, b) ← CX (a, b);
-%(* P2: {q↦qbit\kp /\ (a,b)↦qbit|φ+⟩ *)
-%\end{QCode}
 
@@
 \subsection{Hadamard basis states}
-\begin{QCode}
+\begin{lstlisting}[language=QHaskell]
 qplus : {emp} r : Qbit {Id(r, |+\>)}
@@
             return q
-\end{QCode}
+\end{lstlisting}
 
 \subsection{Creating entanglement}
-\begin{QCode}
+\begin{lstlisting}[language=QHaskell]
 share : \Pi a : Qbit.
@@
               return b
-\end{QCode}
+\end{lstlisting}
 
 \subsection{Bell pair}
-\begin{QCode}
+\begin{lstlisting}[language=QHaskell]
 bell : {emp} (a, b) : (Qbit, Qbit) {Id(a, b) /\ a \in {|0\>, |1\>}}
@@
           return (qa, qb)
-\end{QCode}
+\end{lstlisting}
 
 \subsection{Testing modular Bell pair}
-\begin{QCode}
+\begin{lstlisting}[language=QHaskell]
 testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}
@@
 \subsection{Alice's circuit}
-\begin{QCode}
+\begin{lstlisting}[language=QHaskell]
 alice : \Pi a : Qbit. \Pi e : Qbit
@@
                  (measQbit a, measQbit e)
-\end{QCode}
+\end{lstlisting}
 
 \subsection{Bob's circuit}
-\begin{QCode}
+\begin{lstlisting}[language=QHaskell]
 bob : \Pi m1 m2 : Bool. \Pi e : Qbit
@@
                       return e
-\end{QCode}
+\end{lstlisting}
 
 \subsection{Teleport}
-\begin{QCode}
+\begin{lstlisting}[language=QHaskell]
 teleport : \Pi q : Qbit. x : Pure.
@@
               return tq
-\end{QCode}
+\end{lstlisting}
 

### 2020-04-29 ms-paper/ms-paper.tex
Remove unnecessary packages and color definitions

@@
 \citestyle{acmauthoryear}
-
-\usepackage{booktabs}
-\usepackage{subcaption}
 
@@
 
-\usepackage{color}
-\definecolor{ltblue}{rgb}{0,0.4,0.4}
-\definecolor{dkblue}{rgb}{0,0,1}
 \definecolor{dkgreen}{rgb}{0,0.35,0}
-\definecolor{dkviolet}{rgb}{0.3,0,0.5}
-\definecolor{dkred}{rgb}{0.5,0,0}
 
@@
 	commentstyle=\color{dkgreen},
-	keywordstyle=\bfseries\color{dkblue},
-	stringstyle=\color{dkviolet},
+	keywordstyle=\bfseries\color{blue},
+	stringstyle=\color{violet},
 	literate={+}{{$+$}}1 {/}{{$/$}}1 {*}{{$*$}}1 {=}{{$=$}}1

### 2020-04-29 ms-paper/ms-paper.tex
Add bibentry for Fu, Kishida, Selinger paper on Linear Dependent Type Theory

@@
 
-\paragraph{Linearity} Peter Selinger and collaborators have recently proposed a linear dependent typed version of Proto-Quipper (dubbed Proto-Quipper-D)~\cite{selinger2020}.
+\paragraph{Linearity} Peter Selinger and collaborators have recently proposed a linear dependent typed version of Proto-Quipper (dubbed Proto-Quipper-D)~\cite{selinger2020,fu2020linear}.
 It is an interesting challenge to reconcile linearity in our theory based on their proposal.

### 2020-04-30 ms-paper/ms-paper.tex
The pair types need to be revised; subsubsection was missing mysteriously

@@
 \subsubsection*{Hello Quantum World}
+We start out in an empty state and assert that the result is equal to \texttt{false}.
 \begin{lstlisting}[language=QHaskell]
@@
 
-We start out in an empty state and assert that the result is equal to \texttt{false}.
+\subsubsection*{Coin Toss}
+As the result can be in either of its two values \texttt{false} or \texttt{true}, we need not specify any special postcondition as type checking the result will be sufficient for correctness.
 
-\subsubsection*{Coin Toss}
 \begin{lstlisting}[language=QHaskell]
@@
 
-As the result can be in either of its two values \texttt{false} or \texttt{true}, we need not specify any special postcondition as type checking the result will be sufficient for correctness.
+\subsubsection*{Testing Bell Pair}
+Here, we are asserting that the returned booleans $a$ and $b$ hold the same value, which is what we expect from a Bell pair.
 
-\subsubsection*{Testing Bell Pair}
+\todo{The type should be \type{Bool} $\otimes$ \type{Bool}}
 \begin{lstlisting}[language=QHaskell]
@@
 \end{lstlisting}
-
-Here, we are asserting that the returned booleans $a$ and $b$ hold the same value, which is what we expect from a Bell pair.
 
@@
 \subsection{Bell pair}
+\todo{The type should be \type{Qbit} $\otimes$ \type{Qbit}}
 \begin{lstlisting}[language=QHaskell]

### 2020-05-05 ms-paper/ms-paper.tex
Another missing todo
Update ms-paper.tex

@@
 \subsubsection{Judgments}
-The type system is designed to be bidirectional and syntax directed and hences include separate judgements for intro and elim terms. We are able to use them unchanged for QHTT.
+The type system is designed to be bidirectional and syntax directed and hence includes separate judgements for intro and elim terms. We are able to use them unchanged for QHTT.
 
@@
 \label{app:teleport}
-% \todo{TODO: Refine assertions}
+\todo{TODO: Refine assertions}
 \subsection{Alice's circuit}

### 2020-05-08 ms-paper/ms-paper.tex
Minor updates
Fix minor type, add some todos

@@
 
-\renewcommand*\contentsname{Table of Contents}
 \tableofcontents
@@
 
-\paragraph{Computations} The computations are the usual monadic return and (three) bind operations as in HTT.
+\paragraph{Computations} The computations are the usual monadic return and (three) bind operations as in HTT. \todo{TODO: $\Leftarrow$ is currently overloaded with notation for type checking}
 
@@
 
-We can now show some typing rules:
+We can now state the typing rules:
 
@@
 
-\todo{The type should be \type{Bool} $\otimes$ \type{Bool}}
+\todo{TODO: The type should be \type{Bool} $\otimes$ \type{Bool}}
 \begin{lstlisting}[language=QHaskell]
@@
               qb <= mkQbit false;
-              -- P3: P2 \o (qa |-> state(false))
+              -- P3: P2 \o (qb |-> state(false))
               applyU (ifQ qa (X qb));
@@
 \subsection{Bell pair}
-\todo{The type should be \type{Qbit} $\otimes$ \type{Qbit}}
+\todo{TODO: The type should be \type{Qbit} $\otimes$ \type{Qbit}}
 \begin{lstlisting}[language=QHaskell]

### 2020-05-25 ms-paper/ms-paper.tex
Add minor feedback from QPL reviewers

@@
-\documentclass[acmsmall,nonacm,timestamp]{acmart}
+\documentclass[acmsmall,nonacm,timestamp,review=false,anonymous=false]{acmart}
 
@@
 Here, \texttt{rot} takes a qubit and a two-by-two complex-valued unitary matrix (represented as a function from the matrix indices to $\mathbb{C}$) and lets one define arbitrary rotation on a single qubit.
-\texttt{rot} can then be used to define standard gates such as the hadamard, H, and the Pauli gates, X, Y and Z.
+\texttt{rot} can then be used to define standard gates such as the Hadamard, H, and the Pauli gates, X, Y and Z.
 \texttt{cond} is the unitary conditional that is used to perform branching unitary operations.
@@
 \subsubsection{Judgments}
-The type system is designed to be bidirectional and syntax directed and hence includes separate judgements for intro and elim terms. We are able to use them unchanged for QHTT.
+The type system is designed to be bidirectional and syntax directed and hence it includes separate judgements for intro and elim terms. We are able to use them unchanged for QHTT.
 

### 2020-06-03 ms-paper/ms-paper.tex
Changes to everything apart from the main paper

@@
 \newcommand{\todo}[1]{\textcolor{red}{#1}}
+
+% modified from bidir-survey
+\definecolor{dRed}{rgb}{0.85, 0.0, 0.0}
+\definecolor{dBlue}{rgb}{0.0, 0.0, 0.85}
+
+\newcommand{\chkcolor}{dBlue}
+\newcommand{\syncolor}{dRed}
+\newcommand{\chk}{\,\textcolor{\chkcolor}{\Leftarrow}\,}
+\newcommand{\uncoloredsyn}{{\Rightarrow}}
+\newcommand{\syn}{\,\textcolor{\syncolor}{\uncoloredsyn}\,}
 
@@
 Then we introduce our contributions in merging these ideas together for specification and enforcement of useful semantic properties in section \cref{sec:qhtt} along with its syntax (\cref{sec:syntax}), typing (\cref{sec:typing}) and, some examples (\cref{sec:examples}).
-Some related work follows in \cref{related}.
+Some related work follows in \cref{sec:related}.
 Finally, we conclude and share ideas for future work in the final section (\cref{sec:conclusion}).
@@
 This step also involves generating verification conditions that cannot be checked during this first phase of typechecking.
+We are also looking at alternatives such as the recently proposed sum-over-paths action semantics \cite{amy18}.
 
@@
 \section{Related Work}
-\label{related}
+\label{sec:related}
 %\textit{Objective of this section\ldots}
@@
 \begin{acks}
-%	I would like to thank my advisor John Reppy for suggesting me to look at Hoare Type Theory that led to this master's thesis and for his constant support and encouragement to see it through.
-I also thank my committee members Fred Chong and Ravi Chugh for reading drafts of this document and providing useful feedback.
+	I thank John Reppy, Robert Rand and the anonymous reviewers of QPL 2020 for their feedback on an earlier presentation of this work.
 

### 2020-06-03 ms-paper/ms-paper.tex
Transfer content from QPL preprint, some minor refinements

@@
 As programming languages researchers, we think it will be better to encode such assertions into a static type system both for formal verification and to aid the programmers in writing correct programs from the start.
-Inspired by the use of Hoare triples in the verification of imperative programs and building on the idea of Hoare Type Theory (HTT) for classical programming languages~\cite{nanevski_hoare_2008}, we propose Quantum Hoare Type Theory (by extending the Quantum IO Monad interface\cite{qio}) aimed at enabling both sound static type checking and formal verification of quantum programs.
+Inspired by the use of Hoare triples in the verification of imperative programs and building on the idea of Hoare Type Theory (HTT) for classical programming languages~\cite{nanevski_hoare_2008}, we extend the Quantum IO Monad interface~\cite{qio} and propose Quantum Hoare Type Theory aimed at enabling both sound static type checking and formal verification of quantum programs.
 
 The main idea is that instead of just indexing the QIO monad type with the type of computation result, we can also index it with pre-conditions and post-conditions so as to integrate Hoare-style reasoning into the type system itself.
-The resulting Quantum Hoare Type, $\{P\} x{ : }A \{Q\}$, specifies the preconditions, $P$, that hold on the quantum state before execution; the result, $x$, and its type, $A$, on successful execution; and, the postconditions, $Q$, that are true for the quantum state after successful execution.
+The resulting Quantum Hoare Type, $\{P\}\ x{ : }A\ \{Q\}$, specifies the preconditions, $P$, that hold on the quantum state before execution; the result, $x$, and its type, $A$; and, the postconditions, $Q$, that are true for the quantum state after successful execution.
 In this way the effectful quantum fragment of the program is effectively encapsulated inside the Quantum Hoare monad.
@@
 This paper is organized as follows.
-We discuss background in the next section (\cref{sec:bg}); specifically, quantum computing (\cref{sec:qc}),  Hoare Type Theory in the classical setting (\cref{sec:htt}) and then the Quantum IO Monad interface (\cref{sec:qio}).
-Then we introduce our contributions in merging these ideas together for specification and enforcement of useful semantic properties in section \cref{sec:qhtt} along with its syntax (\cref{sec:syntax}), typing (\cref{sec:typing}) and, some examples (\cref{sec:examples}).
+We discuss background in the next section, \cref{sec:bg}; specifically, quantum computing (\cref{sec:qc}),  Hoare Type Theory in the classical setting (\cref{sec:htt}) and then the Quantum IO Monad interface (\cref{sec:qio}).
+Then we introduce our contributions in merging these ideas together for specification and enforcement of useful semantic properties in section~\cref{sec:qhtt} along with its syntax (\cref{sec:syntax}), typing rules (\cref{sec:typing}) and, some examples (\cref{sec:examples}).
 Some related work follows in \cref{sec:related}.
@@
 \label{sec:bg}
-In this section, we provide background on the core ideas from existing literature that form the foundation for our work -- Hoare Type Theory (HTT) and the Quantum IO Monad interface.
+In this section, we provide background on the core ideas from existing literature that form the foundation for our work --- Hoare Type Theory (HTT) and the Quantum IO Monad interface.
 We start with some basics of quantum computing.
@@
 %We will see how this property can be utilized for computation in the quantum teleportation algorithm that we describe and program in \S\ref{telep}.
-We refer the reader to the standard textbook~\cite{nielsen_quantum_2010} or Nielsen's excellent series of essays~\cite{matuschak2019} for more background.
+We refer the reader to the standard textbook by \citet{nielsen_quantum_2010} or the excellent series of essays by \citet{matuschak2019} for more background.
 
@@
 \label{sec:htt}
-A Hoare type encodes preconditions and postconditions in the same spirit as Hoare triples to allow both specification and verification of effectful classical programs.
-The type shown on the left below can be read as `for a stateful computation executed in a heap that satisfies precondition $P$, return a value of type $A$ in a heap that satisfies postcondition $Q$.' We show an example, the \texttt{alloc} primitive from HTT~\cite{nanevski_hoare_2008}, to demonstrate the expressiveness of a Hoare type:
+A Hoare type, $\Delta.\Psi.\{P\}\; \texttt{x}:A\; \{Q\}$, encodes preconditions and postconditions in the same spirit as Hoare triples to allow both specification and verification of effectful classical programs.
+It can be read as `for a stateful computation executed in a heap that satisfies precondition $P$, return a value of type $A$ in a heap that satisfies postcondition $Q$.' The contexts, $\Delta$ and $\Psi$, contain the variables and heap variables that may appear in both $P$ and $Q$.
+In this presentation, we will omit the contexts in the Hoare type when they are unneeded.
+We show an example to demonstrate the expressiveness of a Hoare type --- the \texttt{alloc} primitive from HTT~\cite{nanevski_hoare_2008}:
 \begin{mathpar}
-	\{P\}\; \texttt{x}:A\; \{Q\}
-
-	\and
-
 	\texttt{alloc} : \forall \alpha. \Pi x:\alpha. \{\texttt{emp}\}\; y:\texttt{nat}\; \{y \mapsto_{\alpha} x\}
@@
 
+This type specifies that \texttt{alloc} is a polymorphic function that takes as input a variable, $x$, of any simple type, $\alpha$, that is executed in an empty heap (meaning it does not affect existing heap), returns a new location bound by a fresh variable, $y$, of type \texttt{nat} and initializes it with the supplied value $x$ of type $\alpha$.
+
 Type checking in HTT involves generation of strongest postconditions at each step of the program.
-Verification is a two-step process: the first phase does basic type checking and verification condition generation which is decidable, the second phase needs to show the validity of the generated verification conditions, which is undecidable.
+Verification is a two-step process: the first phase does basic type checking and verification condition generation which is decidable, the second phase needs to show the validity of the generated verification conditions, which can be undecidable.
 This second phase can be deferred to an automated theorem prover.
 
-We provide more details when we introduce QHTT in section \cref{sec:qhtt}.
+We provide more details when we introduce QHTT in section~\cref{sec:qhtt}. An accessible introduction to Hoare Type Theory is available in the lecture notes by \citet{perconti2012}.
 
@@
 
+QIO interface was developed as a library for the Haskell programming language but its design was influenced~\cite{asg2010} by the category \textbf{FQC} of finite quantum computations (where computations are interpreted as superoperators) as explored by \citet{fqc06} in previous work.
+This is the most we will say about the relation of QIO monad to category theory in this paper.
+
 Here we discuss the relevant bits from the QIO interface.
@@
 The primitive quantum operations defined using the QIO monad below (Haskell syntax) are:
+
+\begin{minipage}{0.95\linewidth}
 \begin{lstlisting}[language=QHaskell]
-mkQbit : Bool -> QIO Qbit       -- initialization
-applyU : U -> QIO ()            -- apply a unitary to quantum state
-measQbit : Qbit -> QIO Bool     -- measurement
+mkQbit :: Bool -> QIO Qbit       -- initialization
+applyU :: U -> QIO ()            -- apply a unitary to quantum state
+measQbit :: Qbit -> QIO Bool     -- measurement
 \end{lstlisting}
+\end{minipage}
 
 Further, \texttt{U} is defined as a monoid with neutral element, \texttt{mempty}, and operation, \texttt{mappend}, that encodes reversible operations on quantum state.
-Here are the core unitary primitives that we will need:
+The core unitary primitives that we will need are:
 
+\begin{minipage}{0.95\linewidth}
 \begin{lstlisting}[language=QHaskell]
@@
 \end{lstlisting}
+\end{minipage}
 
-Here, \texttt{rot} takes a qubit and a two-by-two complex-valued unitary matrix (represented as a function from the matrix indices to $\mathbb{C}$) and lets one define arbitrary rotation on a single qubit.
+\texttt{rot} takes a qubit and a two-by-two complex-valued unitary matrix (represented as a function from the matrix indices to $\mathbb{C}$) and lets one define arbitrary rotation on a single qubit.
 \texttt{rot} can then be used to define standard gates such as the Hadamard, H, and the Pauli gates, X, Y and Z.
@@
 
+\begin{minipage}{0.95\linewidth}
 \begin{lstlisting}[language=QHaskell]
@@
 \end{lstlisting}
+\end{minipage}
 
@@
 
-Quantum state in the QIO interface is modeled as a normalized vector that intuitively stores pairs representing complex amplitudes associated with each basis state (represented as a map from Qbit to Bool types).
+Quantum state in the QIO interface is modeled as a normalized vector that stores pairs representing complex amplitudes associated with each basis state (represented as a map from Qbit to Bool types).
 The operations defined over the vector class ensure that the quantum state is kept normalized throughout.
@@
 
-Figure \ref{fig:syntax} shows the syntax of Quantum Hoare Type Theory. Our presentation closely follows that of HTT~\cite{nanevski_hoare_2008}.
+\Cref{fig:syntax} shows the syntax of Quantum Hoare Type Theory. Our presentation closely follows that of HTT~\cite{nanevski_hoare_2008}.
 
@@
 As done in HTT, some common convenience assertions can be defined using the base primitives.
-Particularly, $\kw{emp}$ denotes that the heap is empty.
-M $\mapsto$ N represents that this is the only mapping in the heap.
-M $\hookrightarrow$ N represents that looking up M in the heap returns N.
-$\kw{seleq}$(H, M, N) asserts that in heap H, location M points to N.
+Particularly, $\kw{emp}$ denotes that the current heap, say $h$, is empty, that is, $\kw{HId}(h, \kw{empty})$.
+$M \mapsto N$ represents the only mapping in the (singleton) heap.
+$M \hookrightarrow N$ says that looking up $M$ in the heap returns $N$ which can be more explicitly written as $\kw{seleq}(H, M, N)$.
 
@@
 
-For the purpose of the type theory, heaps are functional, so that, for example, $\kw{upd}(H, M, N)$ returns a new heap after updating heap H at location M with N.
+For the purpose of the type theory, heaps are functional, so that, for example, $\kw{upd}(H, M, N)$ returns a new heap after updating heap $H$ at location $M$ with $N$.
 
@@
 
-\paragraph{Computations} The computations are the usual monadic return and (three) bind operations as in HTT. \todo{TODO: $\Leftarrow$ is currently overloaded with notation for type checking}
+\paragraph{Computations} The computations are the usual monadic return and three sequencing operations that bind their result to a variable similar to HTT.
+The first, $x \leftarrow K$, executes a suspended computation (such as applying a function that includes a \texttt{do} $E$ in its body); the second, $x \Leftarrow c$ executes a primitive command $c$; and the last, $x =_A M$, is just syntactic sugar for let-binding.
 
@@
 \subsubsection{Judgments}
-The type system is designed to be bidirectional and syntax directed and hence it includes separate judgements for intro and elim terms. We are able to use them unchanged for QHTT.
+We explain the main judgments used in HTT that are required to understand the typing rules in the next section.
+The type system is designed to be bidirectional and syntax directed and hence includes separate judgements for intro and elim terms.
+We are able to use them unchanged for QHTT.
+A recent survey by \citet{dunfield2019bidirectional} provides accessible introduction to bidirectional typing.
 
@@
 
-As usual elim terms synthesize types ($\Rightarrow$) and intro terms are checked against the given type ($\Leftarrow$). Canonical forms are always synthesized. For the pure fragment:
+As usual, elim terms synthesize types ($\syn$) and intro terms are checked against the given type ($\chk$). Canonical forms are always synthesized. For the pure fragment:
 
@@
 	\begin{tabular}{ll}
-		$\Delta \vdash K \Rightarrow A [N']$ & Elim term K has type A and canonical form $N'$ \\
-		$\Delta \vdash M \Leftarrow A [M']$ & Intro term M has type A and canonical form $M'$
+		$\Delta \vdash K \syn A [N']$ & Elim term K has type A and canonical form $N'$ \\
+		$\Delta \vdash M \chk A [M']$ & Intro term M has type A and canonical form $M'$
 	\end{tabular}
 \end{center}
+
+Note that the blue colored $\chk$ symbol we use in the typing rules lives in a different syntactic category from the $\Leftarrow$ symbol used in the syntax (\cref{fig:syntax}) that binds the output of primitive commands.
 
@@
 	\begin{tabular}{ll}
-		$\Delta; P \vdash E \Rightarrow x:A.Q[E']$ & Computation E with precondition P has \textit{strongest} postcondition Q\\ & and returns value x of type A. Its canonical form is $E'$. \\
-		$\Delta; P \vdash E \Leftarrow x:A.Q[E']$ & Computation E with precondition P has postcondition Q\\ & and returns value x of type A. Its canonical form is $E'$.
+		$\Delta; P \vdash E \syn x:A.Q[E']$ & Computation E with precondition P has \textit{strongest} postcondition Q\\ & and returns value x of type A. Its canonical form is $E'$. \\
+		$\Delta; P \vdash E \chk x:A.Q[E']$ & Computation E with precondition P has postcondition Q\\ & and returns value x of type A. Its canonical form is $E'$.
 	\end{tabular}
 \end{center}
+
+Even though, we have shown the full forms of these judgments above, we will omit the canonical forms while presenting typing rules in the next section as they crowd the rules and do not affect the insight to be gained from them.
+
+Finally, the judgement $\Delta; \Psi; \Gamma_1 \Longrightarrow \Gamma_2$ encodes the sequent calculus for the assertion logic.
+Recall that $\Delta$, $\Psi$, and $\Gamma$ denote the variable context, the quantum heap context and the assertion context, respectively.
 
@@
 	\inferrule
-	{\Delta \vdash M : \type{Bool}
-		\\ \Delta, x : \type{Qbit}; P \circ (x \mapsto \kw{state}(M)) \vdash E \Rightarrow y:B . Q}
-	{\Delta; P \vdash x \Leftarrow \kw{mkQbit} M; E \Rightarrow y:B. (\exists x:\type{Qbit}.Q)}
+	{\Delta \vdash M \chk \type{Bool}
+		\\ \Delta, x : \type{Qbit}; P \circ (x \mapsto \kw{state}(M)) \vdash E \syn y:B . Q}
+	{\Delta; P \vdash x \Leftarrow \kw{mkQbit} M; E \syn y:B. (\exists x:\type{Qbit}.Q)}
 \end{mathpar}
 where x is fresh and \texttt{state} is a function that translates a classical representation of quantum state (such as boolean here) to the equivalent quantum state vector.
+
+Initialization can only be performed for a term $M$ that can be type checked as a \type{Bool}.
+Then, we look at the rest of the computation, $E$, in a context that adds $x$ of type \type{Qbit} under a precondition that extends the previous precondition, $P$, with the strongest postcondition for initialization, that is, the newly bound variable $x$ points to the representation of a new qubit state; this should synthesize the strongest postcondition (with respect to the expanded precondition in the context), $Q$, for $E$.
+
+To avoid dangling variables, we need to existentially quantify $x$ in the postcondition, $Q$, of the conclusion.
 
@@
 	\inferrule
-	{\Delta \vdash M : \type{Qbit}
-		\\ \Delta \vdash P \supset (M \hookrightarrow -)
-		\\ \Delta; P \circ ((M \mapsto -) \multimap \kw{emp}) \vdash E \Rightarrow y:B . Q}
-	{\Delta; P \vdash x \Leftarrow \kw{measQbit} M; E \Rightarrow y:B. Q}
+	{\Delta \vdash M \chk \type{Qbit}
+		\\ \Delta; \Psi; P \Longrightarrow (M \hookrightarrow -)
+		\\ \Delta, x:\type{Bool}; P \circ ((M \mapsto -) \multimap \kw{emp}) \vdash E \syn y:B . Q}
+	{\Delta; P \vdash x \Leftarrow \kw{measQbit} M; E \syn y:B. (\exists x:\type{Bool}.Q)}
 \end{mathpar}
 
-In measurement, we first need to ensure that the precondition P implies that M is allocated in the heap and that the fragment of heap that only refers to location M becomes empty.
+The measurement rule can be read in a similar way as the previous rule.
+But in measurement, we need to additionally prove that the verification condition (written as a sequent), the precondition $P$ implies that the qubit $M$ is allocated, holds in the heap context $\Psi$.
+The strongest postcondition here is that the fragment of heap that only refers to location M becomes empty.
 
@@
 This step also involves generating verification conditions that cannot be checked during this first phase of typechecking.
-We are also looking at alternatives such as the recently proposed sum-over-paths action semantics \cite{amy18}.
+We are also looking at alternatives such as the recently proposed sum-over-paths action semantics by \citet{amy18}.
 
@@
 \subsubsection*{Hello Quantum World}
-We start out in an empty state and assert that the result is equal to \texttt{false}.
+\leavevmode
+
+\begin{minipage}{0.95\linewidth}
 \begin{lstlisting}[language=QHaskell]
@@
 \end{lstlisting}
+\end{minipage}
+
+In this trivial program where we initialize a new qubit with \texttt{false} and then immediately measure it, we assert that the result is equal to \texttt{false}.
+The complete specification also implicitly says that the existing quantum state is not affected in any way as both the pre and postcondition include the \texttt{emp} assertion.
 
 \subsubsection*{Coin Toss}
-As the result can be in either of its two values \texttt{false} or \texttt{true}, we need not specify any special postcondition as type checking the result will be sufficient for correctness.
+\leavevmode
 
+\begin{minipage}{0.95\linewidth}
 \begin{lstlisting}[language=QHaskell]
@@
 \end{lstlisting}
+\end{minipage}
+
+In the case of quantum coin toss, the result can be in either of its two boolean values \texttt{false} or \texttt{true}, hence, we need not specify any special postcondition as type checking the result will be sufficient for correctness.
 
 \subsubsection*{Testing Bell Pair}
-Here, we are asserting that the returned booleans $a$ and $b$ hold the same value, which is what we expect from a Bell pair.
+\leavevmode
+\todo{TODO: Should the type be \type{Bool} $\otimes$ \type{Bool}?}
 
-\todo{TODO: The type should be \type{Bool} $\otimes$ \type{Bool}}
+\begin{minipage}{0.95\linewidth}
 \begin{lstlisting}[language=QHaskell]
@@
 \end{lstlisting}
+\end{minipage}
+
+Here, we are asserting that the returned booleans $a$ and $b$ hold the same value, which is what we expect from the first Bell state prepared in this program.
 
@@
 For illustration, we show the assertions annotated as comments as we step through the \texttt{testBell} program below.
-We have elided specific details about unitary application and definition of the previously mentioned \texttt{state} function as they are still being worked out.
+We have elided specific details about unitary application as they are still being worked out.
 
@@
          = do qa <= mkQbit false;
-              -- P1: P0 \o (qa |-> state(false))
+              -- P1: P0 \o (qa |-> |0\>)
               applyU (H qa);
-              -- P2: P1 \o (qa |-> state({(0.5, false), (0.5, true)}))
+              -- P2: P1 \o ((qa |-> |0\>) -o (qa |-> |+\>))
               qb <= mkQbit false;
-              -- P3: P2 \o (qb |-> state(false))
+              -- P3: P2 \o (qb |-> |0\>)
               applyU (ifQ qa (X qb));
-              -- P4: P3 \o ((qa, qb) |-> state({(0.5, (false, false),
-              --                              (0.5, (true, true))}))
+              -- P4: P3 \o ((qa |-> |+\>, qb |-> |0\>) -o (qa, qb) |-> |\Phi+\>)
               (measQbit qa, measQbit qb)
@@
 \end{lstlisting}
-Finally, the \texttt{consq} rule should ensure that the assertion P5 can lead to the specified postcondition in the type of the program.
 
-We show some two more examples in the appendix for a modular version of Bell pair generation (\cref{app:mbp}) that shows use of higher-order features of QHTT and another for teleportation protocol (\cref{app:teleport}).
+In the assertion P2, we show that only the portion of the quantum state referred to by the \texttt{applyU} command is affected.
+Similarly, in P4 we see that only the relevant portion of the heap containing the control and target qubits used in the CNOT operation are affected.
+In this case, we also needed to combine the two qubit state as they are now entangled.
+The last assertion, P5, merely encodes the strongest postcondition for measurement for each of the two qubits.
+
+Finally, the \texttt{consq} rule of HTT ensures that the assertion P5 leads to the specified postcondition in the type of the program.
+
+We show some two more examples in the appendix for a modular version of Bell pair generation (\Cref{app:mbp}) that shows use of higher-order features of QHTT and another for teleportation protocol (\Cref{app:teleport}).
 Both of these involve specifying assertions over quantum state whose theory we still need to work out.
@@
 \paragraph{Circuits as Arrows} Further, Proto-Quipper treats quantum circuits as first class citizens of the language.
-We would like to explore modifying our theory to treat Quantum Hoare types as arrows instead of as monads as was suggested by Vizzotto et al~\cite{so-arrows}.
+We would like to explore modifying our theory to treat Quantum Hoare types as arrows instead of as monads as was suggested by \citet{so-arrows}.
 It makes sense from the perspective of sequential composition as arrows can have an arbitrary number of input/outputs as opposed to monads.
@@
 those in superposition vs.
-those in entanglement~\cite{JorrandPerdrix2009} such as those inspired by the various quantum resource theories (see \cite{rand_type_2019} and references therein).
+those in entanglement~\cite{JorrandPerdrix2009} such as those inspired by the various quantum resource theories or the Heisenberg representation of quantum mechanics~\cite{rand_type_2019,rssl20}.
 This may help us provide more specific postconditions that quantum programmers expect to hold true of their programs.

### 2020-06-08 ms-paper/ms-paper.tex
Remove conditional and move to future work
WIP

@@
 		\begin{minipage}[t]{0.45\columnwidth}%
-			$ \kw{mkQbit} M \mid \kw{measQbit} M \mid \kw{applyU} M \mid \kw{if} M \kw{then} N \kw{else} O$
+			$ \kw{mkQbit} M \mid \kw{measQbit} M \mid \kw{applyU} M$
 		\end{minipage}\\
@@
 These are the quantum-specific effectful commands from QIO for initialization, unitary application and measurement of qubits.
-We also support classical control using the \texttt{if} $M$ \texttt{then} $N$ \texttt{else} $O$ construct similar to HTT.
 
@@
 
+\textbf{Code:}
+
 \begin{minipage}{0.95\linewidth}
@@
          measQbit q
+\end{lstlisting}
+\end{minipage}
+
+\textbf{Verification steps:}
+
+\begin{minipage}{0.95\linewidth}
+\begin{lstlisting}[language=QHaskell]
+hqw : {emp} r : Bool {emp /\ Id(r, false)}
+-- P0: emp
+	= do q <= mkQbit false;
+-- P1: P0 \o (q |-> |0\>)
+		 measQbit q
+-- P2: P1 \o ((q |-> -) -o emp)
 \end{lstlisting}
@@
 There are several avenues of future work to explore:
+
+\paragraph{Quantum Language} Our programming language currently does not support conditionals or any iterative or recursive constructs. We would like to add those in future.
 

### 2020-06-08 ms-paper/ms-paper.tex
Add verification steps for first two examples

@@
 An observation is that translating the programs was a very simple process except for coming up with the right specifications for the programs.
-We take some liberty in writing our typeset syntax as plain text for convenience and provide some commentary on assertions specified for these programs.
+We provide some commentary on assertions specified for these programs.
 
@@
 
-\subsubsection*{Hello Quantum World}
+\subsubsection{Hello Quantum World}
 \leavevmode
-
-\textbf{Code:}
 
@@
 
-\textbf{Verification steps:}
+In this trivial program where we initialize a new qubit with \texttt{false} and then immediately measure it, we assert that the result is equal to \texttt{false}.
+The complete specification also implicitly says that the existing quantum state is not affected in any way as both the pre and postcondition include the \texttt{emp} assertion.
+
+Verification steps follow:
 
@@
 
-In this trivial program where we initialize a new qubit with \texttt{false} and then immediately measure it, we assert that the result is equal to \texttt{false}.
-The complete specification also implicitly says that the existing quantum state is not affected in any way as both the pre and postcondition include the \texttt{emp} assertion.
-
-\subsubsection*{Coin Toss}
+\subsubsection{Coin Toss}
 \leavevmode
@@
 
-\subsubsection*{Testing Bell Pair}
+Verification steps follow:
+
+\begin{minipage}{0.95\linewidth}
+\begin{lstlisting}[language=QHaskell]
+rnd : {emp} r : Bool {emp}
+-- P0: emp
+    = do q <= mkQbit false;
+-- P1: P0 \o (q |-> |0\>)
+         applyU (H q);
+-- P2: P1 \o ((q |-> |0\>) -o (q |-> |+\>))
+         measQbit q
+-- P3: P2 \o ((q |-> -) -o emp)
+\end{lstlisting}
+\end{minipage}
+
+\subsubsection{Testing Bell Pair}
 \leavevmode

### 2020-06-09 ms-paper/ms-paper.tex
Switch to Pierce & Turner (2000) notation for typing
The use of Rightarrow, etc is justified in binding because we want
to treat assignment differently from equality, like in Haskell. See
http://www.haskellforall.com/2017/07/demistifying-haskell-assignment.html

We use Pierce and Turner to avoid confusion with monading binding
notation even though, <= and => are way more common in papers since
2008, see https://arxiv.org/pdf/1908.05839v1.pdf#page=32 and sec. 11
in the same survey paper by Dunfield and Krishnaswami (2019).

Further, as pointed out in the same section, "vertical arrows are
uncle

@@
 \newcommand{\syncolor}{dRed}
-\newcommand{\chk}{\,\textcolor{\chkcolor}{\Leftarrow}\,}
-\newcommand{\uncoloredsyn}{{\Rightarrow}}
+\newcommand{\chk}{\,\textcolor{\chkcolor}{{\stackrel{\leftarrow}{\in}}}\,}
+\newcommand{\uncoloredsyn}{{{\stackrel{\rightarrow}{\in}}}}
 \newcommand{\syn}{\,\textcolor{\syncolor}{\uncoloredsyn}\,}
@@
 
-Note that the blue colored $\chk$ symbol we use in the typing rules lives in a different syntactic category from the $\Leftarrow$ symbol used in the syntax (\cref{fig:syntax}) that binds the output of primitive commands.
-
 The judgments for computations involve synthesizing the strongest postconditions and checking whether a given postcondition applies:
@@
     = do q <= mkQbit false;
-         measQbit q
+         return measQbit q
 \end{lstlisting}
@@
 -- P1: P0 \o (q |-> |0\>)
-         measQbit q
+         return measQbit q
 -- P2: P1 \o ((q |-> -) -o emp)
@@
          applyU (H q);
-         measQbit q
+         return measQbit q
 \end{lstlisting}
@@
 -- P2: P1 \o ((q |-> |0\>) -o (q |-> |+\>))
-         measQbit q
+         return measQbit q
 -- P3: P2 \o ((q |-> -) -o emp)
@@
               applyU (ifQ qa (X qb));
-              (measQbit qa, measQbit qb)
+              return (measQbit qa, measQbit qb)
 \end{lstlisting}
@@
               -- P4: P3 \o ((qa |-> |+\>, qb |-> |0\>) -o (qa, qb) |-> |\Phi+\>)
-              (measQbit qa, measQbit qb)
+              return (measQbit qa, measQbit qb)
               -- P5: P4 \o ((qa |-> -) -o emp) \o ((qb |-> -) -o emp)
@@
          = do (qa, qb) <- bell;
-              (measQbit qa, measQbit qb)
+              return (measQbit qa, measQbit qb)
 \end{lstlisting}

### 2020-06-09 ms-paper/ms-paper.tex
Add note for consequence of removal of conditional

@@
                  applyU (H a);
-                 (measQbit a, measQbit e)
+                 return (measQbit a, measQbit e)
 \end{lstlisting}
 
-\subsection{Bob's circuit}
+\subsection{Bob's circuit} \todo{TODO: With \texttt{if ... then ... else} removed, this will not work}
 \begin{lstlisting}[language=QHaskell]

### 2020-06-09 ms-paper/ms-paper.tex
Add missing notation for the first bell pair

@@
 	{\\in}{{$\in$}}1 {\\o}{{$\circ$}}1 {\\Pi}{{$\Pi$}}1
+	{\\Phi+}{{$\Phi^+$}}2
 	{|->}{{$\mapsto$}}2 {-o}{{$\multimap$}}2

### 2020-06-12 ms-paper/ms-paper.tex
WIP
Better notation for semantic proof obligation

@@
 \newcommand{\syn}{\,\textcolor{\syncolor}{\uncoloredsyn}\,}
+
+\newcommand{\prove}{\models}
 
@@
 
-Finally, the judgement $\Delta; \Psi; \Gamma_1 \Longrightarrow \Gamma_2$ encodes the sequent calculus for the assertion logic.
+Finally, the judgement $\Delta; \Psi; \Gamma_1 \prove \Gamma_2$ encodes the sequent calculus for the assertion logic.
 Recall that $\Delta$, $\Psi$, and $\Gamma$ denote the variable context, the quantum heap context and the assertion context, respectively.
@@
 	{\Delta \vdash M \chk \type{Qbit}
-		\\ \Delta; \Psi; P \Longrightarrow (M \hookrightarrow -)
+		\\ \Delta; \Psi; P \prove (M \hookrightarrow -)
 		\\ \Delta, x:\type{Bool}; P \circ ((M \mapsto -) \multimap \kw{emp}) \vdash E \syn y:B . Q}
@@
 \end{minipage}
+
+\begin{mathpar}
+	\inferrule
+	{ \exists \Psi . (emp * \top) \vdash q \Leftarrow mkQbit false; return measQbit q \chk r : Bool . \forall \Psi . (emp \multimap emp \wedge Id(r, false)) }
+	{do (q \Leftarrow mkQbit false; return measQbit q) \chk \{emp\} r : Bool \{emp \wedge Id(r, false)\}}
+\end{mathpar}
 

### 2020-06-12 ms-paper/ms-paper.tex
Almost done proof for typechecking HQW

@@
 
+WIP derivation tree for above program:
+
 \begin{mathpar}
-	\inferrule
-	{ \exists \Psi . (emp * \top) \vdash q \Leftarrow mkQbit false; return measQbit q \chk r : Bool . \forall \Psi . (emp \multimap emp \wedge Id(r, false)) }
-	{do (q \Leftarrow mkQbit false; return measQbit q) \chk \{emp\} r : Bool \{emp \wedge Id(r, false)\}}
+	\inferrule*[right=HIntro]
+	{\inferrule*[right=consq]
+		{\inferrule*[right=mkQbit]
+			{\inferrule*[right=measQbit]
+				{\inferrule*[right=return]
+					{q : \type{Qbit}, r : \type{Bool} \vdash r \chk \type{Bool}}
+					{q : \type{Qbit}, r : \type{Bool}; P_1 \circ ((q \mapsto -) \multimap emp) \vdash \kw{return} r \syn r : \type{Bool} . R}
+					\\ q : \type{Qbit} \vdash q \chk \type{Qbit}
+					\\ \cdot; P_0 * (q \mapsto \ket{0}) \prove (q \hookrightarrow -)
+				}
+				{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{measQbit} q; E_3 \syn r : \type{Bool} . (\exists r : \type{Bool}.R)}
+				\\ \cdot \vdash \kw{false} \chk \type{Bool}
+			}
+			{
+				\cdot; P_0 \vdash q \Leftarrow \kw{mkQbit false}; E_{23} \syn r : \type{Bool} . (\exists q : \type{Qbit}.R)
+			}
+			\\ r : \type{Bool}; i,m ; R \prove emp \multimap emp \wedge Id(r, \kw{false})
+		}
+		{
+			\cdot; \lambda i . \lambda m . i = m \wedge (emp * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bool} . (emp \multimap emp \wedge Id(r, \kw{false}))
+		}
+	}
+	{\cdot \vdash \kw{do} E \chk \{emp\} r : \type{Bool} \{emp \wedge \kw{Id}(r, \kw{false})\}}
 \end{mathpar}

### 2020-10-01 ms-paper/ms-paper.tex
Add city
Add implementation notes and citations

@@
 	\institution{University of Chicago}
+	\city{Chicago}
 	\state{IL}
@@
 
+\section{Implementation}
+We are implementing QHTT in the F* programming language.
+
+One issue with the original formulation of Hoare Type Theory is that using Hoare triples does not directly lead to a verification condition generation algorithm.
+This results in requiring a complicated bidirectional typing system for implementation.
+If we use predicate transformers as demonstrated by the use of Dijkstra monads in F*, we can avoid these issues \cite{swamy2013verifying}.
+
+The current steps for the implementation are:
+
+\begin{itemize}
+	\item Design a memory model for representing quantum state, along with elementary operations such as read, write, lookup, etc
+	\item Expose a new Quantum effect in F* by using the Dijkstra Monads for Free technique \cite{dm4free2017}.
+	\item Port our language to work well with F* syntax.
+\end{itemize}
+
+This will necessarily take us away from the bidirectional typing approach used in the previous section.
+But the end result for the user would be the same.
+Indeed, F* system is an evolution of the HTT approach that comes with inherent support for effects and automation using an SMT solver.
+
 %\section{Evaluation}

### 2020-10-01 ms-paper/ms-paper.tex
Add note about Unruh's QHL work

@@
 Various Quantum Hoare Logics that exist~\cite{ying_floydhoare_2012} currently do not support frame rules that provide Separation Logic its power.
+The closest approach is that taken by \citet{unruh2019} that allows assertions on quantum state similar to what we would like in our setting, such as whether certain qubits are in classical state, or in uniform superposition, entangled or separable.
+We hope to incorporate these ideas in this work; this may require us to use a subspace semantics for the quantum state as shown in \citet{unruh2019}.
 

### 2020-10-01 ms-paper/ms-paper.tex
Comment out previous derivation

@@
 
-WIP derivation tree for above program:
+% WIP derivation tree for above program:
 
-\begin{mathpar}
-	\inferrule*[right=HIntro]
-	{\inferrule*[right=consq]
-		{\inferrule*[right=mkQbit]
-			{\inferrule*[right=measQbit]
-				{\inferrule*[right=return]
-					{q : \type{Qbit}, r : \type{Bool} \vdash r \chk \type{Bool}}
-					{q : \type{Qbit}, r : \type{Bool}; P_1 \circ ((q \mapsto -) \multimap emp) \vdash \kw{return} r \syn r : \type{Bool} . R}
-					\\ q : \type{Qbit} \vdash q \chk \type{Qbit}
-					\\ \cdot; P_0 * (q \mapsto \ket{0}) \prove (q \hookrightarrow -)
-				}
-				{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{measQbit} q; E_3 \syn r : \type{Bool} . (\exists r : \type{Bool}.R)}
-				\\ \cdot \vdash \kw{false} \chk \type{Bool}
-			}
-			{
-				\cdot; P_0 \vdash q \Leftarrow \kw{mkQbit false}; E_{23} \syn r : \type{Bool} . (\exists q : \type{Qbit}.R)
-			}
-			\\ r : \type{Bool}; i,m ; R \prove emp \multimap emp \wedge Id(r, \kw{false})
-		}
-		{
-			\cdot; \lambda i . \lambda m . i = m \wedge (emp * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bool} . (emp \multimap emp \wedge Id(r, \kw{false}))
-		}
-	}
-	{\cdot \vdash \kw{do} E \chk \{emp\} r : \type{Bool} \{emp \wedge \kw{Id}(r, \kw{false})\}}
-\end{mathpar}
+% \begin{mathpar}
+% 	\inferrule*[right=HIntro]
+% 	{\inferrule*[right=consq]
+% 		{\inferrule*[right=mkQbit]
+% 			{\inferrule*[right=measQbit]
+% 				{\inferrule*[right=return]
+% 					{q : \type{Qbit}, r : \type{Bool} \vdash r \chk \type{Bool}}
+% 					{q : \type{Qbit}, r : \type{Bool}; P_1 \circ ((q \mapsto -) \multimap emp) \vdash \kw{return} r \syn r : \type{Bool} . R}
+% 					\\ q : \type{Qbit} \vdash q \chk \type{Qbit}
+% 					\\ \cdot; P_0 * (q \mapsto \ket{0}) \prove (q \hookrightarrow -)
+% 				}
+% 				{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{measQbit} q; E_3 \syn r : \type{Bool} . (\exists r : \type{Bool}.R)}
+% 				\\ \cdot \vdash \kw{false} \chk \type{Bool}
+% 			}
+% 			{
+% 				\cdot; P_0 \vdash q \Leftarrow \kw{mkQbit false}; E_{23} \syn r : \type{Bool} . (\exists q : \type{Qbit}.R)
+% 			}
+% 			\\ r : \type{Bool}; i,m ; R \prove emp \multimap emp \wedge Id(r, \kw{false})
+% 		}
+% 		{
+% 			\cdot; \lambda i . \lambda m . i = m \wedge (emp * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bool} . (emp \multimap emp \wedge Id(r, \kw{false}))
+% 		}
+% 	}
+% 	{\cdot \vdash \kw{do} E \chk \{emp\} r : \type{Bool} \{emp \wedge \kw{Id}(r, \kw{false})\}}
+% \end{mathpar}
 

### 2020-10-09 ms-paper/ms-paper.tex
Move previous derivation to appendix
Use correct notation for pair type

--- 2020-10-01 ms-paper/ms-paper.tex
@@
 \crefformat{section}{\S#2#1#3}
+
+\usepackage{rotating}
 
@@
 	{|->}{{$\mapsto$}}2 {-o}{{$\multimap$}}2
+	{o*}{{$\otimes$}}1
 }
@@
 
-WIP derivation tree for above program:
-
-\begin{mathpar}
-	\inferrule*[right=HIntro]
-	{\inferrule*[right=consq]
-		{\inferrule*[right=mkQbit]
-			{\inferrule*[right=measQbit]
-				{\inferrule*[right=return]
-					{q : \type{Qbit}, r : \type{Bool} \vdash r \chk \type{Bool}}
-					{q : \type{Qbit}, r : \type{Bool}; P_1 \circ ((q \mapsto -) \multimap emp) \vdash \kw{return} r \syn r : \type{Bool} . R}
-					\\ q : \type{Qbit} \vdash q \chk \type{Qbit}
-					\\ \cdot; P_0 * (q \mapsto \ket{0}) \prove (q \hookrightarrow -)
-				}
-				{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{measQbit} q; E_3 \syn r : \type{Bool} . (\exists r : \type{Bool}.R)}
-				\\ \cdot \vdash \kw{false} \chk \type{Bool}
-			}
-			{
-				\cdot; P_0 \vdash q \Leftarrow \kw{mkQbit false}; E_{23} \syn r : \type{Bool} . (\exists q : \type{Qbit}.R)
-			}
-			\\ r : \type{Bool}; i,m ; R \prove emp \multimap emp \wedge Id(r, \kw{false})
-		}
-		{
-			\cdot; \lambda i . \lambda m . i = m \wedge (emp * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bool} . (emp \multimap emp \wedge Id(r, \kw{false}))
-		}
-	}
-	{\cdot \vdash \kw{do} E \chk \{emp\} r : \type{Bool} \{emp \wedge \kw{Id}(r, \kw{false})\}}
-\end{mathpar}
-
 \subsubsection{Coin Toss}
@@
 
+\section{Sample Proof}
+Complete derivation tree for Hello Quantum World:
+
+\begin{turn}{90}
+	\begin{minipage}{40pc}
+	\begin{mathpar}
+		\inferrule*[right=HIntro]
+		{\inferrule*[right=consq]
+			{\inferrule*[right=mkQbit]
+				{\inferrule*[right=measQbit]
+					{\inferrule*[right=return]
+						{q : \type{Qbit}, r : \type{Bool} \vdash r \chk \type{Bool}}
+						{q : \type{Qbit}, r : \type{Bool}; P_1 \circ ((q \mapsto -) \multimap emp) \vdash \kw{return} r \syn r : \type{Bool} . R}
+						\\ q : \type{Qbit} \vdash q \chk \type{Qbit}
+						\\ \cdot; P_0 * (q \mapsto \ket{0}) \prove (q \hookrightarrow -)
+					}
+					{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{measQbit} q; E_3 \syn r : \type{Bool} . (\exists r : \type{Bool}.R)}
+					\\ \cdot \vdash \kw{false} \chk \type{Bool}
+				}
+				{
+					\cdot; P_0 \vdash q \Leftarrow \kw{mkQbit false}; E_{23} \syn r : \type{Bool} . (\exists q : \type{Qbit}.R)
+				}
+				\\ r : \type{Bool}; i,m ; R \prove emp \multimap emp \wedge Id(r, \kw{false})
+			}
+			{
+				\cdot; \lambda i . \lambda m . i = m \wedge (emp * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bool} . (emp \multimap emp \wedge Id(r, \kw{false}))
+			}
+		}
+		{\cdot \vdash \kw{do} E \chk \{emp\} r : \type{Bool} \{emp \wedge \kw{Id}(r, \kw{false})\}}
+	\end{mathpar}
+	\end{minipage}
+\end{turn}
+
 \section{Modular Bell Pair}
@@
 \subsection{Bell pair}
-\todo{TODO: The type should be \type{Qbit} $\otimes$ \type{Qbit}}
 \begin{lstlisting}[language=QHaskell]
-bell : {emp} (a, b) : (Qbit, Qbit) {Id(a, b) /\ a \in {|0\>, |1\>}}
+bell : {emp} (a, b) : (Qbit o* Qbit) {Id(a, b) /\ a \in {|0\>, |1\>}}
      = do qa <- qplus;
@@
 \begin{lstlisting}[language=QHaskell]
-testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}
+testBell : {emp} (a, b) : (Bool o* Bool) {emp /\ Id(a, b)}
          = do (qa, qb) <- bell;
@@
         {(Id(a, -) /\ entangled(e)}
-           r : (Bool, Bool)
+           r : (Bool o* Bool)
         {emp}

### 2020-10-09 ms-paper/ms-paper.tex
Add notes from the Unruh paper on ghost variables

@@
 We hope to incorporate these ideas in this work; this may require us to use a subspace semantics for the quantum state as shown in \citet{unruh2019}.
+We share preliminary work in that direction in the appendix below.
 
@@
 
+\section{Ideas from Unruh 2019}
+\subsection{Model of the Quantum State}
+A program starts with a fixed but unknown number of qubits. For simplicity, assume they are countably infinite.
+With access to commands for initialization, the program can allocate as many qubits as it needs.
+In our model of the quantum state, the qubits are divided into a few categories: those that are being used (ie, that have been allocated) by the program (program variables), denoted by X; those that are ghosts (only used in the logic, but not available to the program), denoted by G and the rest that are unallocated.
+The ghost qubits are then further categorized into entangled (E) and unentangled (U), which mean exactly what they are called.
+
+All of these variables draw values from classical sets of values such as {0, 1}, that can be considered as their type.
+For operations such as substitution, it is important that the types match.
+At this point, we should make it clear that we are not limited to treating a single variable as a single qubit anymore; a variable can represent many qubits together, for example, if the type of the variable is {0, 1, 2, 3}, it necessarily requires at least two qubits to store these classical values.
+
+Now, the pure quantum state accessible to the program can be considered as the superposition (linear combination) of all of the basis vectors that span the Hilbert space corresponding to the given qubits.
+
+We will also need to talk about mixed states as we deal with measurements that can result in probabilistic outcomes.
+We represent mixed states using density operators, $\rho = \Sigma_i p_i \mathrm{proj}(\psi_i)$, where $p_i$ represents the probability of the state being $\psi_i$.
+
+Applying unitaries to the quantum state is same as before, in that, we apply the operation to the specified variables and identity to all other program variables.
+
+\subsection{Quantum Assertion Language}
+We need to be explicit about what we mean when we say that a certain state satisifes a certain assertion (predicate) in the quantum setting.
+In this model, the predicates are subspaces of the Hilbert space corresponding to a quantum state.
+Here are some examples:
+\begin{itemize}
+	\item $\top$ represents the complete Hilbert space that is always satisfied by any state.
+	\item $\bot$ represents nothing, it is never satisfied.
+	\item $A \wedge B$ represent that both A and B hold (conjunction), it is equivalent to the intersection of the subspaces.
+	\item $A \vee B$ similarly represents disjunction and is modeled as the sum (linear combination) of the subspaces A and B.
+	\item If an operator M is applied to a state satisfying a predicate A, then the new predicate can be written as $M \cdot A$.
+	\item If we want to write that a variable w, or a list of variables W is in a specific state, we can model that predicate as the variable(s) being in the span of the state and use the notation, $W =_q \psi$.
+\end{itemize}
+
+Here are some examples.
+The postcondition of the program \texttt{qplus} above can be written as $r =_q \ket{+}$.
+The postcondition for the program \texttt{bell} as $ab =_q \frac{1}{\sqrt{2}}\ket{00}+\frac{1}{\sqrt{2}}\ket{11}$ and the precondition for the program \texttt{share} as $a =_q \ket{+} \vee a =_q \ket{-}$.
+
+What is the relation between the program variables and the subspaces corresponding to the predicates that we are talking about?
+We need to define the notion of free variables for a predicate $A$.
+fv(A) consists of those variables that solely describe the predicate A.
+
+There are more predicates that will involve ghosts, described in the section after the next.
+
+\subsection{Improvements to the quantum commands}
+\todo{To Do: add if then else, skip, sequencing, and a simpler form of apply; and the syntactic sugar for quantum and classical assignments}
+
+\subsection{Predicates with Ghosts}
+\todo{To Do: add distrib, uniform, separable, classical, etc}
+
+\subsection{Rules of the Hoare logic}
+
+Core rules:
+
+\todo{To Do: total 11 core rules: 6 for the language primitives, 5 structural. Out of which two are for case distinction and the remaining three deal with ghosts}
+
+Derived rules:
+
+\todo{To Do: conseq and others as needed}
+
+\subsection{Remarks}
+Now, the question is whether we can use this memory model (based on subspsace semantics), assertion logic and rules and still continue to use bidirectional typing for the type system?
+If so, it's probably worth finishing up the rule for unitary application using this memory model.
+
+The implementation notes suggest that switching to F* seems to require making too many changes, both in the syntax of the language and in the core theory; which is something probably best left for a future implementation project.
+
 %\section{Sample Proof}

### 2020-10-16 ms-paper/ms-paper.tex
Leave more space above the vertical proof tree
Update status in appendix

@@
 \begin{turn}{90}
-	\begin{minipage}{40pc}
+	\begin{minipage}{42pc}
 	\begin{mathpar}
@@
 \subsection{Improvements to the quantum commands}
-\todo{To Do: add if then else, skip, sequencing, and a simpler form of apply; and the syntactic sugar for quantum and classical assignments}
+
+Added if then else, skip, sequencing, and a simpler form of apply; and the syntactic sugar for quantum and classical assignments in the ott document.
 
 \subsection{Predicates with Ghosts}
-\todo{To Do: add distrib, uniform, separable, classical, etc}
+\todo{To Do: consider adding distrib}
+
+Added uniform, separable and classical in the ott document. distrib seems too general for our needs.
 
@@
 
+\subsection{Implementation Options}
+
+The semantics considered in this work uses abstract linear algebra without depending on a concrete representation (such as on matrices) but current library support for such abstract linear algebra (such as vector spaces and subspaces) in dependent type theoretic systems such as Coq seems limited.
+
+I found Georges Gonthiers' ITP 2011 paper ``Point-Free, Set-Free Concrete Linear Algebra'' (\url{https://hal.inria.fr/hal-00805966/document}) that models abstract linear algebra using matrices but still exposes a point-free interface to vector spaces still in line with our needs.
+The implementation seems to be what is now the mathcomp.algebra.vector library (\url{https://math-comp.github.io/htmldoc/mathcomp.algebra.vector.html}).
+There are couple of concerns:
+
+\begin{itemize}
+	\item this library is limited to finite dimensional vector spaces which is probably okay for our needs but not the same as the infinite dimension semantics presented in Unruh's work.
+	\item since it is still dependent on a concrete matrix representation, it does not give us complete flexibility in renamings of the quantum variables under consideration.
+I need to investigate this further.
+	\item It's unclear to me how production-ready this library might be as I could not find examples of it being used in a project.
+The couple of Coq workshops that I found stop at the matrix algebra library (\url{https://math-comp.github.io/htmldoc/mathcomp.algebra.mxalgebra.html}).
+\end{itemize}
+
+None of these seem completely impossible to overcome for our needs, but we may run into hiccups here and there.
+
+Lean seems to have a better support for linear algebra (\url{https://leanprover-community.github.io/mathlib-overview.html}) but it may require significantly more work to understand that system.
+
+Robert suggests another approach that is to limit the expressivity of the system and implement the limited logic proposed in Häner, Hoefler and Troyer's ``Using Hoare logic for quantum circuit optimization'' (\url{https://arxiv.org/abs/1810.00375}), an improved version of which is going to be presented at OOPSLA in November.
+This may even be possible in F* with some effort.
+
 \subsection{Remarks}

### 2020-10-25 ms-paper/ms-paper.tex
Reduce dependence on QIO

@@
 	%Task
-	Inspired by Hoare Type Theory in the classical paradigm, we propose Quantum Hoare Types by extending the Quantum IO Monad by indexing it with pre- and post-conditions that serve as program specifications.
+	Inspired by Hoare Type Theory in the classical paradigm, we propose Quantum Hoare Types in which a quantum effect monad is indexed with pre- and post-conditions that serve as program specifications.
 	%Object
@@
 As programming languages researchers, we think it will be better to encode such assertions into a static type system both for formal verification and to aid the programmers in writing correct programs from the start.
-Inspired by the use of Hoare triples in the verification of imperative programs and building on the idea of Hoare Type Theory (HTT) for classical programming languages~\cite{nanevski_hoare_2008}, we extend the Quantum IO Monad interface~\cite{qio} and propose Quantum Hoare Type Theory aimed at enabling both sound static type checking and formal verification of quantum programs.
+Inspired by the use of Hoare triples in the verification of imperative programs and building on the idea of Hoare Type Theory (HTT) for classical programming languages~\cite{nanevski_hoare_2008}, we extend a quantum effect monad inspired by the Quantum IO Monad interface~\cite{qio} and propose Quantum Hoare Type Theory aimed at enabling both sound static type checking and formal verification of quantum programs.
 

### 2020-10-25 ms-paper/ms-paper.tex
Update syntax of assertions and q. commands

@@
 		\begin{minipage}[t]{0.45\columnwidth}%
-			$ \type{1} \mid \type{Bool} \mid \type{Qbit} \mid \type{U} \mid \type{Pure} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
+			$ \type{1} \mid \type{Bool} \mid \type{Qbit} \mid \type{U} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
 		\end{minipage}\\ \\
@@
 		\begin{minipage}[t]{0.45\columnwidth}%
-			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid P \supset Q \mid \neg P \mid \exists x{:}A.P \mid \forall x{:}A.P \mid \exists h{:}\kw{heap}.P \mid \forall h{:}\kw{heap}.P \mid \kw{Id}_A(M, N) \mid \kw{HId}(H, G) \mid \kw{indom}(H,M) $
+			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid M \cdot P \mid X \in_q S \mid X =_q \psi \mid \kw{uniform}(X) \mid \kw{separable}(X) \mid \kw{classical}(X) $
 		\end{minipage}\\ \\
-		\textit{Q. Heaps} & $H, G$ & ::= & $ h \mid \kw{empty} \mid \kw{upd}(H, M, N) $\\ \\
-		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K M \mid M : A $\\
-		\textit{Intro terms} & $M, N, O$ & ::= & $ K \mid () \mid \lambda x.M \mid \kw{do} E \mid \kw{true} \mid \kw{false} $\\ \\
+		% \textit{Q. Heaps} & $H, G$ & ::= & $ h \mid \kw{empty} \mid \kw{upd}(H, M, N) $\\ \\
+		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K N \mid N : A $\\
+		\textit{Intro terms} & $N, O$ & ::= & $ K \mid () \mid \lambda x.N \mid \kw{do} E \mid \kw{true} \mid \kw{false} $\\ \\
 		\textit{Q. Commands} & $c$ & ::= &
 		\begin{minipage}[t]{0.45\columnwidth}%
-			$ \kw{mkQbit} M \mid \kw{measQbit} M \mid \kw{applyU} M$
+			$ \kw{init} N \mid \kw{apply} U \kw{to} N \mid \kw{measure} N \mid \kw{if} y \kw{then} E \kw{else} F$
 		\end{minipage}\\
-		\textit{Computations} & $E, F$ & ::= & $ \kw{return} M \mid x \leftarrow K; E \mid x \Leftarrow c; E \mid x =_A M; E $\\ \\
+		\textit{Computations} & $E, F$ & ::= & $ \kw{return} N \mid x \leftarrow K; E \mid x \Leftarrow c; E \mid x =_A N; E $\\ \\
 		\textit{Variable context} & $\Delta$ & ::= & $ \cdot \mid x : A, \Delta $ \\
@@
 \paragraph{Assertions}
-Apart from the usual first order logic, we have assertions for reasoning about propositional equality of terms, $\kw{Id}_A(M, N)$, and heaps, $\kw{HId}(H, G)$.
+The assertion language is inspired by \citet{unruh2019} that follows subspace semantics.
+We need to be explicit about what we mean when we say that a certain state satisifes a certain assertion (predicate) in the quantum setting.
+In this model, the predicates are subspaces of the Hilbert space corresponding to a quantum state.
+Here are some examples:
+\begin{itemize}
+	\item $\top$ represents the complete Hilbert space that is always satisfied by any state.
+	\item $\bot$ represents nothing, it is never satisfied.
+	\item $P \wedge Q$ represent that both P and Q hold (conjunction), it is equivalent to the intersection of the subspaces.
+	\item $P \vee Q$ similarly represents disjunction and is modeled as the sum (linear combination) of the subspaces P and Q.
+	\item If an operator M is applied to a state satisfying a predicate P, then the new predicate can be written as $M \cdot P$.
+	\item If we want to write that a variable w, or a list of variables W is in a specific state, we can model that predicate as the variable(s) being in the span of the state and use the notation, $W =_q \psi$.
+\end{itemize}
 
-As done in HTT, some common convenience assertions can be defined using the base primitives.
-Particularly, $\kw{emp}$ denotes that the current heap, say $h$, is empty, that is, $\kw{HId}(h, \kw{empty})$.
-$M \mapsto N$ represents the only mapping in the (singleton) heap.
-$M \hookrightarrow N$ says that looking up $M$ in the heap returns $N$ which can be more explicitly written as $\kw{seleq}(H, M, N)$.
+Here are some examples.
+The postcondition of the program \texttt{qplus} can be written as $r =_q \ket{+}$.
+The postcondition for the program \texttt{bell} as $ab =_q \frac{1}{\sqrt{2}}\ket{00}+\frac{1}{\sqrt{2}}\ket{11}$ and the precondition for the program \texttt{share} as $a =_q \ket{+} \vee a =_q \ket{-}$.
+
+There are other predicates that involve ghosts. \todo{TODO}
 
@@
 
-For the purpose of the type theory, heaps are functional, so that, for example, $\kw{upd}(H, M, N)$ returns a new heap after updating heap $H$ at location $M$ with $N$.
+% For the purpose of the type theory, heaps are functional, so that, for example, $\kw{upd}(H, M, N)$ returns a new heap after updating heap $H$ at location $M$ with $N$.
 
@@
 \paragraph{Quantum Commands}
-These are the quantum-specific effectful commands from QIO for initialization, unitary application and measurement of qubits.
+These are the quantum-specific effectful commands inspired by QIO for initialization, unitary application and measurement of qubits.
+\texttt{init} $N$ initializes a given variable based on the value of the expression $N$ (typically 0 or 1 if N is a boolean).
+\texttt{apply} $U$ \texttt{to} $X$ applies unitary operation $U$ to variables $X$.
+\texttt{measure} $x$ measures the variable $x$ in the computational basis.
+\texttt{if y then E else F} executes computation $E$ if measuring variable $y$ returns 1 else executes $F$.
 
 \paragraph{Computations} The computations are the usual monadic return and three sequencing operations that bind their result to a variable similar to HTT.
-The first, $x \leftarrow K$, executes a suspended computation (such as applying a function that includes a \texttt{do} $E$ in its body); the second, $x \Leftarrow c$ executes a primitive command $c$; and the last, $x =_A M$, is just syntactic sugar for let-binding.
+The first, $x \leftarrow K$, executes a suspended computation (such as applying a function that includes a \texttt{do} $E$ in its body); the second, $x \Leftarrow c$ executes a primitive command $c$; and the last, $x =_A N$, is just syntactic sugar for let-binding.
 

### 2020-10-25 ms-paper/ms-paper.tex
Metavariable M is now used for operations as in M.P, could use U perhaps?

@@
 		$\Delta \vdash K \syn A [N']$ & Elim term K has type A and canonical form $N'$ \\
-		$\Delta \vdash M \chk A [M']$ & Intro term M has type A and canonical form $M'$
+		$\Delta \vdash N \chk A [N']$ & Intro term N has type A and canonical form $N'$
 	\end{tabular}
@@
 \paragraph{Initialization}
-$x \Leftarrow \kw{mkQbit} M; E$
+$x \Leftarrow \kw{init} N; E$
 
@@
 	\inferrule
-	{\Delta \vdash M \chk \type{Bool}
-		\\ \Delta, x : \type{Qbit}; P \circ (x \mapsto \kw{state}(M)) \vdash E \syn y:B . Q}
-	{\Delta; P \vdash x \Leftarrow \kw{mkQbit} M; E \syn y:B. (\exists x:\type{Qbit}.Q)}
+	{\Delta \vdash N \chk \type{Bool}
+		\\ \Delta, x : \type{Qbit}; P \circ (x \mapsto \kw{state}(N)) \vdash E \syn y:B . Q}
+	{\Delta; P \vdash x \Leftarrow \kw{init} N; E \syn y:B. (\exists x:\type{Qbit}.Q)}
 \end{mathpar}
@@
 
-Initialization can only be performed for a term $M$ that can be type checked as a \type{Bool}.
+Initialization can only be performed for a term $N$ that can be type checked as a \type{Bool}.
 Then, we look at the rest of the computation, $E$, in a context that adds $x$ of type \type{Qbit} under a precondition that extends the previous precondition, $P$, with the strongest postcondition for initialization, that is, the newly bound variable $x$ points to the representation of a new qubit state; this should synthesize the strongest postcondition (with respect to the expanded precondition in the context), $Q$, for $E$.
@@
 \paragraph{Measurement}
-$x \Leftarrow \kw{measQbit} M; E$
+$x \Leftarrow \kw{measure} N; E$
 
@@
 	\inferrule
-	{\Delta \vdash M \chk \type{Qbit}
-		\\ \Delta; \Psi; P \prove (M \hookrightarrow -)
-		\\ \Delta, x:\type{Bool}; P \circ ((M \mapsto -) \multimap \kw{emp}) \vdash E \syn y:B . Q}
-	{\Delta; P \vdash x \Leftarrow \kw{measQbit} M; E \syn y:B. (\exists x:\type{Bool}.Q)}
+	{\Delta \vdash N \chk \type{Qbit}
+		\\ \Delta; \Psi; P \prove (N \hookrightarrow -)
+		\\ \Delta, x:\type{Bool}; P \circ ((N \mapsto -) \multimap \kw{emp}) \vdash E \syn y:B . Q}
+	{\Delta; P \vdash x \Leftarrow \kw{measure} N; E \syn y:B. (\exists x:\type{Bool}.Q)}
 \end{mathpar}
@@
 The measurement rule can be read in a similar way as the previous rule.
-But in measurement, we need to additionally prove that the verification condition (written as a sequent), the precondition $P$ implies that the qubit $M$ is allocated, holds in the heap context $\Psi$.
-The strongest postcondition here is that the fragment of heap that only refers to location M becomes empty.
+But in measurement, we need to additionally prove that the verification condition (written as a sequent), the precondition $P$ implies that the qubit $N$ is allocated, holds in the heap context $\Psi$.
+The strongest postcondition here is that the fragment of heap that only refers to location N becomes empty.
 

### 2020-10-25 ms-paper/ms-paper.tex
Update all examples

@@
 hqw : {emp} r : Bool {emp /\ Id(r, false)}
-    = do q <= mkQbit false;
-         return measQbit q
+    = do q <= init false;
+         return measure q
 \end{lstlisting}
@@
 -- P0: emp
-    = do q <= mkQbit false;
+    = do q <= init false;
 -- P1: P0 \o (q |-> |0\>)
-         return measQbit q
+         return measure q
 -- P2: P1 \o ((q |-> -) -o emp)
@@
 rnd : {emp} r : Bool {emp}
-    = do q <= mkQbit false;
-         applyU (H q);
-         return measQbit q
+    = do q <= init false;
+         apply H to q;
+         return measure q
 \end{lstlisting}
@@
 -- P0: emp
-    = do q <= mkQbit false;
+    = do q <= init false;
 -- P1: P0 \o (q |-> |0\>)
-         applyU (H q);
+         apply H to q;
 -- P2: P1 \o ((q |-> |0\>) -o (q |-> |+\>))
-         return measQbit q
+         return measure q
 -- P3: P2 \o ((q |-> -) -o emp)
@@
 \todo{TODO: Should the type be \type{Bool} $\otimes$ \type{Bool}?}
+\todo{TODO: syntax for applying multi-qubit gate}
 
@@
 testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}  -- a,b \in {true, false}
-         = do qa <= mkQbit false;
-              applyU (H qa);
-              qb <= mkQbit false;
-              applyU (ifQ qa (X qb));
-              return (measQbit qa, measQbit qb)
+         = do qa <= init false;
+              apply H to qa;
+              qb <= init false;
+              apply CNOT to [qa, qb];
+              return (measure qa, measure qb)
 \end{lstlisting}
@@
               -- P0: emp
-         = do qa <= mkQbit false;
+         = do qa <= init false;
               -- P1: P0 \o (qa |-> |0\>)
-              applyU (H qa);
+              apply H to qa;
               -- P2: P1 \o ((qa |-> |0\>) -o (qa |-> |+\>))
-              qb <= mkQbit false;
+              qb <= init false;
               -- P3: P2 \o (qb |-> |0\>)
-              applyU (ifQ qa (X qb));
+              apply CNOT to [qa, qb];
               -- P4: P3 \o ((qa |-> |+\>, qb |-> |0\>) -o (qa, qb) |-> |\Phi+\>)
-              return (measQbit qa, measQbit qb)
+              return (measure qa, measure qb)
               -- P5: P4 \o ((qa |-> -) -o emp) \o ((qb |-> -) -o emp)
@@
 
-In the assertion P2, we show that only the portion of the quantum state referred to by the \texttt{applyU} command is affected.
+In the assertion P2, we show that only the portion of the quantum state referred to by the \texttt{apply} command is affected.
 Similarly, in P4 we see that only the relevant portion of the heap containing the control and target qubits used in the CNOT operation are affected.
@@
 		{\inferrule*[right=consq]
-			{\inferrule*[right=mkQbit]
-				{\inferrule*[right=measQbit]
+			{\inferrule*[right=init]
+				{\inferrule*[right=measure]
 					{\inferrule*[right=return]
@@
 					}
-					{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{measQbit} q; E_3 \syn r : \type{Bool} . (\exists r : \type{Bool}.R)}
+					{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{measure} q; E_3 \syn r : \type{Bool} . (\exists r : \type{Bool}.R)}
 					\\ \cdot \vdash \kw{false} \chk \type{Bool}
@@
 				{
-					\cdot; P_0 \vdash q \Leftarrow \kw{mkQbit false}; E_{23} \syn r : \type{Bool} . (\exists q : \type{Qbit}.R)
+					\cdot; P_0 \vdash q \Leftarrow \kw{init false}; E_{23} \syn r : \type{Bool} . (\exists q : \type{Qbit}.R)
 				}
@@
 qplus : {emp} r : Qbit {Id(r, |+\>)}
-      = do q <= mkQbit false;
-           applyU (H q);
+      = do q <= init false;
+           apply H to q;
            return q
@@
 qminus : {emp} r : Qbit {Id(r, |-\>)}
-       = do q <= mkQbit true;
-            applyU (H q);
+       = do q <= init true;
+            apply H to q;
             return q
@@
         {Id(a, b) /\ a \in {|0\>, |1\>}}
-      = \a.do b <= mkQbit false;
-              applyU (ifQ a (X b));
+      = \a.do b <= init false;
+              apply CNOT to [a, b];
               return b
@@
          = do (qa, qb) <- bell;
-              return (measQbit qa, measQbit qb)
+              return (measure qa, measure qb)
 \end{lstlisting}
@@
         {emp}
-      = \a.\e.do applyU (ifQ a (X e));
-                 applyU (H a);
-                 return (measQbit a, measQbit e)
+      = \a.\e.do apply CNOT to [a, e];
+                 apply H to a;
+                 return (measure a, measure e)
 \end{lstlisting}
 
-\subsection{Bob's circuit} \todo{TODO: With \texttt{if ... then ... else} removed, this will not work}
+\subsection{Bob's circuit}
 \begin{lstlisting}[language=QHaskell]
@@
       {(Id(r, -)}
-      = \m1.\m2.\e.do if m1 then applyU (Z e) else ();
-                      if m2 then applyU (X e) else ();
+      = \m1.\m2.\e.do if m1 then (apply Z to e) else ();
+                      if m2 then (apply X to e) else ();
                       return e

### 2020-10-26 ms-paper/ms-paper.tex
emp -> \top

@@
 	{o*}{{$\otimes$}}1
+	{\\top}{{$\top$}}1
+	{\\bot}{{$\bot$}}1
 }
@@
 		\\ \Delta; \Psi; P \prove (N \hookrightarrow -)
-		\\ \Delta, x:\type{Bool}; P \circ ((N \mapsto -) \multimap \kw{emp}) \vdash E \syn y:B . Q}
+		\\ \Delta, x:\type{Bool}; P \circ ((N \mapsto -) \multimap \top) \vdash E \syn y:B . Q}
 	{\Delta; P \vdash x \Leftarrow \kw{measure} N; E \syn y:B. (\exists x:\type{Bool}.Q)}
@@
 \begin{lstlisting}[language=QHaskell]
-hqw : {emp} r : Bool {emp /\ Id(r, false)}
+hqw : {\top} r : Bool {\top /\ Id(r, false)}
     = do q <= init false;
@@
 In this trivial program where we initialize a new qubit with \texttt{false} and then immediately measure it, we assert that the result is equal to \texttt{false}.
-The complete specification also implicitly says that the existing quantum state is not affected in any way as both the pre and postcondition include the \texttt{emp} assertion.
+The complete specification also implicitly says that the existing quantum state is not affected in any way as both the pre and postcondition include the $\top$ assertion.
 
@@
 \begin{lstlisting}[language=QHaskell]
-hqw : {emp} r : Bool {emp /\ Id(r, false)}
--- P0: emp
+hqw : {\top} r : Bool {\top /\ Id(r, false)}
+-- P0: \top
     = do q <= init false;
@@
          return measure q
--- P2: P1 \o ((q |-> -) -o emp)
+-- P2: P1 \o ((q |-> -) -o \top)
 \end{lstlisting}
@@
 \begin{lstlisting}[language=QHaskell]
-rnd : {emp} r : Bool {emp}
+rnd : {\top} r : Bool {\top}
     = do q <= init false;
@@
 \begin{lstlisting}[language=QHaskell]
-rnd : {emp} r : Bool {emp}
--- P0: emp
+rnd : {\top} r : Bool {\top}
+-- P0: \top
     = do q <= init false;
@@
          return measure q
--- P3: P2 \o ((q |-> -) -o emp)
+-- P3: P2 \o ((q |-> -) -o \top)
 \end{lstlisting}
@@
 \begin{lstlisting}[language=QHaskell]
-testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}  -- a,b \in {true, false}
+testBell : {\top} (a, b) : (Bool, Bool) {\top /\ Id(a, b)}  -- a,b \in {true, false}
          = do qa <= init false;
@@
 \begin{lstlisting}[language=QHaskell]
-testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}
-              -- P0: emp
+testBell : {\top} (a, b) : (Bool, Bool) {\top /\ Id(a, b)}
+              -- P0: \top
          = do qa <= init false;
@@
               return (measure qa, measure qb)
-              -- P5: P4 \o ((qa |-> -) -o emp) \o ((qb |-> -) -o emp)
+              -- P5: P4 \o ((qa |-> -) -o \top) \o ((qb |-> -) -o \top)
 \end{lstlisting}
@@
 						{q : \type{Qbit}, r : \type{Bool} \vdash r \chk \type{Bool}}
-						{q : \type{Qbit}, r : \type{Bool}; P_1 \circ ((q \mapsto -) \multimap emp) \vdash \kw{return} r \syn r : \type{Bool} . R}
+						{q : \type{Qbit}, r : \type{Bool}; P_1 \circ ((q \mapsto -) \multimap \top) \vdash \kw{return} r \syn r : \type{Bool} . R}
 						\\ q : \type{Qbit} \vdash q \chk \type{Qbit}
@@
 				}
-				\\ r : \type{Bool}; i,m ; R \prove emp \multimap emp \wedge Id(r, \kw{false})
+				\\ r : \type{Bool}; i,m ; R \prove \top \multimap \top \wedge Id(r, \kw{false})
 			}
 			{
-				\cdot; \lambda i . \lambda m . i = m \wedge (emp * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bool} . (emp \multimap emp \wedge Id(r, \kw{false}))
+				\cdot; \lambda i . \lambda m . i = m \wedge (\top * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bool} . (\top \multimap \top \wedge Id(r, \kw{false}))
 			}
 		}
-		{\cdot \vdash \kw{do} E \chk \{emp\} r : \type{Bool} \{emp \wedge \kw{Id}(r, \kw{false})\}}
+		{\cdot \vdash \kw{do} E \chk \{\top\} r : \type{Bool} \{\top \wedge \kw{Id}(r, \kw{false})\}}
 	\end{mathpar}
@@
 \begin{lstlisting}[language=QHaskell]
-qplus : {emp} r : Qbit {Id(r, |+\>)}
+qplus : {\top} r : Qbit {Id(r, |+\>)}
       = do q <= init false;
@@
 
-qminus : {emp} r : Qbit {Id(r, |-\>)}
+qminus : {\top} r : Qbit {Id(r, |-\>)}
        = do q <= init true;
@@
 \begin{lstlisting}[language=QHaskell]
-bell : {emp} (a, b) : (Qbit o* Qbit) {Id(a, b) /\ a \in {|0\>, |1\>}}
+bell : {\top} (a, b) : (Qbit o* Qbit) {Id(a, b) /\ a \in {|0\>, |1\>}}
      = do qa <- qplus;
@@
 \begin{lstlisting}[language=QHaskell]
-testBell : {emp} (a, b) : (Bool o* Bool) {emp /\ Id(a, b)}
+testBell : {\top} (a, b) : (Bool o* Bool) {\top /\ Id(a, b)}
          = do (qa, qb) <- bell;
@@
            r : (Bool o* Bool)
-        {emp}
+        {\top}
       = \a.\e.do apply CNOT to [a, e];

### 2020-10-26 ms-paper/ms-paper.tex
Convert booleans to bits

@@
 		\begin{minipage}[t]{0.45\columnwidth}%
-			$ \type{1} \mid \type{Bool} \mid \type{Qbit} \mid \type{U} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
+			$ \type{1} \mid \type{Bit} \mid \type{Qbit} \mid \type{U} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
 		\end{minipage}\\ \\
@@
 		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K N \mid N : A $\\
-		\textit{Intro terms} & $N, O$ & ::= & $ K \mid () \mid \lambda x.N \mid \kw{do} E \mid \kw{true} \mid \kw{false} $\\ \\
+		\textit{Intro terms} & $N, O$ & ::= & $ K \mid () \mid \lambda x.N \mid \kw{do} E \mid \kw{1} \mid \kw{0} $\\ \\
 		\textit{Q. Commands} & $c$ & ::= &
@@
 
-\paragraph{Types} We include primitive types for unit, booleans, qubits and unitary operations, and a type \type{Pure} for representing pure quantum state vectors (complex valued vectors in Hilbert space); these correspond to similar types in the QIO work.
+\paragraph{Types} We include primitive types for unit, bits, qubits and unitary operations, and a type \type{Pure} for representing pure quantum state vectors (complex valued vectors in Hilbert space); these correspond to similar types in the QIO work.
 There are also type constructors for pairs, dependent functions and the Hoare type from HTT.
@@
 These are the quantum-specific effectful commands inspired by QIO for initialization, unitary application and measurement of qubits.
-\texttt{init} $N$ initializes a given variable based on the value of the expression $N$ (typically 0 or 1 if N is a boolean).
+\texttt{init} $N$ initializes a given variable based on the value of the expression $N$ (typically 0 or 1 if N is a bit).
 \texttt{apply} $U$ \texttt{to} $X$ applies unitary operation $U$ to variables $X$.
@@
 	\inferrule
-	{\Delta \vdash N \chk \type{Bool}
+	{\Delta \vdash N \chk \type{Bit}
 		\\ \Delta, x : \type{Qbit}; P \circ (x \mapsto \kw{state}(N)) \vdash E \syn y:B . Q}
@@
 \end{mathpar}
-where x is fresh and \texttt{state} is a function that translates a classical representation of quantum state (such as boolean here) to the equivalent quantum state vector.
+where x is fresh and \texttt{state} is a function that translates a classical representation of quantum state (such as boolean here) to the equivalent quantum state vector. \todo{TODO}
 
-Initialization can only be performed for a term $N$ that can be type checked as a \type{Bool}.
+Initialization can only be performed for a term $N$ that can be type checked as a \type{Bit}.
 Then, we look at the rest of the computation, $E$, in a context that adds $x$ of type \type{Qbit} under a precondition that extends the previous precondition, $P$, with the strongest postcondition for initialization, that is, the newly bound variable $x$ points to the representation of a new qubit state; this should synthesize the strongest postcondition (with respect to the expanded precondition in the context), $Q$, for $E$.
@@
 		\\ \Delta; \Psi; P \prove (N \hookrightarrow -)
-		\\ \Delta, x:\type{Bool}; P \circ ((N \mapsto -) \multimap \top) \vdash E \syn y:B . Q}
-	{\Delta; P \vdash x \Leftarrow \kw{measure} N; E \syn y:B. (\exists x:\type{Bool}.Q)}
+		\\ \Delta, x:\type{Bit}; P \circ ((N \mapsto -) \multimap \top) \vdash E \syn y:B . Q}
+	{\Delta; P \vdash x \Leftarrow \kw{measure} N; E \syn y:B. (\exists x:\type{Bit}.Q)}
 \end{mathpar}
@@
 \begin{lstlisting}[language=QHaskell]
-hqw : {\top} r : Bool {\top /\ Id(r, false)}
-    = do q <= init false;
+hqw : {\top} r : Bit {\top /\ r=0}
+    = do q <= init 0;
          return measure q
@@
 
-In this trivial program where we initialize a new qubit with \texttt{false} and then immediately measure it, we assert that the result is equal to \texttt{false}.
+In this trivial program where we initialize a new qubit with 0 and then immediately measure it, we assert that the result is equal to 0.
 The complete specification also implicitly says that the existing quantum state is not affected in any way as both the pre and postcondition include the $\top$ assertion.
@@
 \begin{lstlisting}[language=QHaskell]
-hqw : {\top} r : Bool {\top /\ Id(r, false)}
+hqw : {\top} r : Bit {\top /\ r=0}
 -- P0: \top
-    = do q <= init false;
+    = do q <= init 0;
 -- P1: P0 \o (q |-> |0\>)
@@
 \begin{lstlisting}[language=QHaskell]
-rnd : {\top} r : Bool {\top}
-    = do q <= init false;
+rnd : {\top} r : Bit {\top}
+    = do q <= init 0;
          apply H to q;
@@
 
-In the case of quantum coin toss, the result can be in either of its two boolean values \texttt{false} or \texttt{true}, hence, we need not specify any special postcondition as type checking the result will be sufficient for correctness.
+In the case of quantum coin toss, the result can be in either 0 or 1, hence, we need not specify any special postcondition as type checking the result will be sufficient for correctness.
 
@@
 \begin{lstlisting}[language=QHaskell]
-rnd : {\top} r : Bool {\top}
+rnd : {\top} r : Bit {\top}
 -- P0: \top
-    = do q <= init false;
+    = do q <= init 0;
 -- P1: P0 \o (q |-> |0\>)
@@
 \leavevmode
-\todo{TODO: Should the type be \type{Bool} $\otimes$ \type{Bool}?}
+\todo{TODO: Should the type be \type{Bit} $\otimes$ \type{Bit}?}
 \todo{TODO: syntax for applying multi-qubit gate}
@@
 \begin{lstlisting}[language=QHaskell]
-testBell : {\top} (a, b) : (Bool, Bool) {\top /\ Id(a, b)}  -- a,b \in {true, false}
-         = do qa <= init false;
+testBell : {\top} (a, b) : (Bit, Bit) {\top /\ Id(a, b)}  -- a,b \in {0, 1}
+         = do qa <= init 0;
               apply H to qa;
-              qb <= init false;
+              qb <= init 0;
               apply CNOT to [qa, qb];
@@
 
-Here, we are asserting that the returned booleans $a$ and $b$ hold the same value, which is what we expect from the first Bell state prepared in this program.
+Here, we are asserting that the returned bits $a$ and $b$ hold the same value, which is what we expect from the first Bell state prepared in this program. \todo{TODO}
 
@@
 \begin{lstlisting}[language=QHaskell]
-testBell : {\top} (a, b) : (Bool, Bool) {\top /\ Id(a, b)}
+testBell : {\top} (a, b) : (Bit, Bit) {\top /\ Id(a, b)}
               -- P0: \top
-         = do qa <= init false;
+         = do qa <= init 0;
               -- P1: P0 \o (qa |-> |0\>)
@@
               -- P2: P1 \o ((qa |-> |0\>) -o (qa |-> |+\>))
-              qb <= init false;
+              qb <= init 0;
               -- P3: P2 \o (qb |-> |0\>)
@@
 					{\inferrule*[right=return]
-						{q : \type{Qbit}, r : \type{Bool} \vdash r \chk \type{Bool}}
-						{q : \type{Qbit}, r : \type{Bool}; P_1 \circ ((q \mapsto -) \multimap \top) \vdash \kw{return} r \syn r : \type{Bool} . R}
+						{q : \type{Qbit}, r : \type{Bit} \vdash r \chk \type{Bit}}
+						{q : \type{Qbit}, r : \type{Bit}; P_1 \circ ((q \mapsto -) \multimap \top) \vdash \kw{return} r \syn r : \type{Bit} . R}
 						\\ q : \type{Qbit} \vdash q \chk \type{Qbit}
@@
 					}
-					{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{measure} q; E_3 \syn r : \type{Bool} . (\exists r : \type{Bool}.R)}
-					\\ \cdot \vdash \kw{false} \chk \type{Bool}
+					{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{measure} q; E_3 \syn r : \type{Bit} . (\exists r : \type{Bit}.R)}
+					\\ \cdot \vdash \kw{0} \chk \type{Bit}
 				}
 				{
-					\cdot; P_0 \vdash q \Leftarrow \kw{init false}; E_{23} \syn r : \type{Bool} . (\exists q : \type{Qbit}.R)
+					\cdot; P_0 \vdash q \Leftarrow \kw{init 0}; E_{23} \syn r : \type{Bit} . (\exists q : \type{Qbit}.R)
 				}
-				\\ r : \type{Bool}; i,m ; R \prove \top \multimap \top \wedge Id(r, \kw{false})
+				\\ r : \type{Bit}; i,m ; R \prove \top \multimap \top \wedge Id(r, \kw{0})
 			}
 			{
-				\cdot; \lambda i . \lambda m . i = m \wedge (\top * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bool} . (\top \multimap \top \wedge Id(r, \kw{false}))
+				\cdot; \lambda i . \lambda m . i = m \wedge (\top * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bit} . (\top \multimap \top \wedge Id(r, \kw{0}))
 			}
 		}
-		{\cdot \vdash \kw{do} E \chk \{\top\} r : \type{Bool} \{\top \wedge \kw{Id}(r, \kw{false})\}}
+		{\cdot \vdash \kw{do} E \chk \{\top\} r : \type{Bit} \{\top \wedge \kw{Id}(r, \kw{0})\}}
 	\end{mathpar}
@@
 qplus : {\top} r : Qbit {Id(r, |+\>)}
-      = do q <= init false;
+      = do q <= init 0;
            apply H to q;
@@
 qminus : {\top} r : Qbit {Id(r, |-\>)}
-       = do q <= init true;
+       = do q <= init 1;
             apply H to q;
@@
         {Id(a, b) /\ a \in {|0\>, |1\>}}
-      = \a.do b <= init false;
+      = \a.do b <= init 0;
               apply CNOT to [a, b];
@@
 \begin{lstlisting}[language=QHaskell]
-testBell : {\top} (a, b) : (Bool o* Bool) {\top /\ Id(a, b)}
+testBell : {\top} (a, b) : (Bit o* Bit) {\top /\ Id(a, b)}
          = do (qa, qb) <- bell;
@@
         {(Id(a, -) /\ entangled(e)}
-           r : (Bool o* Bool)
+           r : (Bit o* Bit)
         {\top}
@@
 \begin{lstlisting}[language=QHaskell]
-bob : \Pi m1 m2 : Bool. \Pi e : Qbit
+bob : \Pi m1 m2 : Bit. \Pi e : Qbit
       {entangled(e)}

### 2020-10-26 ms-paper/ms-paper.tex
Get rid of state function
Revise share example

@@
 	{\\bot}{{$\bot$}}1
+	{=q}{{$=_q$}}2
+	{alpha}{{$\alpha$}}1 {beta}{{$\beta$}}1
 }
@@
 	{\Delta \vdash N \chk \type{Bit}
-		\\ \Delta, x : \type{Qbit}; P \circ (x \mapsto \kw{state}(N)) \vdash E \syn y:B . Q}
+		\\ \Delta, x : \type{Qbit}; P \circ (x \mapsto \ket{N}) \vdash E \syn y:B . Q}
 	{\Delta; P \vdash x \Leftarrow \kw{init} N; E \syn y:B. (\exists x:\type{Qbit}.Q)}
 \end{mathpar}
-where x is fresh and \texttt{state} is a function that translates a classical representation of quantum state (such as boolean here) to the equivalent quantum state vector. \todo{TODO}
+where x is fresh.
 
@@
 \subsubsection{Hello Quantum World}
+ \todo{TODO}
 \leavevmode
@@
 
-Verification steps follow:
+Verification steps follow:  \todo{TODO}
 
@@
 
-Verification steps follow:
+Verification steps follow: \todo{TODO}
 
@@
 share : \Pi a : Qbit.
-        {a \in {|+\>, |-\>}}   -- a \in {b, c} is short for Id(a,b) \/ Id(a,c)
-           b : Qbit
-        {Id(a, b) /\ a \in {|0\>, |1\>}}
+        {a =q (alpha |0\> + beta |1\>)}
+           (a, b) : (Qbit o* Qbit)
+        {(a, b) =q (alpha |00\> + beta |11\>)}
       = \a.do b <= init 0;
-              apply CNOT to [a, b];
-              return b
+              (a, b) <= apply CNot to [a, b];
+              return (a, b)
 \end{lstlisting}

### 2020-10-26 ms-paper/ms-paper.tex
Revise examples in appendix B

@@
 \begin{lstlisting}[language=QHaskell]
-qplus : {\top} r : Qbit {Id(r, |+\>)}
+qplus : {\top} r : Qbit {r =q |+\>}
       = do q <= init 0;
-           apply H to q;
+           q <= apply H to q;
            return q
 
-qminus : {\top} r : Qbit {Id(r, |-\>)}
+qminus : {\top} r : Qbit {r =q |-\>}
        = do q <= init 1;
-            apply H to q;
+            q <= apply H to q;
             return q
@@
 \begin{lstlisting}[language=QHaskell]
-bell : {\top} (a, b) : (Qbit o* Qbit) {Id(a, b) /\ a \in {|0\>, |1\>}}
+bell00 : {\top} (a, b) : (Qbit o* Qbit) {(a,b) =q |\Phi+\>}
      = do qa <- qplus;
-          qb <- share qa;
+          (qa, qb) <- share qa;
           return (qa, qb)
@@
 \subsection{Testing modular Bell pair}
+\todo{TODO: how is classical equality ensured?}
 \begin{lstlisting}[language=QHaskell]
-testBell : {\top} (a, b) : (Bit o* Bit) {\top /\ Id(a, b)}
-         = do (qa, qb) <- bell;
-              return (measure qa, measure qb)
+testBell : {\top} (a, b) : (Bit o* Bit) {\top /\ a = b}
+         = do (qa, qb) <- bell00;
+              a <= measure qa;
+              b <= measure qb;
+              return (a, b)
 \end{lstlisting}

### 2020-10-26 ms-paper/ms-paper.tex
Refine teleport example in appendix C

@@
 	language     = Haskell,
-	morekeywords = {Qbit, U, C, Pure},
+	morekeywords = {Qbit, U, C, Bit},
 }
@@
 \label{app:teleport}
-\todo{TODO: Refine assertions}
 \subsection{Alice's circuit}
 \begin{lstlisting}[language=QHaskell]
-alice : \Pi a : Qbit. \Pi e : Qbit
-        {(Id(a, -) /\ entangled(e)}
-           r : (Bit o* Bit)
-        {\top}
-      = \a.\e.do apply CNOT to [a, e];
-                 apply H to a;
-                 return (measure a, measure e)
+alice : \Pi q : Qbit. \Pi a : Qbit
+        {q =q (alpha |0\> + beta |1\>) /\ (a, e) =q |\Phi+\>}    -- ie, a is entangled with ghost var e
+           (m1, m2) : (Bit o* Bit)
+        {  (m1, m2) = 00 /\ e =q (alpha |0\> + beta |1\>)
+         \/ (m1, m2) = 01 /\ e =q (alpha |1\> + beta |0\>)
+         \/ (m1, m2) = 10 /\ e =q (alpha |0\> - beta |1\>)
+         \/ (m1, m2) = 11 /\ e =q (alpha |1\> - beta |0\>)}
+      = \q.\a.do (q, a) <= apply CNOT to [q, a];
+                 q <= apply H to q;
+                 rq <= measure q;
+                 ra <= measure a;
+                 return (rq, ra)
 \end{lstlisting}
@@
 \subsection{Bob's circuit}
+\todo{TODO provide correct assertions using ghost e}
 \begin{lstlisting}[language=QHaskell]
-bob : \Pi m1 m2 : Bit. \Pi e : Qbit
-      {entangled(e)}
+bob : \Pi m1 m2 : Bit. \Pi b : Qbit
+      {(b, e) =q |\Phi+\>}
          r : Qbit
-      {(Id(r, -)}
-      = \m1.\m2.\e.do if m1 then (apply Z to e) else ();
-                      if m2 then (apply X to e) else ();
-                      return e
+      {\top}
+      = \m1.\m2.\b.do (m1, b) <= apply C_Z to [m1, b];
+                      (m2, b) <= apply C_X to [m2, b];
+                      return b
 \end{lstlisting}
@@
 \begin{lstlisting}[language=QHaskell]
-teleport : \Pi q : Qbit. x : Pure.
-           {Id(q, x)}
-              r : Qbit
-           {Id(r, x)}
-         = \q.do (a, b) <- bell;
+teleport : \Pi q : Qbit.
+           {q =q (alpha |0\> + beta |1\>)}
+              b : Qbit
+           {r =q (alpha |0\> + beta |1\>)}
+         = \q.do (a, b) <- bell00;
                  (m1, m2) <- alice q a;
-                 tq <- bob m1 m2 b;
-                 return tq
+                 b <- bob m1 m2 b;
+                 return b
 \end{lstlisting}

### 2020-10-26 ms-paper/ms-paper.tex
Revise remaining examples, update some text for consistency

@@
 		\begin{minipage}[t]{0.45\columnwidth}%
-			$ \kw{init} N \mid \kw{apply} U \kw{to} N \mid \kw{measure} N \mid \kw{if} y \kw{then} E \kw{else} F$
+			$ \kw{init} N \mid \kw{apply} U \kw{to} N \mid \kw{measure} N$
 		\end{minipage}\\
 		\textit{Computations} & $E, F$ & ::= & $ \kw{return} N \mid x \leftarrow K; E \mid x \Leftarrow c; E \mid x =_A N; E $\\ \\
-		\textit{Variable context} & $\Delta$ & ::= & $ \cdot \mid x : A, \Delta $ \\
-		\textit{Q. Heap context} & $\Psi$ & ::= & $ \cdot \mid h, \Psi$ \\
-		\textit{Assertion context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
+%		\textit{Variable context} & $\Delta$ & ::= & $ \cdot \mid x : A, \Delta $ \\
+%		\textit{Q. Heap context} & $\Psi$ & ::= & $ \cdot \mid h, \Psi$ \\
+%		\textit{Assertion context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
 	\end{tabular}
@@
 
-\paragraph{Types} We include primitive types for unit, bits, qubits and unitary operations, and a type \type{Pure} for representing pure quantum state vectors (complex valued vectors in Hilbert space); these correspond to similar types in the QIO work.
+\paragraph{Types} We include primitive types for unit, bits, qubits and unitary operations; these correspond to similar types in the QIO work.
 There are also type constructors for pairs, dependent functions and the Hoare type from HTT.
@@
 Here are some examples.
-The postcondition of the program \texttt{qplus} can be written as $r =_q \ket{+}$.
-The postcondition for the program \texttt{bell} as $ab =_q \frac{1}{\sqrt{2}}\ket{00}+\frac{1}{\sqrt{2}}\ket{11}$ and the precondition for the program \texttt{share} as $a =_q \ket{+} \vee a =_q \ket{-}$.
+The postcondition of the program \texttt{qplus} below can be written as $r =_q \ket{+}$.
+The postcondition for the program \texttt{bell00} as $(a, b) =_q \frac{1}{\sqrt{2}}\ket{00}+\frac{1}{\sqrt{2}}\ket{11}$ or equivalently as $(a, b) =_q \ket{\Phi^+}$..
 
@@
 \texttt{measure} $x$ measures the variable $x$ in the computational basis.
-\texttt{if y then E else F} executes computation $E$ if measuring variable $y$ returns 1 else executes $F$.
+%\texttt{if y then E else F} executes computation $E$ if measuring variable $y$ returns 1 else executes $F$.
 
@@
 
-Verification steps follow:  \todo{TODO}
+Verification steps follow:  \todo{TODO how to ensure classical equality?}
 
@@
 \begin{lstlisting}[language=QHaskell]
-hqw : {\top} r : Bit {\top /\ r=0}
+hqw : {\top} r : Bit {\top /\ r = 0}
 -- P0: \top
@@
     = do q <= init 0;
-         apply H to q;
+         q <= apply H to q;
          return measure q
@@
 -- P1: P0 \o (q |-> |0\>)
-         apply H to q;
+         q <= apply H to q;
 -- P2: P1 \o ((q |-> |0\>) -o (q |-> |+\>))
@@
 \leavevmode
-\todo{TODO: Should the type be \type{Bit} $\otimes$ \type{Bit}?}
 \todo{TODO: syntax for applying multi-qubit gate}
@@
 \begin{lstlisting}[language=QHaskell]
-testBell : {\top} (a, b) : (Bit, Bit) {\top /\ Id(a, b)}  -- a,b \in {0, 1}
+testBell00 : {\top} (a, b) : (Bit o* Bit) {\top /\ a = b}
          = do qa <= init 0;
-              apply H to qa;
+              qa <= apply H to qa;
               qb <= init 0;
-              apply CNOT to [qa, qb];
+              (qa, qb) <= apply CNOT to [qa, qb];
               return (measure qa, measure qb)
@@
 
-Here, we are asserting that the returned bits $a$ and $b$ hold the same value, which is what we expect from the first Bell state prepared in this program. \todo{TODO}
+Here, we are asserting that the returned bits $a$ and $b$ hold the same value, which is what we expect from the first Bell state prepared in this program.
 
@@
 We would like to describe how QHTT lets us verify the correctness of the above program based on the specification provided as the type of the function.
-For illustration, we show the assertions annotated as comments as we step through the \texttt{testBell} program below.
+For illustration, we show the assertions annotated as comments as we step through the \texttt{testBell00} program below.
 We have elided specific details about unitary application as they are still being worked out.
@@
 \begin{lstlisting}[language=QHaskell]
-testBell : {\top} (a, b) : (Bit, Bit) {\top /\ Id(a, b)}
+testBell00 : {\top} (a, b) : (Bit o* Bit) {\top /\ a = b}
               -- P0: \top
@@
               -- P1: P0 \o (qa |-> |0\>)
-              apply H to qa;
+              qa <= apply H to qa;
               -- P2: P1 \o ((qa |-> |0\>) -o (qa |-> |+\>))
@@
               -- P3: P2 \o (qb |-> |0\>)
-              apply CNOT to [qa, qb];
+              (qa, qb) <= apply CNOT to [qa, qb];
               -- P4: P3 \o ((qa |-> |+\>, qb |-> |0\>) -o (qa, qb) |-> |\Phi+\>)
@@
 
-\subsection{Quantum Assertion Language}
-We need to be explicit about what we mean when we say that a certain state satisifes a certain assertion (predicate) in the quantum setting.
-In this model, the predicates are subspaces of the Hilbert space corresponding to a quantum state.
-Here are some examples:
-\begin{itemize}
-	\item $\top$ represents the complete Hilbert space that is always satisfied by any state.
-	\item $\bot$ represents nothing, it is never satisfied.
-	\item $A \wedge B$ represent that both A and B hold (conjunction), it is equivalent to the intersection of the subspaces.
-	\item $A \vee B$ similarly represents disjunction and is modeled as the sum (linear combination) of the subspaces A and B.
-	\item If an operator M is applied to a state satisfying a predicate A, then the new predicate can be written as $M \cdot A$.
-	\item If we want to write that a variable w, or a list of variables W is in a specific state, we can model that predicate as the variable(s) being in the span of the state and use the notation, $W =_q \psi$.
-\end{itemize}
-
-Here are some examples.
-The postcondition of the program \texttt{qplus} above can be written as $r =_q \ket{+}$.
-The postcondition for the program \texttt{bell} as $ab =_q \frac{1}{\sqrt{2}}\ket{00}+\frac{1}{\sqrt{2}}\ket{11}$ and the precondition for the program \texttt{share} as $a =_q \ket{+} \vee a =_q \ket{-}$.
-
-What is the relation between the program variables and the subspaces corresponding to the predicates that we are talking about?
+%What is the relation between the program variables and the subspaces corresponding to the predicates that we are talking about?
 We need to define the notion of free variables for a predicate $A$.
@@
 
-There are more predicates that will involve ghosts, described in the section after the next.
-
 \subsection{Improvements to the quantum commands}
 
-Added if then else, skip, sequencing, and a simpler form of apply; and the syntactic sugar for quantum and classical assignments in the ott document.
+Added if then else and a simpler form of apply; and the syntactic sugar for quantum and classical assignments in the ott document.
 

### 2020-10-28 ms-paper/ms-paper.tex
simplify ott build

@@
 alice : \Pi q : Qbit. \Pi a : Qbit
-        {q =q (alpha |0\> + beta |1\>) /\ (a, e) =q |\Phi+\>}    -- ie, a is entangled with ghost var e
+        {q =q (alpha |0\> + beta |1\>) /\ (a, e) =q |\Phi+\>}    -- a is entangled with ghost var e
            (m1, m2) : (Bit o* Bit)

### 2020-10-28 ott/build.sh

@@ -1,3 +1,3 @@
 #!/bin/bash
 
-ott -i qhtt.ott -o qhtt.tex && make qhtt
+ott -show_sort true -show_defns true -i qhtt.ott -o qhtt.tex

### 2020-10-29 ms-paper/ms-paper.tex
Use the term assertion only for runtime assertions

@@
 
-As programming languages researchers, we think it will be better to encode such assertions into a static type system both for formal verification and to aid the programmers in writing correct programs from the start.
-Inspired by the use of Hoare triples in the verification of imperative programs and building on the idea of Hoare Type Theory (HTT) for classical programming languages~\cite{nanevski_hoare_2008}, we extend a quantum effect monad inspired by the Quantum IO Monad interface~\cite{qio} and propose Quantum Hoare Type Theory aimed at enabling both sound static type checking and formal verification of quantum programs.
+As programming languages researchers, we think it will be better to encode such assertions (propositions) into a static type system both for formal verification and to aid the programmers in writing correct programs from the start.\footnote{In this work, we will use the terms \textit{proposition} and \textit{predicate} interchangeably to signify their static nature as opposed to the runtime nature of \textit{assertions}.} Inspired by the use of Hoare triples in the verification of imperative programs and building on the idea of Hoare Type Theory (HTT) for classical programming languages~\cite{nanevski_hoare_2008}, we extend a quantum effect monad inspired by the Quantum IO Monad interface~\cite{qio} and propose Quantum Hoare Type Theory aimed at enabling both sound static type checking and formal verification of quantum programs.
 
@@
 		\textit{Types} & $A, B, C$ & ::= &
-		\begin{minipage}[t]{0.45\columnwidth}%
-			$ \type{1} \mid \type{Bit} \mid \type{Qbit} \mid \type{U} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
+		\begin{minipage}[t]{0.5\columnwidth}%
+			$ \type{1} \mid \type{Qbit} \mid \type{U} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
 		\end{minipage}\\ \\
-		\textit{Assertions} & $P, Q, R$ & ::= &
-		\begin{minipage}[t]{0.45\columnwidth}%
-			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid M \cdot P \mid X \in_q S \mid X =_q \psi \mid \kw{uniform}(X) \mid \kw{separable}(X) \mid \kw{classical}(X) $
+		\textit{Propositions} & $P, Q, R$ & ::= &
+		\begin{minipage}[t]{0.5\columnwidth}%
+			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid U \cdot P \mid X \in_q S \mid X =_q \psi \mid \kw{uniform}(X) \mid \kw{separable}(X) \mid \kw{classical}(X) $
 		\end{minipage}\\ \\
@@
 		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K N \mid N : A $\\
-		\textit{Intro terms} & $N, O$ & ::= & $ K \mid () \mid \lambda x.N \mid \kw{do} E \mid \kw{1} \mid \kw{0} $\\ \\
+		\textit{Intro terms} & $N, O$ & ::= & $ K \mid () \mid \lambda x.N \mid \kw{do} E $\\ \\
 		\textit{Q. Commands} & $c$ & ::= &
-		\begin{minipage}[t]{0.45\columnwidth}%
-			$ \kw{init} N \mid \kw{apply} U \kw{to} N \mid \kw{measure} N$
+		\begin{minipage}[t]{0.5\columnwidth}%
+			$ \kw{init0} \mid \kw{init1} \mid \kw{apply} U \kw{to} N \mid \kw{measure} N$
 		\end{minipage}\\
@@
 %		\textit{Q. Heap context} & $\Psi$ & ::= & $ \cdot \mid h, \Psi$ \\
-%		\textit{Assertion context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
+%		\textit{Proposition context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
+		\textit{Unitaries} & $U$ & ::= & $ I \mid X \mid Y \mid Z \mid H \mid CX \mid CZ \mid S \mid T$
 	\end{tabular}
@@
 
-\paragraph{Types} We include primitive types for unit, bits, qubits and unitary operations; these correspond to similar types in the QIO work.
-There are also type constructors for pairs, dependent functions and the Hoare type from HTT.
+\paragraph{Types} We include primitive types for unit, qubits, and unitary operations. There are also type constructors for pairs, dependent functions and the Hoare type from HTT.
 
-\paragraph{Assertions}
-The assertion language is inspired by \citet{unruh2019} that follows subspace semantics.
-We need to be explicit about what we mean when we say that a certain state satisifes a certain assertion (predicate) in the quantum setting.
-In this model, the predicates are subspaces of the Hilbert space corresponding to a quantum state.
+\paragraph{Propositions}
+The predicate language is inspired by \citet{unruh2019} that follows subspace semantics.
+We need to be explicit about what we mean when we say that a certain state satisifes a certain predicate in the quantum setting.
+In this model, the propositions are subspaces of the Hilbert space corresponding to a quantum state.
 Here are some examples:
@@
 	\item $P \vee Q$ similarly represents disjunction and is modeled as the sum (linear combination) of the subspaces P and Q.
-	\item If an operator M is applied to a state satisfying a predicate P, then the new predicate can be written as $M \cdot P$.
-	\item If we want to write that a variable w, or a list of variables W is in a specific state, we can model that predicate as the variable(s) being in the span of the state and use the notation, $W =_q \psi$.
+	\item If an operator M is applied to a state satisfying a proposition P, then the new proposition can be written as $U \cdot P$.
+	\item If we want to write that a variable w, or a list of variables W is in a specific state, we can model that proposition as the variable(s) being in the span of the state and use the notation, $W =_q \psi$.
 \end{itemize}
@@
 
-There are other predicates that involve ghosts. \todo{TODO}
+There are other propositions that involve ghosts. \todo{TODO}
 
@@
 
-Finally, the judgement $\Delta; \Psi; \Gamma_1 \prove \Gamma_2$ encodes the sequent calculus for the assertion logic.
-Recall that $\Delta$, $\Psi$, and $\Gamma$ denote the variable context, the quantum heap context and the assertion context, respectively.
+Finally, the judgement $\Delta; \Psi; \Gamma_1 \prove \Gamma_2$ encodes the sequent calculus for the predicate logic.
+Recall that $\Delta$, $\Psi$, and $\Gamma$ denote the variable context, the quantum heap context and the proposition context, respectively.
 
@@
 Here we show the strongest postconditions for the primitive quantum commands of QHTT encoded in their typing rules.
-Given a quantum command for initialization, unitary application or measurement, its strongest postcondition is an assertion that most precisely captures the relationship between the initial state and the modified state after the execution.
+Given a quantum command for initialization, unitary application or measurement, its strongest postcondition is a proposition that most precisely captures the relationship between the initial state and the modified state after the execution.
 These rules along with the \texttt{consq} (consequent) rule of HTT (that ensures weakening of the given strongest postcondition to an arbitrary postcondition) work together to ensure composition of verification.
@@
 An observation is that translating the programs was a very simple process except for coming up with the right specifications for the programs.
-We provide some commentary on assertions specified for these programs.
+We provide some commentary on propositions specified for these programs.
 
@@
 In this trivial program where we initialize a new qubit with 0 and then immediately measure it, we assert that the result is equal to 0.
-The complete specification also implicitly says that the existing quantum state is not affected in any way as both the pre and postcondition include the $\top$ assertion.
+The complete specification also implicitly says that the existing quantum state is not affected in any way as both the pre and postcondition include the $\top$ proposition.
 
@@
 We would like to describe how QHTT lets us verify the correctness of the above program based on the specification provided as the type of the function.
-For illustration, we show the assertions annotated as comments as we step through the \texttt{testBell00} program below.
+For illustration, we show the propositions annotated as comments as we step through the \texttt{testBell00} program below.
 We have elided specific details about unitary application as they are still being worked out.
@@
 
-In the assertion P2, we show that only the portion of the quantum state referred to by the \texttt{apply} command is affected.
+In the proposition P2, we show that only the portion of the quantum state referred to by the \texttt{apply} command is affected.
 Similarly, in P4 we see that only the relevant portion of the heap containing the control and target qubits used in the CNOT operation are affected.
 In this case, we also needed to combine the two qubit state as they are now entangled.
-The last assertion, P5, merely encodes the strongest postcondition for measurement for each of the two qubits.
+The last proposition, P5, merely encodes the strongest postcondition for measurement for each of the two qubits.
 
-Finally, the \texttt{consq} rule of HTT ensures that the assertion P5 leads to the specified postcondition in the type of the program.
+Finally, the \texttt{consq} rule of HTT ensures that the proposition P5 leads to the specified postcondition in the type of the program.
 
 We show some two more examples in the appendix for a modular version of Bell pair generation (\Cref{app:mbp}) that shows use of higher-order features of QHTT and another for teleportation protocol (\Cref{app:teleport}).
-Both of these involve specifying assertions over quantum state whose theory we still need to work out.
+Both of these involve specifying propositions over quantum state whose theory we still need to work out.
 
@@
 This results in requiring a complicated bidirectional typing system for implementation.
-If we use predicate transformers as demonstrated by the use of Dijkstra monads in F*, we can avoid these issues \cite{swamy2013verifying}.
+If we use proposition transformers as demonstrated by the use of Dijkstra monads in F*, we can avoid these issues \cite{swamy2013verifying}.
 
@@
 
-\paragraph{Quantum Assertion Logic} An obvious need is to come up with an assertion logic similar to Separation Logic~\cite{reynolds2002} for quantum computing so as to be able to reason about only the interesting portions of the quantum state while still ensuring correctness of non-local effects such as entanglement.
+\paragraph{Quantum Predicate Logic} An obvious need is to come up with a predicate logic similar to Separation Logic~\cite{reynolds2002} for quantum computing so as to be able to reason about only the interesting portions of the quantum state while still ensuring correctness of non-local effects such as entanglement.
 Various Quantum Hoare Logics that exist~\cite{ying_floydhoare_2012} currently do not support frame rules that provide Separation Logic its power.
-The closest approach is that taken by \citet{unruh2019} that allows assertions on quantum state similar to what we would like in our setting, such as whether certain qubits are in classical state, or in uniform superposition, entangled or separable.
+The closest approach is that taken by \citet{unruh2019} that allows predicates on quantum state similar to what we would like in our setting, such as whether certain qubits are in classical state, or in uniform superposition, entangled or separable.
 We hope to incorporate these ideas in this work; this may require us to use a subspace semantics for the quantum state as shown in \citet{unruh2019}.
@@
 \subsection{Bob's circuit}
-\todo{TODO provide correct assertions using ghost e}
+\todo{TODO provide correct propositions using ghost e}
 \begin{lstlisting}[language=QHaskell]
@@
 
-Robert suggests another approach that is to limit the expressivity of the system and implement the limited logic proposed in Häner, Hoefler and Troyer's ``Using Hoare logic for quantum circuit optimization'' (\url{https://arxiv.org/abs/1810.00375}), an improved version of which is going to be presented at OOPSLA in November.
+Robert suggests another approach that is to limit the expressiveness of the system and implement the limited logic proposed in Häner, Hoefler and Troyer's ``Using Hoare logic for quantum circuit optimization'' (\url{https://arxiv.org/abs/1810.00375}), an improved version of which is going to be presented at OOPSLA in November.
 This may even be possible in F* with some effort.
@@
 \subsection{Remarks}
-Now, the question is whether we can use this memory model (based on subspsace semantics), assertion logic and rules and still continue to use bidirectional typing for the type system?
+Now, the question is whether we can use this memory model (based on subspace semantics), predicate logic and rules and still continue to use bidirectional typing for the type system?
 If so, it's probably worth finishing up the rule for unitary application using this memory model.

### 2020-10-29 ms-paper/ms-paper.tex
Simplify syntax some more

@@
 		\begin{minipage}[t]{0.5\columnwidth}%
-			$ \type{1} \mid \type{Qbit} \mid \type{U} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
+			$ \type{1} \mid \type{Qbit} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
 		\end{minipage}\\ \\
@@
 		\begin{minipage}[t]{0.5\columnwidth}%
-			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid U \cdot P \mid X \in_q S \mid X =_q \psi \mid \kw{uniform}(X) \mid \kw{separable}(X) \mid \kw{classical}(X) $
+			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid U \cdot P \mid \textbf{X} =_q \psi \mid \kw{uniform}(\textbf{X}) \mid \kw{separable}(\textbf{X}) \mid \kw{classical}(\textbf{X}) $
 		\end{minipage}\\ \\
@@
 
-\paragraph{Types} We include primitive types for unit, qubits, and unitary operations. There are also type constructors for pairs, dependent functions and the Hoare type from HTT.
+\paragraph{Types} There are primitive types for unit and qubits. There are also type constructors for pairs, dependent functions and the Hoare type from HTT.
 
@@
 In this model, the propositions are subspaces of the Hilbert space corresponding to a quantum state.
+We use the metavariable \textbf{X} to refer to a list of quantum variables.
 Here are some examples:
@@
 	\item $P \vee Q$ similarly represents disjunction and is modeled as the sum (linear combination) of the subspaces P and Q.
-	\item If an operator M is applied to a state satisfying a proposition P, then the new proposition can be written as $U \cdot P$.
-	\item If we want to write that a variable w, or a list of variables W is in a specific state, we can model that proposition as the variable(s) being in the span of the state and use the notation, $W =_q \psi$.
+	\item If a unitary operator U is applied to a state satisfying a proposition P, then the new proposition can be written as $U \cdot P$.
+	\item If we want to write that a list of variables \textbf{X} is in a specific state, we can model that proposition as the variables being in the span of the state and use the notation, $\textbf{X} =_q \psi$.
 \end{itemize}
@@
 The postcondition of the program \texttt{qplus} below can be written as $r =_q \ket{+}$.
-The postcondition for the program \texttt{bell00} as $(a, b) =_q \frac{1}{\sqrt{2}}\ket{00}+\frac{1}{\sqrt{2}}\ket{11}$ or equivalently as $(a, b) =_q \ket{\Phi^+}$..
+The postcondition for the program \texttt{bell00} as $(a, b) =_q \frac{1}{\sqrt{2}}\ket{00}+\frac{1}{\sqrt{2}}\ket{11}$ or equivalently as $(a, b) =_q \ket{\Phi^+}$.
 
-There are other propositions that involve ghosts. \todo{TODO}
+There are other propositions that involve ghosts:
+\begin{itemize}
+	\item uniform(\textbf{X}) is an abbreviation for $(\textbf{X}, \textbf{e}) = \frac{1}{\sqrt{2}}(\ket{00}+\ket{11})$ where \textbf{e} is  a fresh entangled ghost variable.
+It means that X is uniformly distributed.
+	\item separable(\textbf{X}) is an abbreviation for $\textbf{X} \equiv_{q} \textbf{u}$ where \textbf{u} is a fresh unentangled ghost variable.
+	\item class(\textbf{X}) is an abbreviation for $\textbf{X} \equiv_{cl} \textbf{u}$ where \textbf{u} is a fresh unentangled ghost variable.
+The classical equality symbol means that the variables on both side will measure to be the exact same values.
+That is there is no superposition involved.
+\end{itemize}
 

### 2020-10-30 ms-paper/ms-paper.tex
Reintroduce booleans, measure -> meas

@@
 	language     = Haskell,
-	morekeywords = {Qbit, U, C, Bit},
+	morekeywords = {Qbit, U, C, meas, apply, to},
 }
@@
 
-Here we discuss the relevant bits from the QIO interface.
+Here we discuss the relevant concepts from the QIO interface.
 QIO provides a type for referring to qubits, \texttt{Qbit}; a type for unitary operations, \texttt{U}; and, the \texttt{QIO} type operator which is a monad indexed by the result of the quantum computation.
@@
 
-Complete grammar written using Ott tool at \url{https://github.com/k4rtik/QHT/blob/master/ott/qhtt.ott}.
+%Complete grammar written using Ott tool at \url{https://github.com/k4rtik/QHT/blob/master/ott/qhtt.ott}.
 
@@
 		\textit{Types} & $A, B, C$ & ::= &
-		\begin{minipage}[t]{0.5\columnwidth}%
-			$ \type{1} \mid \type{Qbit} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
+		\begin{minipage}[t]{0.45\columnwidth}%
+			$ \type{1} \mid \type{Bool} \mid \type{Qbit} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
 		\end{minipage}\\ \\
@@
 		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K N \mid N : A $\\
-		\textit{Intro terms} & $N, O$ & ::= & $ K \mid () \mid \lambda x.N \mid \kw{do} E $\\ \\
+		\textit{Intro terms} & $N, O$ & ::= & $ K \mid () \mid \lambda x.N \mid \kw{do} E \mid \kw{true} \mid \kw{false} $\\ \\
 		\textit{Q. Commands} & $c$ & ::= &
 		\begin{minipage}[t]{0.5\columnwidth}%
-			$ \kw{init0} \mid \kw{init1} \mid \kw{apply} U \kw{to} N \mid \kw{measure} N$
+			$ \kw{\textbf{init}} N \mid \kw{\textbf{apply}} U \kw{\textbf{to}} N \mid \kw{\textbf{meas}} N$
 		\end{minipage}\\
-		\textit{Computations} & $E, F$ & ::= & $ \kw{return} N \mid x \leftarrow K; E \mid x \Leftarrow c; E \mid x =_A N; E $\\ \\
+		\textit{Computations} & $E, F$ & ::= & $ \kw{return} N \mid x \leftarrow K; E \mid x \Leftarrow c; E \mid x = N; E $\\ \\
 %		\textit{Variable context} & $\Delta$ & ::= & $ \cdot \mid x : A, \Delta $ \\
@@
 
-\paragraph{Types} There are primitive types for unit and qubits. There are also type constructors for pairs, dependent functions and the Hoare type from HTT.
+\paragraph{Types} There are primitive types for unit, booleans, and qubits. There are also type constructors for pairs, dependent functions and the Hoare type from HTT.
 
@@
 	\item class(\textbf{X}) is an abbreviation for $\textbf{X} \equiv_{cl} \textbf{u}$ where \textbf{u} is a fresh unentangled ghost variable.
-The classical equality symbol means that the variables on both side will measure to be the exact same values.
-That is there is no superposition involved.
+The classical equality symbol means that the variables on both side will measure to be the exact same value.
+That is, there is no superposition involved.
 \end{itemize}
+
+In addition to these quantum-specific propositions, the system also supports first-order logic (not shown) for reasoning about the pure fragment of the language.
 
@@
 These are the quantum-specific effectful commands inspired by QIO for initialization, unitary application and measurement of qubits.
-\texttt{init} $N$ initializes a given variable based on the value of the expression $N$ (typically 0 or 1 if N is a bit).
+\texttt{init} $N$ initializes a given variable based on the value of the expression $N$ (typically \ket{0} or \ket{1} if N is a boolean).
 \texttt{apply} $U$ \texttt{to} $X$ applies unitary operation $U$ to variables $X$.
-\texttt{measure} $x$ measures the variable $x$ in the computational basis.
+\texttt{meas} $x$ measures the variable $x$ in the computational basis.
 %\texttt{if y then E else F} executes computation $E$ if measuring variable $y$ returns 1 else executes $F$.
@@
 \paragraph{Computations} The computations are the usual monadic return and three sequencing operations that bind their result to a variable similar to HTT.
-The first, $x \leftarrow K$, executes a suspended computation (such as applying a function that includes a \texttt{do} $E$ in its body); the second, $x \Leftarrow c$ executes a primitive command $c$; and the last, $x =_A N$, is just syntactic sugar for let-binding.
+The first, $x \leftarrow K$, executes a suspended computation (such as applying a function that includes a \texttt{do} $E$ in its body); the second, $x \Leftarrow c$ executes a primitive command $c$; and the last, $x = N$, is just syntactic sugar for let-binding.
 
@@
 	\inferrule
-	{\Delta \vdash N \chk \type{Bit}
-		\\ \Delta, x : \type{Qbit}; P \circ (x \mapsto \ket{N}) \vdash E \syn y:B . Q}
+	{\Delta \vdash N \chk \type{Bool}
+		\\ \Delta, x : \type{Qbit}; P \circ (x =_q \mathit{bool2ket}(N)) \vdash E \syn y:B . Q}
 	{\Delta; P \vdash x \Leftarrow \kw{init} N; E \syn y:B. (\exists x:\type{Qbit}.Q)}
@@
 
-Initialization can only be performed for a term $N$ that can be type checked as a \type{Bit}.
-Then, we look at the rest of the computation, $E$, in a context that adds $x$ of type \type{Qbit} under a precondition that extends the previous precondition, $P$, with the strongest postcondition for initialization, that is, the newly bound variable $x$ points to the representation of a new qubit state; this should synthesize the strongest postcondition (with respect to the expanded precondition in the context), $Q$, for $E$.
+Initialization can only be performed for a term $N$ that can be type checked as a \type{Bool}.
+Then, we look at the rest of the computation, $E$, in a context that adds $x$ of type \type{Qbit} under a precondition that extends the previous precondition, $P$, with the strongest postcondition for initialization, that is, the newly bound variable $x$ is equal to the \ket{0} or \ket{1}; this should synthesize the strongest postcondition (with respect to the expanded precondition in the context), $Q$, for $E$.
 
@@
 \paragraph{Measurement}
-$x \Leftarrow \kw{measure} N; E$
+$x \Leftarrow \kw{meas} N; E$
 
@@
 	{\Delta \vdash N \chk \type{Qbit}
-		\\ \Delta; \Psi; P \prove (N \hookrightarrow -)
-		\\ \Delta, x:\type{Bit}; P \circ ((N \mapsto -) \multimap \top) \vdash E \syn y:B . Q}
-	{\Delta; P \vdash x \Leftarrow \kw{measure} N; E \syn y:B. (\exists x:\type{Bit}.Q)}
+		\\ \Delta; \Psi; P \prove (N =_q -)
+		\\ \Delta, x:\type{Bool}; P \circ ((N =_q -) \multimap \top) \vdash E \syn y:B . Q}
+	{\Delta; P \vdash x \Leftarrow \kw{meas} N; E \syn y:B. (\exists x:\type{Bool}.Q)}
 \end{mathpar}
@@
 The strongest postcondition here is that the fragment of heap that only refers to location N becomes empty.
+\todo{TODO: need to introduce ghost variable for measurement}
 
@@
 
-In this trivial program where we initialize a new qubit with 0 and then immediately measure it, we assert that the result is equal to 0.
+In this trivial program where we initialize a new qubit with false and then immediately measure it, we assert that the result is false.
 The complete specification also implicitly says that the existing quantum state is not affected in any way as both the pre and postcondition include the $\top$ proposition.
 
-Verification steps follow:  \todo{TODO how to ensure classical equality?}
+Verification steps follow:
 
@@
 					{\inferrule*[right=return]
-						{q : \type{Qbit}, r : \type{Bit} \vdash r \chk \type{Bit}}
-						{q : \type{Qbit}, r : \type{Bit}; P_1 \circ ((q \mapsto -) \multimap \top) \vdash \kw{return} r \syn r : \type{Bit} . R}
+						{q : \type{Qbit}, r : \type{Bool} \vdash r \chk \type{Bool}}
+						{q : \type{Qbit}, r : \type{Bool}; P_1 \circ ((q \mapsto -) \multimap \top) \vdash \kw{return} r \syn r : \type{Bool} . R}
 						\\ q : \type{Qbit} \vdash q \chk \type{Qbit}
@@
 					}
-					{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{measure} q; E_3 \syn r : \type{Bit} . (\exists r : \type{Bit}.R)}
-					\\ \cdot \vdash \kw{0} \chk \type{Bit}
+					{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{meas} q; E_3 \syn r : \type{Bool} . (\exists r : \type{Bool}.R)}
+					\\ \cdot \vdash \kw{0} \chk \type{Bool}
 				}
 				{
-					\cdot; P_0 \vdash q \Leftarrow \kw{init 0}; E_{23} \syn r : \type{Bit} . (\exists q : \type{Qbit}.R)
+					\cdot; P_0 \vdash q \Leftarrow \kw{init false}; E_{23} \syn r : \type{Bool} . (\exists q : \type{Qbit}.R)
 				}
-				\\ r : \type{Bit}; i,m ; R \prove \top \multimap \top \wedge Id(r, \kw{0})
+				\\ r : \type{Bool}; i,m ; R \prove \top \multimap \top \wedge Id(r, \kw{0})
 			}
 			{
-				\cdot; \lambda i . \lambda m . i = m \wedge (\top * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bit} . (\top \multimap \top \wedge Id(r, \kw{0}))
+				\cdot; \lambda i . \lambda m . i = m \wedge (\top * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bool} . (\top \multimap \top \wedge Id(r, \kw{0}))
 			}
 		}
-		{\cdot \vdash \kw{do} E \chk \{\top\} r : \type{Bit} \{\top \wedge \kw{Id}(r, \kw{0})\}}
+		{\cdot \vdash \kw{do} E \chk \{\top\} r : \type{Bool} \{\top \wedge \kw{Id}(r, \kw{0})\}}
 	\end{mathpar}
@@
 qplus : {\top} r : Qbit {r =q |+\>}
-      = do q <= init 0;
+      = do q <= init false;
            q <= apply H to q;
@@
 qminus : {\top} r : Qbit {r =q |-\>}
-       = do q <= init 1;
+       = do q <= init true;
             q <= apply H to q;
@@
         {(a, b) =q (alpha |00\> + beta |11\>)}
-      = \a.do b <= init 0;
+      = \a.do b <= init false;
               (a, b) <= apply CNot to [a, b];
@@
 \begin{lstlisting}[language=QHaskell]
-testBell : {\top} (a, b) : (Bit o* Bit) {\top /\ a = b}
+testBell : {\top} (a, b) : (Bool o* Bool) {\top /\ a = b}
          = do (qa, qb) <- bell00;
-              a <= measure qa;
-              b <= measure qb;
+              a <= meas qa;
+              b <= meas qb;
               return (a, b)
@@
         {q =q (alpha |0\> + beta |1\>) /\ (a, e) =q |\Phi+\>}    -- a is entangled with ghost var e
-           (m1, m2) : (Bit o* Bit)
+           (m1, m2) : (Bool o* Bool)
         {  (m1, m2) = 00 /\ e =q (alpha |0\> + beta |1\>)
@@
                  q <= apply H to q;
-                 rq <= measure q;
-                 ra <= measure a;
+                 rq <= meas q;
+                 ra <= meas a;
                  return (rq, ra)
@@
 \begin{lstlisting}[language=QHaskell]
-bob : \Pi m1 m2 : Bit. \Pi b : Qbit
+bob : \Pi m1 m2 : Bool. \Pi b : Qbit
       {(b, e) =q |\Phi+\>}

### 2020-10-30 ms-paper/ms-paper.tex
Update examples in 3.3, add entangle-disentangle example

@@
 \subsubsection{Hello Quantum World}
- \todo{TODO}
 \leavevmode
@@
 \begin{lstlisting}[language=QHaskell]
-hqw : {\top} r : Bit {\top /\ r=0}
-    = do q <= init 0;
-         return measure q
+hqw : {\top} r : Bool {\top /\ r = false}
+    = do q <= init false;
+         r = meas q;
+         return r
 \end{lstlisting}
@@
 \begin{lstlisting}[language=QHaskell]
-hqw : {\top} r : Bit {\top /\ r = 0}
--- P0: \top
-    = do q <= init 0;
--- P1: P0 \o (q |-> |0\>)
-         return measure q
--- P2: P1 \o ((q |-> -) -o \top)
+hqw : {\top} r : Bool {\top /\ r = false}
+         -- \top
+    = do q <= init false;
+         -- q =q |0\> /\ class(q)
+         r = meas q;
+         -- (q, e) =q |00\> /\ r = false -- could use a sugared version for measured
+         return r
+         -- r = false -- if no qubit left, should be able to switch to \top here
 \end{lstlisting}
@@
 \begin{lstlisting}[language=QHaskell]
-rnd : {\top} r : Bit {\top}
-    = do q <= init 0;
-         q <= apply H to q;
-         return measure q
+rnd : {\top} r : Bool {\top}
+    = do q <= init false;
+         apply H to q;
+         r = meas q;
+         return r
 \end{lstlisting}
@@
 
-In the case of quantum coin toss, the result can be in either 0 or 1, hence, we need not specify any special postcondition as type checking the result will be sufficient for correctness.
+In the case of quantum coin toss, the result can be either true or false, hence, we need not specify any special postcondition as type checking the result will be sufficient for correctness.
 
-Verification steps follow: \todo{TODO}
+Verification steps follow:
 
@@
 \begin{lstlisting}[language=QHaskell]
-rnd : {\top} r : Bit {\top}
--- P0: \top
-    = do q <= init 0;
--- P1: P0 \o (q |-> |0\>)
-         q <= apply H to q;
--- P2: P1 \o ((q |-> |0\>) -o (q |-> |+\>))
-         return measure q
--- P3: P2 \o ((q |-> -) -o \top)
+rnd : {\top} r : Bool {\top}
+         -- \top
+    = do q <= init false;
+         -- q =q |0\> /\ class(q)
+         apply H to q;
+         -- q =q |+\> /\ uniform(q)
+         r = meas q;
+         -- (q, e) =q |++\> -- perhaps we should be able to say uniform(r)?
+         return r
+         -- r = false \/ r = true -- or left unspecified
 \end{lstlisting}
@@
 \leavevmode
-\todo{TODO: syntax for applying multi-qubit gate}
 
@@
 \begin{lstlisting}[language=QHaskell]
-testBell00 : {\top} (a, b) : (Bit o* Bit) {\top /\ a = b}
-         = do qa <= init 0;
-              qa <= apply H to qa;
-              qb <= init 0;
-              (qa, qb) <= apply CNOT to [qa, qb];
-              return (measure qa, measure qb)
+testBell00 : {\top} (a, b) : (Bool o* Bool) {\top /\ a = b}
+         = do qa <= init false;
+              apply H to qa;
+              qb <= init false;
+              <= apply CNOT to (qa, qb);
+              a = meas qa;
+              b = meas qb;
+              return (a, b)
 \end{lstlisting}
@@
 
-Here, we are asserting that the returned bits $a$ and $b$ hold the same value, which is what we expect from the first Bell state prepared in this program.
+Here, we are asserting that the returned booleans $a$ and $b$ hold the same value, which is what we expect from the first Bell state prepared in this program.
 
@@
 We have elided specific details about unitary application as they are still being worked out.
-
+\newpage
 \begin{lstlisting}[language=QHaskell]
-testBell00 : {\top} (a, b) : (Bit o* Bit) {\top /\ a = b}
-              -- P0: \top
-         = do qa <= init 0;
-              -- P1: P0 \o (qa |-> |0\>)
-              qa <= apply H to qa;
-              -- P2: P1 \o ((qa |-> |0\>) -o (qa |-> |+\>))
-              qb <= init 0;
-              -- P3: P2 \o (qb |-> |0\>)
-              (qa, qb) <= apply CNOT to [qa, qb];
-              -- P4: P3 \o ((qa |-> |+\>, qb |-> |0\>) -o (qa, qb) |-> |\Phi+\>)
-              return (measure qa, measure qb)
-              -- P5: P4 \o ((qa |-> -) -o \top) \o ((qb |-> -) -o \top)
+testBell00 : {\top} (a, b) : (Bool o* Bool) {\top /\ a = b}
+              -- \top
+         = do qa <= init false;
+              -- qa =q |0\>, class(qa)
+              apply H to qa;
+              -- qa =q |+\>, uniform(qa)
+              qb <= init false;
+              -- qa =q |+\>, uniform(qa), qb =q |0\>, class(qb)
+              apply CNOT to (qa, qb);
+              -- (qa, qb) =q |++\>  -- can use a weaker but more general `EPR' here
+              -- /\ uniform(qa, qb) -- this won't apply then
+              a = meas qa;
+              --   (qa, e1) =q |00\>, (qb, e2) =q |00\>, a = false
+              -- \/ (qa, e1) =q |11\>, (qb, e2) =q |11\>, a = true
+              b = meas qb;
+              --   (qa, e1) =q |00\>, (qb, e2) =q |00\>, a = false, b = false
+              -- \/ (qa, e1) =q |11\>, (qb, e2) =q |11\>, a = true,  b = true
+              return (a, b)
 \end{lstlisting}
@@
 %\subsection{Soundness / Type Checking Algorithm}
+
+\subsection{Entangle disentangle}
+\todo{OPTIONAL:}
+
+\begin{lstlisting}[language=QHaskell]
+entangle-disentangle : {T} (a, b) : (Qbit o* Qbit) {a = b}
+= do
+    -- P0: T
+    a <= init false;
+    -- P1: a =q |0\>, class(a)
+    b <= init false;
+    -- P2: P1, b =q |0\>, class(b)
+    apply H to a;
+    -- P3: H_a.P2     =   a =q |+\>, b =q |0\>, uniform(a), class(b)
+    apply CX to (a, b);
+    -- P4: CX_ab.P3   =   (a, b) =q |++\>, uniform(a, b)
+    apply CX to (a, b);
+    -- P5: CX_ab.P4   =   (a =q |+\>, b =q |0\>) \/ (a =q |0\>, b =q |+\>)
+    apply H to a;
+    -- P6: H_a.P5     =   (a =q |0\>, b =q |0\>) \/ (a =q |+\>, b =q |+\>)
+    return (a, b)
+\end{lstlisting}
 

### 2020-10-30 ms-paper/ms-paper.tex
Use classic(X) everywhere not classical(X)
A typo and unneeded statement

@@
 		\begin{minipage}[t]{0.5\columnwidth}%
-			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid U \cdot P \mid \textbf{X} =_q \psi \mid \kw{uniform}(\textbf{X}) \mid \kw{separable}(\textbf{X}) \mid \kw{classical}(\textbf{X}) $
+			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid U \cdot P \mid \textbf{X} =_q \psi \mid \kw{uniform}(\textbf{X}) \mid \kw{separable}(\textbf{X}) \mid \kw{class}(\textbf{X}) $
 		\end{minipage}\\ \\
@@
 The predicate language is inspired by \citet{unruh2019} that follows subspace semantics.
-We need to be explicit about what we mean when we say that a certain state satisifes a certain predicate in the quantum setting.
+We need to be explicit about what we mean when we say that a certain state satisfies a certain predicate in the quantum setting.
 In this model, the propositions are subspaces of the Hilbert space corresponding to a quantum state.
@@
 For illustration, we show the propositions annotated as comments as we step through the \texttt{testBell00} program below.
-We have elided specific details about unitary application as they are still being worked out.
 \newpage
@@
 \subsection{Testing modular Bell pair}
-\todo{TODO: how is classical equality ensured?}
 \begin{lstlisting}[language=QHaskell]

### 2020-11-05 ms-paper/ms-paper.tex
Add references for subspace semantics
Add 1936 paper in bib

@@
 We provide more details when we introduce QHTT in section~\cref{sec:qhtt}. An accessible introduction to Hoare Type Theory is available in the lecture notes by \citet{perconti2012}.
+
+\subsection{A Crash Course on Subspace Semantics}
+
+Main references: \cite{logicqm36,ying_duan_feng_ji_2009,aqhl2019,unruh2019}.
+
+\citet{ying_duan_feng_ji_2009} present a systematic exposition of predicate transformer semantics for quantum programs focusing on the Birkhoff-von Neumann quantum logic by restricting to the quantum predicates expressible using projection operators.
+In this section, we will review how projection operators defined over Hilbert space can be viewed as predicates that will help us define the proposition language used in our Quantum Hoare Type Theory.
+We will also define various logial operations on these predicates that will help us reason about quantum programs.
+
+This restriction to projection operators, which are Hermitian operators with 0 or 1 as their eigenvalues, will let us reason about quantum programs similar to classical boolean predicates, without needing a quantum analogue of probabilistic logic.
+One can think of projection operators as performing a yes or no measurement for the chosen subspace.
+Further, we note that projection operators in Hilbert space correspond one to one with closed subspaces of this space.
 

### 2020-11-06 ms-paper/ms-paper.tex
Add definitions of Hermitian and projection operators

--- 2020-10-30 ms-paper/ms-paper.tex
@@
 We provide more details when we introduce QHTT in section~\cref{sec:qhtt}. An accessible introduction to Hoare Type Theory is available in the lecture notes by \citet{perconti2012}.
+
+\subsection{A Crash Course on Subspace Semantics}
+
+Main references: \cite{logicqm36,ying_duan_feng_ji_2009,aqhl2019,unruh2019}.
+
+\citet{ying_duan_feng_ji_2009} present a systematic exposition of predicate transformer semantics for quantum programs utilizing the algebraic structure of the Birkhoff-von Neumann quantum logic by restricting to quantum predicates expressible using projection operators.
+In this section, we will review how projection operators defined over Hilbert space can be viewed as predicates that will help us define the proposition language used in our Quantum Hoare Type Theory.
+We will also define various logial operations on these predicates that will help us reason about quantum programs.
+
+This restriction to projection operators, which are Hermitian operators with 0 or 1 as their eigenvalues, will let us reason about quantum programs in a fashion similar to classical Boolean predicates.
+Further, we note that projection operators in Hilbert space correspond one to one with closed subspaces of this space.
+Let us start with some basics.
+
+\begin{definition}[Hermitian operator]
+	A Hermitian operator, $A$, in a finite-dimensional complex Hilbert space, $\mathcal{H}$, is a self-adjoint operator in $\mathcal{H}$. It is equal to its complex conjugate: $A = A^\dagger$.
+\end{definition}
+Hermitian operators describe physical observables and their eigenvalues are real numbers.
+\citet{dhondt_quantum_2006} define a quantum predicate in $\mathcal{H}$ as a Hermitian operator with eigenvalues within the unit interval ($[0, 1]$).
+
+\citet{ying_duan_feng_ji_2009} point out a problem with that approach that it is not always possible to define certain logical operations such as conjunction and disjunction on some of these quantum predicates as they do not for a lattice.
+These operations are necessary to be able to compositionally reason about quantum programs.
+They, in turn, restrict themselves to quantum predicates that are expressed using a special class of Hermitian operators, called projection operators or projectors.
+It is well-known that projection operators in Hilbert space correspond one to one with closed subspaces in that space.
+Further, \citet{logicqm36} showed that the set of closed subspaces of Hilbert space form an orthomodular lattice that is used to define a logic for quantum mechanics.
+Similar to \cite{ying_duan_feng_ji_2009}, we will restrict ourselves to projection operators (or equivalently closed subspaces) to define quantum predicates.
+
+\begin{definition}[Projection operator]
+	Let $X$ be a closed subspace of $\mathcal{H}$ and a vector $\ket{\psi} \in \mathcal{H}$.
+Then, there uniquely exist $\ket{\psi_0} \in X$ and $\ket{\psi_1} \in X^\bot$ (complement of X) such that $\ket{\psi} = \ket{\psi_0} + \ket{\psi_1}$.
+The vector $\ket{\psi_0}$ is called the projection of $\ket{\psi}$ onto subspace $X$ and is equal to $\ket{\psi_0} = P_X \ket{\psi}$.
+The operator $P_X$ is called the projection operator (or projector) onto $X$.
+\end{definition}
 

### 2020-11-06 ms-paper/ms-paper.tex
Add definition of span

@@
 \end{definition}
+
+\todo{Consider defining orthomodular lattices}
+
+Let us define the notation and basic definitions we will need for talking about subspaces of a Hilbert space.
+
+\begin{definition}[Span of a subspace]
+	For any subspace $X$ in a Hilbert space $\mathcal{H}$, written ($X \subseteq \mathcal{H}$), $\mathit{span}(X)$ is the smallest closed subspace of $\mathcal{H}$ containing $X$.
+In other words, $\mathit{span}(X)$ is the set of linear combinations of vectors in $X$.
+\end{definition}
+
+Further, $S(\mathcal{H})$ is the set of closed subspaces of $\mathcal{H}$.
+If we write $\mathcal{P(H)}$ for the set of all quantum predicates on $\mathcal{H}$, then $S(\mathcal{H}) \subseteq \mathcal{P(H)}$.
+
+Now we can define the logical operations on quantum predicates:
 

### 2020-11-06 ms-paper/ms-paper.tex
Finish initial version of crash course in subspace semantics

@@
 
-\subsection{A Crash Course on Subspace Semantics}
+\subsection{A Crash Course in Subspace Semantics for Quantum Predicates}
 
@@
 
-\citet{ying_duan_feng_ji_2009} point out a problem with that approach that it is not always possible to define certain logical operations such as conjunction and disjunction on some of these quantum predicates as they do not for a lattice.
+\citet{ying_duan_feng_ji_2009} point out a problem with the above approach that it is not always possible to define certain logical operations (such as conjunction and disjunction) on some of these quantum predicates as they do not for a lattice.
 These operations are necessary to be able to compositionally reason about quantum programs.
@@
 It is well-known that projection operators in Hilbert space correspond one to one with closed subspaces in that space.
-Further, \citet{logicqm36} showed that the set of closed subspaces of Hilbert space form an orthomodular lattice that is used to define a logic for quantum mechanics.
-Similar to \cite{ying_duan_feng_ji_2009}, we will restrict ourselves to projection operators (or equivalently closed subspaces) to define quantum predicates.
+Further, \citet{logicqm36} showed that the set of closed subspaces of Hilbert space form an orthomodular lattice that they use to define a logic for quantum mechanics.
+Similar to \citet{ying_duan_feng_ji_2009}, we will restrict ourselves to projection operators (or equivalently closed subspaces) to define quantum predicates.
 
@@
 	Let $X$ be a closed subspace of $\mathcal{H}$ and a vector $\ket{\psi} \in \mathcal{H}$.
-Then, there uniquely exist $\ket{\psi_0} \in X$ and $\ket{\psi_1} \in X^\bot$ (complement of X) such that $\ket{\psi} = \ket{\psi_0} + \ket{\psi_1}$.
+Then, there uniquely exist $\ket{\psi_0} \in X$ and $\ket{\psi_1} \in X^\bot$ (the orthocomplement of X) such that $\ket{\psi} = \ket{\psi_0} + \ket{\psi_1}$.
 The vector $\ket{\psi_0}$ is called the projection of $\ket{\psi}$ onto subspace $X$ and is equal to $\ket{\psi_0} = P_X \ket{\psi}$.
@@
 
-\todo{Consider defining orthomodular lattices}
+\todo{Consider defining orthomodular lattices formally}
 
@@
 
-Now we can define the logical operations on quantum predicates:
+Now we can define the logical operations on quantum predicates following the algebraic structure of the set of closed subspaces of $\mathcal{H}$.
+Recall that the closed subspaces form an orthomodular lattice (with set inclusion as the partial order) which does not satisfy the full distributive law of Boolean algebra but will be sufficient for our needs.
+
+\begin{itemize}
+	\item $\top$ represents the complete Hilbert space, $\mathcal{H}$, under consideration.
+	\item $\bot$ represents the zero-dimension subspace of Hilbert space, $\{0\}$.
+	\item $X \wedge Y$ is taken to be the set intersection of the two subspaces: $X \cap Y$.
+	\item $X \vee Y$ is modeled as the sum (linear combination) of the subspaces $X$ and $Y$.
+In other words, it is the closed subspace of the union of the two subspaces: $\overline{\mathit{span}(X \cup Y)}$, where $\overline{X}$ denotes that $X$ is a closed subspace.
+	\item $X^\bot$ denotes the orthocomplement of $X$ which includes all vectors in $\mathcal{H}$ that are orthogonal to $X$.
+\end{itemize}
+
+\todo{TODO: also need to talk about mixed state / density operators}
 

### 2020-11-06 ms-paper/ms-paper.tex
Remove QIO from bg, move HTT before QC, fix von Neumann name

@@
 
-This paper is organized as follows.
+%This paper is organized as follows.
 We discuss background in the next section, \cref{sec:bg}; specifically, quantum computing (\cref{sec:qc}), Hoare Type Theory in the classical setting (\cref{sec:htt}) and then the Quantum IO Monad interface (\cref{sec:qio}).
@@
 \label{sec:bg}
-In this section, we provide background on the core ideas from existing literature that form the foundation for our work --- Hoare Type Theory (HTT) and the Quantum IO Monad interface.
-We start with some basics of quantum computing.
-% and refer the reader to Nielsen's excellent series of essays~\cite{matuschak2019} for a first introduction and to the standard textbook~\cite{nielsen2010} for advanced material.
+In this section, we provide background on the core ideas from existing literature that form the foundation for our work --- Hoare Type Theory (HTT), quantum computing, and quantum logic defined using closed subspaces.
+
+\subsection{Hoare Type Theory}
+\label{sec:htt}
+A Hoare type, $\Delta.\Psi.\{P\}\; \texttt{x}:A\; \{Q\}$, encodes preconditions and postconditions in the same spirit as Hoare triples to allow both specification and verification of effectful classical programs.
+It can be read as `for a stateful computation executed in a heap that satisfies precondition $P$, return a value of type $A$ in a heap that satisfies postcondition $Q$.' The contexts, $\Delta$ and $\Psi$, contain the variables and heap variables that may appear in both $P$ and $Q$.
+In this presentation, we will omit the contexts in the Hoare type when they are unneeded.
+We show an example to demonstrate the expressiveness of a Hoare type --- the \texttt{alloc} primitive from HTT~\cite{nanevski_hoare_2008}:
+\begin{mathpar}
+	\texttt{alloc} : \forall \alpha. \Pi x:\alpha. \{\texttt{emp}\}\; y:\texttt{nat}\; \{y \mapsto_{\alpha} x\}
+\end{mathpar}
+
+This type specifies that \texttt{alloc} is a polymorphic function that takes as input a variable, $x$, of any simple type, $\alpha$, that is executed in an empty heap (meaning it does not affect existing heap), returns a new location bound by a fresh variable, $y$, of type \texttt{nat} and initializes it with the supplied value $x$ of type $\alpha$.
+
+Type checking in HTT involves generation of strongest postconditions at each step of the program.
+Verification is a two-step process: the first phase does basic type checking and verification condition generation which is decidable, the second phase needs to show the validity of the generated verification conditions, which can be undecidable.
+This second phase can be deferred to an automated theorem prover.
+
+We provide more details when we introduce QHTT in section~\cref{sec:qhtt}. An accessible introduction to Hoare Type Theory is available in the lecture notes by \citet{perconti2012}.
 
@@
 
-\subsection{Hoare Type Theory}
-\label{sec:htt}
-A Hoare type, $\Delta.\Psi.\{P\}\; \texttt{x}:A\; \{Q\}$, encodes preconditions and postconditions in the same spirit as Hoare triples to allow both specification and verification of effectful classical programs.
-It can be read as `for a stateful computation executed in a heap that satisfies precondition $P$, return a value of type $A$ in a heap that satisfies postcondition $Q$.' The contexts, $\Delta$ and $\Psi$, contain the variables and heap variables that may appear in both $P$ and $Q$.
-In this presentation, we will omit the contexts in the Hoare type when they are unneeded.
-We show an example to demonstrate the expressiveness of a Hoare type --- the \texttt{alloc} primitive from HTT~\cite{nanevski_hoare_2008}:
-\begin{mathpar}
-	\texttt{alloc} : \forall \alpha. \Pi x:\alpha. \{\texttt{emp}\}\; y:\texttt{nat}\; \{y \mapsto_{\alpha} x\}
-\end{mathpar}
-
-This type specifies that \texttt{alloc} is a polymorphic function that takes as input a variable, $x$, of any simple type, $\alpha$, that is executed in an empty heap (meaning it does not affect existing heap), returns a new location bound by a fresh variable, $y$, of type \texttt{nat} and initializes it with the supplied value $x$ of type $\alpha$.
-
-Type checking in HTT involves generation of strongest postconditions at each step of the program.
-Verification is a two-step process: the first phase does basic type checking and verification condition generation which is decidable, the second phase needs to show the validity of the generated verification conditions, which can be undecidable.
-This second phase can be deferred to an automated theorem prover.
-
-We provide more details when we introduce QHTT in section~\cref{sec:qhtt}. An accessible introduction to Hoare Type Theory is available in the lecture notes by \citet{perconti2012}.
-
 \subsection{A Crash Course in Subspace Semantics for Quantum Predicates}
@@
 \end{definition}
-Hermitian operators describe physical observables and their eigenvalues are real numbers.
+%Hermitian operators describe physical observables and their eigenvalues are real numbers.
 \citet{dhondt_quantum_2006} define a quantum predicate in $\mathcal{H}$ as a Hermitian operator with eigenvalues within the unit interval ($[0, 1]$).
 
-\citet{ying_duan_feng_ji_2009} point out a problem with the above approach that it is not always possible to define certain logical operations (such as conjunction and disjunction) on some of these quantum predicates as they do not for a lattice.
-These operations are necessary to be able to compositionally reason about quantum programs.
-They, in turn, restrict themselves to quantum predicates that are expressed using a special class of Hermitian operators, called projection operators or projectors.
+\citet{ying_duan_feng_ji_2009} restrict themselves to quantum predicates that are expressed using a special class of Hermitian operators, called projection operators or projectors.
 It is well-known that projection operators in Hilbert space correspond one to one with closed subspaces in that space.
@@
 \todo{TODO: also need to talk about mixed state / density operators}
-
-\subsection{Quantum IO Monad}
-\label{sec:qio}
-
-The Quantum IO (QIO) monad~\cite{qio} is a purely functional interface for quantum programming that provides a separation between the unitary (reversible) and non-unitary (irreversible) fragments of quantum computation.
-It provides isolation of quantum effects inside a monad similar to what we need to provide in our Quantum Hoare monad.
-
-QIO interface was developed as a library for the Haskell programming language but its design was influenced~\cite{asg2010} by the category \textbf{FQC} of finite quantum computations (where computations are interpreted as superoperators) as explored by \citet{fqc06} in previous work.
-This is the most we will say about the relation of QIO monad to category theory in this paper.
-
-Here we discuss the relevant concepts from the QIO interface.
-QIO provides a type for referring to qubits, \texttt{Qbit}; a type for unitary operations, \texttt{U}; and, the \texttt{QIO} type operator which is a monad indexed by the result of the quantum computation.
-The primitive quantum operations defined using the QIO monad below (Haskell syntax) are:
-
-\begin{minipage}{0.95\linewidth}
-\begin{lstlisting}[language=QHaskell]
-mkQbit :: Bool -> QIO Qbit       -- initialization
-applyU :: U -> QIO ()            -- apply a unitary to quantum state
-measQbit :: Qbit -> QIO Bool     -- measurement
-\end{lstlisting}
-\end{minipage}
-
-Further, \texttt{U} is defined as a monoid with neutral element, \texttt{mempty}, and operation, \texttt{mappend}, that encodes reversible operations on quantum state.
-The core unitary primitives that we will need are:
-
-\begin{minipage}{0.95\linewidth}
-\begin{lstlisting}[language=QHaskell]
-rot :: Qbit -> ((Bool, Bool) -> C) -> U
-cond :: Qbit -> (Bool -> U) -> U
-\end{lstlisting}
-\end{minipage}
-
-\texttt{rot} takes a qubit and a two-by-two complex-valued unitary matrix (represented as a function from the matrix indices to $\mathbb{C}$) and lets one define arbitrary rotation on a single qubit.
-\texttt{rot} can then be used to define standard gates such as the Hadamard, H, and the Pauli gates, X, Y and Z.
-\texttt{cond} is the unitary conditional that is used to perform branching unitary operations.
-For the present paper, we use a simpler conditional, \texttt{ifQ}, that is defined as follows:
-
-\begin{minipage}{0.95\linewidth}
-\begin{lstlisting}[language=QHaskell]
-ifQ :: Qbit -> U -> U
-ifQ q u = cond q (\x -> if x then u else mempty)
-\end{lstlisting}
-\end{minipage}
-
-That is, \texttt{ifQ} acts as the standard control operator in the quantum circuit model that runs its second argument based on the truth value of its first argument.
-Then, it is easy to define controlled gates such as CNOT with the expression $\kw{ifQ q1 (X q2)}$.
-
-Quantum state in the QIO interface is modeled as a normalized vector that stores pairs representing complex amplitudes associated with each basis state (represented as a map from Qbit to Bool types).
-The operations defined over the vector class ensure that the quantum state is kept normalized throughout.
-
-QIO further ensures that the monoidal structure of unitary operations lets one run a unitary operation over the complete quantum state.
-We elide details here and refer the reader to \cite{qio} for more.
 
@@
 		\textit{Computations} & $E, F$ & ::= & $ \kw{return} N \mid x \leftarrow K; E \mid x \Leftarrow c; E \mid x = N; E $\\ \\
-%		\textit{Variable context} & $\Delta$ & ::= & $ \cdot \mid x : A, \Delta $ \\
-%		\textit{Q. Heap context} & $\Psi$ & ::= & $ \cdot \mid h, \Psi$ \\
-%		\textit{Proposition context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
+		\textit{Variable context} & $\Delta$ & ::= & $ \cdot \mid x : A, \Delta $ \\
+		\textit{Q. Heap context} & $\Psi$ & ::= & $ \cdot \mid h, \Psi$ \\
+		\textit{Proposition context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
 		\textit{Unitaries} & $U$ & ::= & $ I \mid X \mid Y \mid Z \mid H \mid CX \mid CZ \mid S \mid T$

### 2020-11-09 ms-paper/ms-paper.tex
Remove unneeded page breaks

@@
 For illustration, we show the propositions annotated as comments as we step through the \texttt{testBell00} program below.
-\newpage
+
 \begin{lstlisting}[language=QHaskell]
@@
 \end{lstlisting}
-
-\section{Implementation}
-We are implementing QHTT in the F* programming language.
-
-One issue with the original formulation of Hoare Type Theory is that using Hoare triples does not directly lead to a verification condition generation algorithm.
-This results in requiring a complicated bidirectional typing system for implementation.
-If we use proposition transformers as demonstrated by the use of Dijkstra monads in F*, we can avoid these issues \cite{swamy2013verifying}.
-
-The current steps for the implementation are:
-
-\begin{itemize}
-	\item Design a memory model for representing quantum state, along with elementary operations such as read, write, lookup, etc
-	\item Expose a new Quantum effect in F* by using the Dijkstra Monads for Free technique \cite{dm4free2017}.
-	\item Port our language to work well with F* syntax.
-\end{itemize}
-
-This will necessarily take us away from the bidirectional typing approach used in the previous section.
-But the end result for the user would be the same.
-Indeed, F* system is an evolution of the HTT approach that comes with inherent support for effects and automation using an SMT solver.
-
-%\section{Evaluation}
-
-%\subsection{More Examples}
-%\textit{add at least one example with scaling knob}
-
-%\subsection{Circuit-complexity Considerations}
 
@@
 
-\newpage
-
 \bibliography{references}
-
-\newpage
 

### 2020-11-09 ms-paper/ms-paper.tex
Add definition of a subspace and mention Hilbert space

@@
 \label{sec:qc}
-The basic unit of information in quantum computing is a qubit, that can be represented using a unit vector in a two-dimensional complex vector space.
-Using Dirac's bra-ket notation, we can write an arbitrary qubit, $\ket{\psi}$, as a linear combination of the computational basis vectors: $\ket{\psi} = \alpha\ket{0} + \beta\ket{1}$, where $\ket{0}=\begin{bsmallmatrix}
+The basic unit of information in quantum computing is a qubit, that can be represented using a unit vector in a two-dimensional complex vector space, $\mathbb{C}^2$.
+Using Dirac's bra-ket notation, we can write an arbitrary qubit, $\ket{\psi}$, as a linear combination (or \textit{superposition}) of the computational basis vectors: $\ket{\psi} = \alpha\ket{0} + \beta\ket{1}$, where $\ket{0}=\begin{bsmallmatrix}
 1 \\
@@
 To work with multiple qubits, we need to take their tensor product, denoted by $\otimes$.
-Computation is performed by the application of unitary quantum gates, such as the Hadamard (H) gate or the controlled NOT (CX) gate, that can be represented as square matrices.
+Computation is performed by the application of unitary quantum gates, such as the Hadamard (H) gate or the controlled-Not (CX) gate, that can be represented as square matrices.
 Mathematically, a gate application is equivalent to multiplying the matrix representing the gate with the state vector of the qubit(s).
 
-This basic unit of information is not the only reason why quantum computing is fundamentally different from classical computing---certain quantum mechanical phenomenon are highly counterintuitive!
+This basic unit of information is not the only reason why quantum computing is fundamentally different from classical computing---certain quantum mechanical phenomenon are highly counter-intuitive!
 One such phenomenon is \textit{quantum entanglement} which, at its most basic level, involves a correlation between two qubits such that when one is measured, the outcome obtained necessarily influences the measurement outcome of the other qubit (even if the two qubits are far apart).
-%We will see how this property can be utilized for computation in the quantum teleportation algorithm that we describe and program in \S\ref{telep}.
 We refer the reader to the standard textbook by \citet{nielsen_quantum_2010} or the excellent series of essays by \citet{matuschak2019} for more background.
+
+We assume basic familiarity with linear algebra. Here we define some mathematical preliminaries specific to quantum computing that we will need.
+
+\begin{definition}[Vector Subspace]
+	A subspace of a vector space is a subset of its vectors which is closed under its operations of vector addition and scalar multiplication.
+\end{definition}
+
+All vector spaces are equipped with at least two (\textit{trivial}) subspaces: the singleton set, $\{ 0 \}$, consisting of the zero vector and the vector space itself.
+
+We will work in a specific vector space called \textit{Hilbert space}, $\mathcal{H}$, which for our purposes can be thought of as an n-dimensional complex vector space equipped with an inner product ($\mathbb{C}^n$).
+It is tricky to precisely define a Hilbert space as it needs to satisfy certain axioms from mathematical analysis but we will not need them in this work as we only consider finite-dimensional Hilbert spaces.
+We have already seen an important example of a Hilbert space above when we described a qubit: $\mathbb{C}^2$.
 

### 2020-11-11 ms-paper/ms-paper.tex
Move examples to individual files and switch to unicode literal chars

@@
 	language     = Haskell,
-	morekeywords = {Qbit, U, C, meas, apply, to},
+	morekeywords = {bool, qbit, U, C, meas, apply, to},
 }
@@
 	literate={+}{{$+$}}1 {/}{{$/$}}1 {*}{{$*$}}1 {=}{{$=$}}1
-	{>}{{$>$}}1 {<}{{$<$}}1 {\\}{{$\lambda$}}1
+	{√}{{$\surd$}}1
+	{>}{{$>$}}1 {<}{{$<$}}1 {λ}{{$\lambda$}}1
 	{\\\\}{{\char`\\\char`\\}}1
-	{->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
-	{<=}{{$\Leftarrow$}}2 {=>}{{$\Rightarrow$}}2
+	{⇐}{{$\Leftarrow$}}1 {⇒}{{$\Rightarrow$}}2
 	{\ .}{{$\circ$}}2 {\ .\ }{{$\circ$}}2
-	{>>}{{>>}}2 {>>=}{{>>=}}2
-	{|}{{$\vert$}}1 {\\>}{{$\rangle$}}1
-	{/\\}{{$\wedge$}}1 {\\/}{{$\vee$}}1
-	{\\in}{{$\in$}}1 {\\o}{{$\circ$}}1 {\\Pi}{{$\Pi$}}1
-	{\\Phi+}{{$\Phi^+$}}2
-	{|->}{{$\mapsto$}}2 {-o}{{$\multimap$}}2
-	{o*}{{$\otimes$}}1
-	{\\top}{{$\top$}}1
-	{\\bot}{{$\bot$}}1
-	{=q}{{$=_q$}}2
-	{alpha}{{$\alpha$}}1 {beta}{{$\beta$}}1
+	{|}{{$\vert$}}1 {⟩}{{$\rangle$}}1
+	{∧}{{$\wedge$}}1 {∨}{{$\vee$}}1
+	{∈}{{$\in$}}1 {Π}{{$\Pi$}}1
+	{⊗}{{$\otimes$}}1
+	{⊤}{{$\top$}}1
+	{=q}{{$=_q$}}2 {=c}{{$=_c$}}2
+	{α}{{$\alpha$}}1 {β}{{$\beta$}}1
 }
@@
 		\textit{Types} & $A, B, C$ & ::= &
-		\begin{minipage}[t]{0.45\columnwidth}%
-			$ \type{1} \mid \type{Bool} \mid \type{Qbit} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
+		\begin{minipage}[t]{0.6\columnwidth}%
+			$ \type{unit} \mid \type{bool} \mid \type{prop} \mid \type{qbit} \mid A \otimes B \mid \Pi x{ : }A.B \mid \{P\} x{ : }A \{Q\}$
 		\end{minipage}\\ \\
 		\textit{Propositions} & $P, Q, R$ & ::= &
-		\begin{minipage}[t]{0.5\columnwidth}%
-			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid U \cdot P \mid \textbf{X} =_q \psi \mid \kw{uniform}(\textbf{X}) \mid \kw{separable}(\textbf{X}) \mid \kw{class}(\textbf{X}) $
+		\begin{minipage}[t]{0.6\columnwidth}%
+			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid \neg P \mid U \cdot P \mid \textbf{X} =_q \psi$
 		\end{minipage}\\ \\
 		% \textit{Q. Heaps} & $H, G$ & ::= & $ h \mid \kw{empty} \mid \kw{upd}(H, M, N) $\\ \\
-		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K N \mid N : A $\\
-		\textit{Intro terms} & $N, O$ & ::= & $ K \mid () \mid \lambda x.N \mid \kw{do} E \mid \kw{true} \mid \kw{false} $\\ \\
+		\textit{Elim terms} & $K, L$ & ::= & $ x \mid \kw{fst} K \mid \kw{snd} K \mid K\ N \mid M : A $\\
+		\textit{Intro terms} & $M, N, O$ & ::= & $ K \mid () \mid \kw{true} \mid \kw{false} \mid (M, N) \mid \lambda x.M \mid \kw{do} E$\\ \\
+		\textit{Unitaries} & $U$ & ::= & $ I \mid X \mid Y \mid Z \mid H \mid CX \mid CZ \mid S$\\
 		\textit{Q. Commands} & $c$ & ::= &
-		\begin{minipage}[t]{0.5\columnwidth}%
-			$ \kw{\textbf{init}} N \mid \kw{\textbf{apply}} U \kw{\textbf{to}} N \mid \kw{\textbf{meas}} N$
+		\begin{minipage}[t]{0.6\columnwidth}%
+			$ \kw{init} N \mid \kw{apply} U \kw{to} N \mid \kw{meas} N \mid
+			\kw{if} M \kw{then} E \kw{else} F$
 		\end{minipage}\\
-		\textit{Computations} & $E, F$ & ::= & $ \kw{return} N \mid x \leftarrow K; E \mid x \Leftarrow c; E \mid x = N; E $\\ \\
-		\textit{Variable context} & $\Delta$ & ::= & $ \cdot \mid x : A, \Delta $ \\
-		\textit{Q. Heap context} & $\Psi$ & ::= & $ \cdot \mid h, \Psi$ \\
-		\textit{Proposition context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
-		\textit{Unitaries} & $U$ & ::= & $ I \mid X \mid Y \mid Z \mid H \mid CX \mid CZ \mid S \mid T$
+		\textit{Computations} & $E, F$ & ::= & $ \kw{return} N \mid \kw{run} x = K \kw{in} E \mid x = c; E \mid x = N; E $\\ \\
+		\textit{Context} & $\Delta$ & ::= & $ \cdot \mid x : A, P, \Delta $ \\
+%		\textit{Q. Heap context} & $\Psi$ & ::= & $ \cdot \mid h, \Psi$ \\
+%		\textit{Proposition context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
+
 	\end{tabular}
@@
 
-There are other propositions that involve ghosts:
-\begin{itemize}
-	\item uniform(\textbf{X}) is an abbreviation for $(\textbf{X}, \textbf{e}) = \frac{1}{\sqrt{2}}(\ket{00}+\ket{11})$ where \textbf{e} is  a fresh entangled ghost variable.
-It means that X is uniformly distributed.
-	\item separable(\textbf{X}) is an abbreviation for $\textbf{X} \equiv_{q} \textbf{u}$ where \textbf{u} is a fresh unentangled ghost variable.
-	\item class(\textbf{X}) is an abbreviation for $\textbf{X} \equiv_{cl} \textbf{u}$ where \textbf{u} is a fresh unentangled ghost variable.
-The classical equality symbol means that the variables on both side will measure to be the exact same value.
-That is, there is no superposition involved.
-\end{itemize}
-
 In addition to these quantum-specific propositions, the system also supports first-order logic (not shown) for reasoning about the pure fragment of the language.
@@
 
-We use the terms quantum state and quantum heaps interchangeably throughout.
+\todo{TODO}
+
+%We use the terms quantum state and quantum heaps interchangeably throughout.
 A quantum heap is a partial function from qubits (that can in turn be thought of as locations, represented using natural numbers) to at most one quantum state vector.
@@
 \texttt{apply} $U$ \texttt{to} $X$ applies unitary operation $U$ to variables $X$.
-\texttt{meas} $x$ measures the variable $x$ in the computational basis.
-%\texttt{if y then E else F} executes computation $E$ if measuring variable $y$ returns 1 else executes $F$.
+\texttt{meas} $x$ measures the variable $x$ in the computational basis and returns a classical result.
+\texttt{if y then E else F} executes computation $E$ if measuring variable $y$ returns 1 else executes $F$.
 
@@
 	\inferrule
-	{\Delta \vdash N \chk \type{Bool}
-		\\ \Delta, x : \type{Qbit}; P \circ (x =_q \mathit{bool2ket}(N)) \vdash E \syn y:B . Q}
-	{\Delta; P \vdash x \Leftarrow \kw{init} N; E \syn y:B. (\exists x:\type{Qbit}.Q)}
+	{\Delta \vdash N \chk \type{bool}
+		\\ \Delta, x : \type{qbit}; P \circ (x =_q \mathit{bool2ket}(N)) \vdash E \syn y:B . Q}
+	{\Delta; P \vdash x \Leftarrow \kw{init} N; E \syn y:B. (\exists x:\type{qbit}.Q)}
 \end{mathpar}
@@
 
-Initialization can only be performed for a term $N$ that can be type checked as a \type{Bool}.
-Then, we look at the rest of the computation, $E$, in a context that adds $x$ of type \type{Qbit} under a precondition that extends the previous precondition, $P$, with the strongest postcondition for initialization, that is, the newly bound variable $x$ is equal to the \ket{0} or \ket{1}; this should synthesize the strongest postcondition (with respect to the expanded precondition in the context), $Q$, for $E$.
+Initialization can only be performed for a term $N$ that can be type checked as a \type{bool}.
+Then, we look at the rest of the computation, $E$, in a context that adds $x$ of type \type{qbit} under a precondition that extends the previous precondition, $P$, with the strongest postcondition for initialization, that is, the newly bound variable $x$ is equal to the \ket{0} or \ket{1}; this should synthesize the strongest postcondition (with respect to the expanded precondition in the context), $Q$, for $E$.
 
@@
 	\inferrule
-	{\Delta \vdash N \chk \type{Qbit}
+	{\Delta \vdash N \chk \type{qbit}
 		\\ \Delta; \Psi; P \prove (N =_q -)
-		\\ \Delta, x:\type{Bool}; P \circ ((N =_q -) \multimap \top) \vdash E \syn y:B . Q}
-	{\Delta; P \vdash x \Leftarrow \kw{meas} N; E \syn y:B. (\exists x:\type{Bool}.Q)}
+		\\ \Delta, x:\type{bool}; P \circ ((N =_q -) \multimap \top) \vdash E \syn y:B . Q}
+	{\Delta; P \vdash x \Leftarrow \kw{meas} N; E \syn y:B. (\exists x:\type{bool}.Q)}
 \end{mathpar}
@@
 \begin{minipage}{0.95\linewidth}
-\begin{lstlisting}[language=QHaskell]
-hqw : {\top} r : Bool {\top /\ r = false}
-    = do q <= init false;
-         r = meas q;
-         return r
-\end{lstlisting}
+\lstinputlisting[language=QHaskell,lastline=4]{hqw.qh}
 \end{minipage}
@@
 \begin{minipage}{0.95\linewidth}
-\begin{lstlisting}[language=QHaskell]
-hqw : {\top} r : Bool {\top /\ r = false}
-         -- \top
-    = do q <= init false;
-         -- q =q |0\> /\ class(q)
-         r = meas q;
-         -- (q, e) =q |00\> /\ r = false -- could use a sugared version for measured
-         return r
-         -- r = false -- if no qubit left, should be able to switch to \top here
-\end{lstlisting}
+\lstinputlisting[language=QHaskell,firstline=6]{hqw.qh}
 \end{minipage}
@@
 \begin{minipage}{0.95\linewidth}
-\begin{lstlisting}[language=QHaskell]
-rnd : {\top} r : Bool {\top}
-    = do q <= init false;
-         apply H to q;
-         r = meas q;
-         return r
-\end{lstlisting}
+\lstinputlisting[language=QHaskell,lastline=5]{toss.qh}
 \end{minipage}
@@
 \begin{minipage}{0.95\linewidth}
-\begin{lstlisting}[language=QHaskell]
-rnd : {\top} r : Bool {\top}
-         -- \top
-    = do q <= init false;
-         -- q =q |0\> /\ class(q)
-         apply H to q;
-         -- q =q |+\> /\ uniform(q)
-         r = meas q;
-         -- (q, e) =q |++\> -- perhaps we should be able to say uniform(r)?
-         return r
-         -- r = false \/ r = true -- or left unspecified
-\end{lstlisting}
+\lstinputlisting[language=QHaskell,firstline=7]{toss.qh}
 \end{minipage}
@@
 \begin{minipage}{0.95\linewidth}
-\begin{lstlisting}[language=QHaskell]
-testBell00 : {\top} (a, b) : (Bool o* Bool) {\top /\ a = b}
-         = do qa <= init false;
-              apply H to qa;
-              qb <= init false;
-              <= apply CNOT to (qa, qb);
-              a = meas qa;
-              b = meas qb;
-              return (a, b)
-\end{lstlisting}
+\lstinputlisting[language=QHaskell,lastline=8]{bell.qh}
 \end{minipage}
@@
 
-\begin{lstlisting}[language=QHaskell]
-testBell00 : {\top} (a, b) : (Bool o* Bool) {\top /\ a = b}
-              -- \top
-         = do qa <= init false;
-              -- qa =q |0\>, class(qa)
-              apply H to qa;
-              -- qa =q |+\>, uniform(qa)
-              qb <= init false;
-              -- qa =q |+\>, uniform(qa), qb =q |0\>, class(qb)
-              apply CNOT to (qa, qb);
-              -- (qa, qb) =q |++\>  -- can use a weaker but more general `EPR' here
-              -- /\ uniform(qa, qb) -- this won't apply then
-              a = meas qa;
-              --   (qa, e1) =q |00\>, (qb, e2) =q |00\>, a = false
-              -- \/ (qa, e1) =q |11\>, (qb, e2) =q |11\>, a = true
-              b = meas qb;
-              --   (qa, e1) =q |00\>, (qb, e2) =q |00\>, a = false, b = false
-              -- \/ (qa, e1) =q |11\>, (qb, e2) =q |11\>, a = true,  b = true
-              return (a, b)
-\end{lstlisting}
+\lstinputlisting[language=QHaskell,firstline=10]{bell.qh}
 
 In the proposition P2, we show that only the portion of the quantum state referred to by the \texttt{apply} command is affected.
-Similarly, in P4 we see that only the relevant portion of the heap containing the control and target qubits used in the CNOT operation are affected.
+Similarly, in P4 we see that only the relevant portion of the heap containing the control and target qubits used in the CX operation are affected.
 In this case, we also needed to combine the two qubit state as they are now entangled.
@@
 
-\begin{lstlisting}[language=QHaskell]
-entangle-disentangle : {T} (a, b) : (Qbit o* Qbit) {a = b}
-= do
-    -- P0: T
-    a <= init false;
-    -- P1: a =q |0\>, class(a)
-    b <= init false;
-    -- P2: P1, b =q |0\>, class(b)
-    apply H to a;
-    -- P3: H_a.P2     =   a =q |+\>, b =q |0\>, uniform(a), class(b)
-    apply CX to (a, b);
-    -- P4: CX_ab.P3   =   (a, b) =q |++\>, uniform(a, b)
-    apply CX to (a, b);
-    -- P5: CX_ab.P4   =   (a =q |+\>, b =q |0\>) \/ (a =q |0\>, b =q |+\>)
-    apply H to a;
-    -- P6: H_a.P5     =   (a =q |0\>, b =q |0\>) \/ (a =q |+\>, b =q |+\>)
-    return (a, b)
-\end{lstlisting}
+\lstinputlisting[language=QHaskell]{entdisent.qh}
 
@@
 					{\inferrule*[right=return]
-						{q : \type{Qbit}, r : \type{Bool} \vdash r \chk \type{Bool}}
-						{q : \type{Qbit}, r : \type{Bool}; P_1 \circ ((q \mapsto -) \multimap \top) \vdash \kw{return} r \syn r : \type{Bool} . R}
-						\\ q : \type{Qbit} \vdash q \chk \type{Qbit}
+						{q : \type{qbit}, r : \type{bool} \vdash r \chk \type{bool}}
+						{q : \type{qbit}, r : \type{bool}; P_1 \circ ((q \mapsto -) \multimap \top) \vdash \kw{return} r \syn r : \type{bool} . R}
+						\\ q : \type{qbit} \vdash q \chk \type{qbit}
 						\\ \cdot; P_0 * (q \mapsto \ket{0}) \prove (q \hookrightarrow -)
 					}
-					{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{meas} q; E_3 \syn r : \type{Bool} . (\exists r : \type{Bool}.R)}
-					\\ \cdot \vdash \kw{0} \chk \type{Bool}
+					{q : \type{qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{meas} q; E_3 \syn r : \type{bool} . (\exists r : \type{bool}.R)}
+					\\ \cdot \vdash \kw{0} \chk \type{bool}
 				}
 				{
-					\cdot; P_0 \vdash q \Leftarrow \kw{init false}; E_{23} \syn r : \type{Bool} . (\exists q : \type{Qbit}.R)
+					\cdot; P_0 \vdash q \Leftarrow \kw{init false}; E_{23} \syn r : \type{bool} . (\exists q : \type{qbit}.R)
 				}
-				\\ r : \type{Bool}; i,m ; R \prove \top \multimap \top \wedge Id(r, \kw{0})
+				\\ r : \type{bool}; i,m ; R \prove \top \multimap \top \wedge Id(r, \kw{0})
 			}
 			{
-				\cdot; \lambda i . \lambda m . i = m \wedge (\top * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bool} . (\top \multimap \top \wedge Id(r, \kw{0}))
+				\cdot; \lambda i . \lambda m . i = m \wedge (\top * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{bool} . (\top \multimap \top \wedge Id(r, \kw{0}))
 			}
 		}
-		{\cdot \vdash \kw{do} E \chk \{\top\} r : \type{Bool} \{\top \wedge \kw{Id}(r, \kw{0})\}}
+		{\cdot \vdash \kw{do} E \chk \{\top\} r : \type{bool} \{\top \wedge \kw{Id}(r, \kw{0})\}}
 	\end{mathpar}
@@
 \subsection{Hadamard basis states}
-\begin{lstlisting}[language=QHaskell]
-qplus : {\top} r : Qbit {r =q |+\>}
-      = do q <= init false;
-           q <= apply H to q;
-           return q
-
-qminus : {\top} r : Qbit {r =q |-\>}
-       = do q <= init true;
-            q <= apply H to q;
-            return q
-\end{lstlisting}
+\lstinputlisting[language=QHaskell]{hadbasis.qh}
 
 \subsection{Creating entanglement}
-\begin{lstlisting}[language=QHaskell]
-share : \Pi a : Qbit.
-        {a =q (alpha |0\> + beta |1\>)}
-           (a, b) : (Qbit o* Qbit)
-        {(a, b) =q (alpha |00\> + beta |11\>)}
-      = \a.do b <= init false;
-              (a, b) <= apply CNot to [a, b];
-              return (a, b)
-\end{lstlisting}
+\lstinputlisting[language=QHaskell]{share.qh}
 
 \subsection{Bell pair}
-\begin{lstlisting}[language=QHaskell]
-bell00 : {\top} (a, b) : (Qbit o* Qbit) {(a,b) =q |\Phi+\>}
-     = do qa <- qplus;
-          (qa, qb) <- share qa;
-          return (qa, qb)
-\end{lstlisting}
+\lstinputlisting[language=QHaskell]{bell00.qh}
 
 \subsection{Testing modular Bell pair}
-\begin{lstlisting}[language=QHaskell]
-testBell : {\top} (a, b) : (Bool o* Bool) {\top /\ a = b}
-         = do (qa, qb) <- bell00;
-              a <= meas qa;
-              b <= meas qb;
-              return (a, b)
-\end{lstlisting}
+\lstinputlisting[language=QHaskell]{testbell.qh}
 
@@
 \subsection{Alice's circuit}
-\begin{lstlisting}[language=QHaskell]
-alice : \Pi q : Qbit. \Pi a : Qbit
-        {q =q (alpha |0\> + beta |1\>) /\ (a, e) =q |\Phi+\>}    -- a is entangled with ghost var e
-           (m1, m2) : (Bool o* Bool)
-        {  (m1, m2) = 00 /\ e =q (alpha |0\> + beta |1\>)
-         \/ (m1, m2) = 01 /\ e =q (alpha |1\> + beta |0\>)
-         \/ (m1, m2) = 10 /\ e =q (alpha |0\> - beta |1\>)
-         \/ (m1, m2) = 11 /\ e =q (alpha |1\> - beta |0\>)}
-      = \q.\a.do (q, a) <= apply CNOT to [q, a];
-                 q <= apply H to q;
-                 rq <= meas q;
-                 ra <= meas a;
-                 return (rq, ra)
-\end{lstlisting}
+\lstinputlisting[language=QHaskell,lastline=12]{teleport.qh}
 
@@
 \todo{TODO provide correct propositions using ghost e}
-\begin{lstlisting}[language=QHaskell]
-bob : \Pi m1 m2 : Bool. \Pi b : Qbit
-      {(b, e) =q |\Phi+\>}
-         r : Qbit
-      {\top}
-      = \m1.\m2.\b.do (m1, b) <= apply C_Z to [m1, b];
-                      (m2, b) <= apply C_X to [m2, b];
-                      return b
-\end{lstlisting}
+\lstinputlisting[language=QHaskell,firstline=14,lastline=20]{teleport.qh}
 
 \subsection{Teleport}
-\begin{lstlisting}[language=QHaskell]
-teleport : \Pi q : Qbit.
-           {q =q (alpha |0\> + beta |1\>)}
-              b : Qbit
-           {r =q (alpha |0\> + beta |1\>)}
-         = \q.do (a, b) <- bell00;
-                 (m1, m2) <- alice q a;
-                 b <- bob m1 m2 b;
-                 return b
-\end{lstlisting}
+\lstinputlisting[language=QHaskell,firstline=22]{teleport.qh}
 

### 2020-11-13 ms-paper/ms-paper.tex
Revise projections and their appearance

@@
 	{∧}{{$\wedge$}}1 {∨}{{$\vee$}}1
-	{∈}{{$\in$}}1 {Π}{{$\Pi$}}1
-	{⊗}{{$\otimes$}}1
+	{∈}{{$\in$}}1 {Π}{{$\Pi$}}1 {∃}{{$\exists$}}1
+	{⊗}{{$\otimes$}}1 {⊕}{{$\oplus$}}1
 	{⊤}{{$\top$}}1
@@
 	{α}{{$\alpha$}}1 {β}{{$\beta$}}1
+	{_i}{{$_i$}}1
+	{\ .\ }{{$\cdot$}}1
 }
@@
 	\item $X \vee Y$ is modeled as the sum (linear combination) of the subspaces $X$ and $Y$.
-In other words, it is the closed subspace of the union of the two subspaces: $\overline{\mathit{span}(X \cup Y)}$, where $\overline{X}$ denotes that $X$ is a closed subspace.
+In other words, it is the closed subspace of the union of the two subspaces: $\overline{\mathit{span}(X \cup Y)}$, where $\overline{X}$ denotes the closure of $X$.
 	\item $X^\bot$ denotes the orthocomplement of $X$ which includes all vectors in $\mathcal{H}$ that are orthogonal to $X$.
@@
 
+\section{Other examples}
+
+\subsection{QPL}
+
+\lstinputlisting[language=QHaskell]{qpl.qh}
+
+\subsection{Quantum Lambda Calculus}
+\lstinputlisting[language=QHaskell]{qlc.qh}
+
+\subsection{QWire Core}
+\lstinputlisting[language=QHaskell]{paykin.qh}
+
 \end{document}
